# Base types {#base-types}

Underlying every R object is a C structure (or struct) that describes how that object is stored in memory. The struct includes the contents of the object, the information needed for memory management, and, most importantly for this section, a __type__. This is the __base type__ of an R object.

Base types are not really an object system because only the R core team can create new types. Functions that behave differently for different base types are almost always written in C, where dispatch occurs using switch statements (e.g., `switch(TYPEOF(x))`). As a result, new base types are added very rarely: the most recent change, in 2011, added two exotic types that you never see in R, but are useful for diagnosing memory problems (`NEWSXP` and `FREESXP`). Prior to that, the last type added was a special base type for S4 objects (`S4SXP`) in 2005. \indexc{SEXP} \index{base types} \index{objects!base types}

Even if you never write C code, it's important to understand base types because everything else is built on top of them: S3 objects can be built on top of any base type, S4 objects use a special base type, and R6 objects are a combination of S3 and environments (another base type). To see if an object is a pure base type, i.e., it doesn't also have S3, S4, or R6 behaviour, check that `is.object(x)` returns `FALSE`.

## Types

In [data structures] you learned about the most important set of base types, the vectors (logical, integer, real, complex, character, list, and raw). Other important types are:

* Functions, including primitive and special functions.

* Environments.

* Symbols and pairlists, used to represent the "abstract syntax tree"
  of parsed R code..

In total, there are XXX base types, but it's not important to understand most of them unless you are writing C code.

You can determine an object's base type with `typeof()`:  \indexc{typeof()} 

```{r}
f <- function() {}
typeof(f)

typeof(sum)
```

Be careful with the `is.` functions as they use a different set of naming conventions:

```{r}
is.function(f)
is.primitive(sum)
```

Generally, "is" functions that check if an object is a specific type are ok. "is" functions that check for a family of types are often surprising. For example, `is.atomic(NULL)` is true, and as well as checking than an object is a vector, `is.vector()` also checks that it has no attributes apart from names.

You may have heard of `mode()` and `storage.mode()`. I recommend ignoring these functions because they're just aliases of the names returned by `typeof()`, and exist solely for S compatibility. Read their source code if you want to understand exactly what they do. \indexc{mode()}

