# Base types {#base-types}

> Everything that exists in R is an object
> --- John Chambers

Everything in R is an object; a thing that exists and that you can manipulate. Some objects are also "object-oriented": they belong to a "class", and that class is used to select between different implementations (i.e. methods). This is what `is.object()` tests for (it really should be called `is.oo()`).

While some objects have a class, every object has a __base type__. An object in R can not exist without a base type. Base types are not a fully-fledged OOP system because they have no ability for extension. Functions that behave differently for different base types are almost always written in C, where dispatch occurs using switch statements (e.g., `switch(TYPEOF(x))`). This means only R-core can create new types.

As a consequence, new base types are added very rarely. The most recent change, in 2011, added two exotic types that you never see in R, but are needed for diagnosing memory problems (`NEWSXP` and `FREESXP`). Prior to that, the last type added was a special base type for S4 objects (`S4SXP`) in 2005. \indexc{SEXP} \index{base types} \index{objects!base types}

It's important to understand the base types because everything else is built on top of them: S3 objects can be built on top of most base types, S4 objects can be built on anything S3 can be, or can use a special base type, and R6 is built on top of the environment base type. S3 and S4 also need to interact with the base type system, so we'll need some vocab. The precision of distinguishing between and S3 object and it's underlying base type will also be useful.

## Types in R

Use `typeof()` to determine the base type of an object:  \indexf{typeof}

```{r}
typeof(mean)
typeof(mtcars)
```

Most base types:

* NULL, logical, integer, double, complex, character, list, raw
* symbol, pairlist, language
* environment, 
* closure, special, builtin
* externalptr, weakref

* expression
* bytecode
* S4

* promise
* ...
* any

You can see if an object is a pure base type, i.e., it doesn't also have S3, S4, or R6 behaviour, with `is.object(x)`:

```{r}
is.object(list())
is.object(mtcars)
```

Not that `class()` will always return something even if it's not an object:

```{r}
class(1:10)
```

Generally, this returns the type of base objects, except:

* a base object with attribute dim -> matrix, array
* closure, special, builtin -> function
* double -> numeric
* symbol -> name
* calls to `if`, `while`, `for`, `=`, `<-`, `(`, and `{` 

These differences are typically unimportant provide that you remember to never call `class()` on an object where `is.object()` is `FALSE`.


You may have heard of `mode()` and `storage.mode()`. I recommend ignoring these functions because they're just aliases of the names returned by `typeof()`, and exist solely for S compatibility. Read their source code if you want to understand exactly what they do. \indexc{mode()}

### The is functions

Be careful with the `is.` functions as they use a different set of naming conventions:

```{r}
is.function(mean)
is.primitive(sum)
```

Generally, "is" functions that check if an object is a specific type are ok. "is" functions that check for a family of types are often surprising. For example, `is.atomic(NULL)` is true, and as well as checking that an object is a vector, `is.vector()` also checks that it has no attributes apart from names.

* Specific base type: `is.call()`, `is.character()`, `is.complex()`, 
  `is.double()`, `is.environment()`, `is.expression()`, `is.integer()`,
  `is.list()`, `is.logical()`, `is.name()`, `is.null()`, `is.pairlist()`,
  `is.raw()`, `is.symbol()`

* Combination of base types: `is.atomic()`, `is.function()`,
  `is.language()`, `is.primitive()`, `is.recursive()`

* Has an S3 class: `is.data.frame()`, `is.factor()`, `is.mts()`, 
  `is.numeric_version()`, `is.ordered()`, `is.package_version()`, `is.qr()`, 
  `is.raster()`, `is.relistable()`, `is.stepfun()`, `is.table()`, `is.ts()`,
  `is.tskernel()`, 

* Check attributes: `is.array()`, `is.leaf()`, `is.matrix()`, `is.vector()`

* Vectorised math ops: `is.finite()`, `is.infinite()`, `is.na()`, `is.nan()`

* Finally there are a bunch of special purpose functions that don't 
  fall into any other category: `is.element(x, y)`, `is.empty.model()`
  `is.loaded()`, `is.object()`, `is.R()`, `is.single()`, `is.unsorted()`

One, `is.numeric()`, is sufficiently special that it gets it's own section.

### The numeric "type"

We need a little extra discussion of the numeric "type" because it's used in three different ways in different places in R.

1.  In some places it's used as an alias for "double".  For example
    `as.numeric()` is identical to `as.double()`, and `numeric()` is
    identical to `double()`.
    
1.  In a couple of places it means a base type of either integer or double. 

    ```{r, error = TRUE}
    mode(1)
    mode(1L)
    mode(factor("x"))
    ```
    
    As well as mode, this convention is used for the implicit class and 
    for S4 classes.

1.  In some places it means an object built on a base type of integer or 
    double that has numeric behaviour (i.e. arithmetic makes sense and you
    can order by numeric values)
    
    ```{r}
    is.numeric(1)
    is.numeric(1L)
    is.numeric(factor("x"))
    ```

Finally, there are few places in which R uses "real" instead of double; `NA_real_` is the one place that you're likely to encounter this in practice.

### Switchpatch

If you're writing your own function and you want different behaviour based on the base type, use this pattern:

```{r}
bytes <- function(x) {
  switch(
    typeof(x),
    integer = 4,
    numeric = 8,
    logical = 4,
    NA
  )
}
```

We call it __switchpatch__, a portmanteau of switch and dispatch.

## Types in C

The internal types, or formally the SEXPTYPE, are defined in C code by an enum, a C progamming construct that maps strings to integers. I've reorganise the enum into blocks of related types, but you can get a rough sense of timeline of the types by looking at their associated number.

```c
typedef enum {
  NILSXP      =  0, // NULL

  // Atomic vectors
  LGLSXP      = 10, // logical
  INTSXP      = 13, // integer
  REALSXP     = 14, // double
  CPLXSXP     = 15, // complex
  STRSXP      = 16, // character
  RAWSXP      = 24, // raw
  VECSXP      = 19, // list

  // Language components
  SYMSXP      =  1, // symbols/names
  LISTSXP     =  2, // pairlists
  CLOSXP      =  3, // functions
  ENVSXP      =  4, // environments
  PROMSXP     =  5, // promises
  LANGSXP     =  6, // calls
  SPECIALSXP  =  7, // internal functions
  BUILTINSXP  =  8, // primitive functions
  
  // Internal
  S4SXP       = 25, // S4 objects not built on existing type
  ANYSXP      = 18, // ANY args for S4 dispatch
  DOTSXP      = 17, // ...
  EXPRSXP     = 20, // expression()
  BCODESXP    = 21, // byte code
  EXTPTRSXP   = 22, // external pointer
  WEAKREFSXP  = 23, // weak reference
  CHARSXP     =  9, // "scalar" string type (internal only)
  NEWSXP      = 30, // fresh node creaed in new page
  FREESXP     = 31, // node released by GC
} SEXPTYPE;
```
