---
title: R's C interface
layout: default
---

```{r, echo = FALSE}
library(inline, quietly = TRUE)
```

# R's C interface {#c-api}

## Introduction

Reading the source code of R is an extremely powerful technique for improving your R programming. However, at some point you will hit a brick wall: many R functions are implemented in C. This guide gives you a basic introduction to C and to R's internal C api, giving you the knowledge needed to read the R internals. The contents of this draw heavily from Section 5 ("System and foreign language interfaces") of [Writing R extensions](http://cran.r-project.org/doc/manuals/R-exts.html), but focus on best practices and modern tools. This means it does not cover the old `.C` interface, the old api defined in `Rdefines.h` and rarely used language features. You are unlikely to see these in modern R code. To see the complete C api provided the R, look at the header file `Rinternals.h`. It's easiest to find and display this file from within R:

```{r, eval = FALSE}
rinternals <- file.path(R.home("include"), "Rinternals.h")
file.show(rinternals)
```

I do not recommend using C for writing new high-performance code. Instead use Rcpp and write C++. The Rcpp API protects you from many of the historical idiosyncracies of the R API, takes care of memory management for you, and provides many useful helper methods.

### Prerequisites

To understand existing C code, it's useful to generate simple examples of your own that you can experiment with. To that end, all examples in this chapter use the inline package, which makes it extremely easy to compile and link C code to your current R session.  Get it by running `install.packages("Rcpp")`.

You'll also need a C compiler. Windows users can use Duncan Murdoch's [Rtools](http://cran.r-project.org/bin/windows/Rtools/). Mac users will need the [Xcode command line tools](http://developer.apple.com/). Most Linux distributions will come with the necessary compilers.

## Differences between R and C

Even if you've never used C before, you still should be able to read C code relatively easily because the basic structure is similar to R. Some important differences from R include:

* Variables can only store specific types of object, and must be declared before 
  use.

* Objects are modified in place, unless you specifically copy the object.

* Indices start at 0, not 1.

* You must use a semi-colon at end of each expression.

* You must have an explicit return statement.

* Assignment is done with `=`, not `<-`.

To learn C more formally, I recommend the classic "[The C programming language](http://amzn.com/0131101633?tag=devtools-20)" by Kernigan and Ritchie.

## Calling C functions from R

Generally, to call a C function from R requires two pieces: a C function, and an R wrapper function that uses `.Call()`. The simple function below adds two numbers together and illustrates some of the important features of coding in C (creating new R vectors, coercing input arguments to the appropriate type and dealing with garbage collection).  

```c
// In C ----------------------------------------
#include <R.h>
#include <Rinternals.h>

SEXP add(SEXP a, SEXP b) {
  SEXP result;

  PROTECT(result = allocVector(REALSXP, 1));
  REAL(result)[0] = asReal(a) + asReal(b);
  UNPROTECT(1);

  return result;
}
```

```{r, eval = FALSE}
# In R ----------------------------------------
add <- function(a, b) {
  .Call("add", a, b)
}
```

(An alternative to using `.Call` is to use `.External`.  It is used almost identically, except that the C function will recieve a single argument containing a `LISTSXP`, a pairlist from which the arguments can be extracted. This makes it possible to write functions that take a variable number of arguments. However, it's not commonly used in base R, `inline` does not currently support `.External` functions so I don't discuss it further in this chapter).

In this chapter we'll produce the two pieces in one step by using the `inline` package. This allows us to write:

```{r, cache = TRUE}
add <- cfunction(c(a = "integer", b = "integer"), "
  SEXP result;

  PROTECT(result = allocVector(REALSXP, 1));
  REAL(result)[0] = asReal(a) + asReal(b);
  UNPROTECT(1);

  return result;
")
add(1, 5)
```

Before we begin writing and reading C code, we need to know a little about the basic data structures.

## Basic data structures

At the C-level, all R objects are stored in a common datatype, the `SEXP`. (Technically, this is a pointer to a structure with typedef `SEXPREC`). A `SEXP` is a variant type, with subtypes for all R's data structures. The most important types are:

* `REALSXP`: numeric vectors
* `INTSXP`: integer vectors
* `LGLSXP`: logical vectors
* `STRSXP`: character vectors
* `VECSXP`: lists
* `CLOSXP`: functions (closures)
* `ENVSXP`: environments

__Beware:__ At the C level, R's lists are `VECSXP`s not `LISTSXP`s. This is because early implementations of R used Lisp-like linked lists (now known as "pairlists") before moving to the S-like generic vectors that we now know as lists.

Character vectors are a little more complicated than the other atomic vectors. A `STRSXP`s contains a vector of `CHARSXP`s, where each `CHARSXP` points to C-style string stored in a global pool. This design allows individual `CHARSXP`'s to be shared between multiple strings, reducing memory usage. See [object size](#object_size) for more details.

There are also `SEXP`s for less common object types:

* `CPLXSXP`: complex vectors
* `LISTSXP`: "pair" lists. At the R level, you only need to care about the distinction lists and pairlists for function arguments, but internally they are used in many more places.
* `DOTSXP`: '...'
* `SYMSXP`: names/symbols
* `NILSXP`: `NULL`

And `SEXP`s for internal objects, objects that are usually only created and used by C functions, not R functions:

* `LANGSXP`: language constructs
* `CHARSXP`: "scalar" strings
* `PROMSXP`: promises, lazily evaluated function arguments
* `EXPRSXP`: expressions

There's no built-in R function to easily access these names, but we can write one: 
  
```{r, cache = TRUE}
sexp_type <- cfunction(c(x = "ANY"), '
  switch (TYPEOF(x)) {
    case NILSXP:      return mkString("NILSXP");
    case SYMSXP:      return mkString("SYMSXP");
    case LISTSXP:     return mkString("LISTSXP");
    case CLOSXP:      return mkString("CLOSXP");
    case ENVSXP:      return mkString("ENVSXP");
    case PROMSXP:     return mkString("PROMSXP");
    case LANGSXP:     return mkString("LANGSXP");
    case SPECIALSXP:  return mkString("SPECIALSXP");
    case BUILTINSXP:  return mkString("BUILTINSXP");
    case CHARSXP:     return mkString("CHARSXP");
    case LGLSXP:      return mkString("LGLSXP");
    case INTSXP:      return mkString("INTSXP");
    case REALSXP:     return mkString("REALSXP");
    case CPLXSXP:     return mkString("CPLXSXP");
    case STRSXP:      return mkString("STRSXP");
    case DOTSXP:      return mkString("DOTSXP");
    case ANYSXP:      return mkString("ANYSXP");
    case VECSXP:      return mkString("VECSXP");
    case EXPRSXP:     return mkString("EXPRSXP");
    case BCODESXP:    return mkString("BCODESXP");
    case EXTPTRSXP:   return mkString("EXTPTRSXP");
    case WEAKREFSXP:  return mkString("WEAKREFSXP");
    case S4SXP:       return mkString("S4SXP");
    case RAWSXP:      return mkString("RAWSXP");
    default:          return mkString("<unknown>");
}')
sexp_type(10)
sexp_type(10L)
sexp_type("a")
sexp_type(T)
sexp_type(list(a = 1))
sexp_type(pairlist(a = 1))
```

(code adapated from R's `inspect.c`)

## Coercion and object creation

At the heart of every C function will be a set of conversions between R data structures and C data structures. Inputs and output will always be R data structures (`SEXP`s) and you will need to convert them to C data structures in order to do any work. An additional complication is the garbage collector: if you don't claim every R object you create, the garbage collector will think they are unused and delete them.

### Object creation and garbage collection

The simplest way to create an new R-level object is `allocVector`, which takes two arguments, the type of `SEXP` (or `SEXPTYPE`) to create, and the length of the vector. The following code creates a three element list containing a logical vector, a numeric vector and an integer vector:

```{r, cache = TRUE}
dummy <- cfunction(body = '
  SEXP vec, real, lgl, ints;

  PROTECT(real = allocVector(REALSXP, 2));
  PROTECT(lgl = allocVector(LGLSXP, 10));
  PROTECT(ints = allocVector(INTSXP, 10));

  PROTECT(vec = allocVector(VECSXP, 3));
  SET_VECTOR_ELT(vec, 0, real);
  SET_VECTOR_ELT(vec, 1, lgl);
  SET_VECTOR_ELT(vec, 2, ints);

  UNPROTECT(4);
  return vec;
')
dummy()
```

You might wonder what all the `PROTECT` calls do. They tell R that we're currently using each object, and not to delete it if the garbage collector is activated. (We don't need to protect objects that R already knows about, like function arguments).

You also need to make sure that every protected object is unprotected. `UNPROTECT` takes a single integer argument, `n`, and unprotects the last n objects that were protected. The number of protects and unprotects must match, if not, R will warn about a "stack imbalance in .Call".  Other specialised forms of `PROTECT` and `UNPROTECT` are needed in some circumstances: `UNPROTECT_PTR(s)` unprotects the object pointed to by the `SEXP` s, `PROTECT_WITH_INDEX` saves an index of the protection location that can be used to replace the protected value using `REPROTECT`. Consult the R externals section on [garbage collection](http://cran.r-project.org/doc/manuals/R-exts.html#Garbage-Collection) for more details.

If you run `dummy()` a few times, you'll notice the output is basically random. This is because `allocVector` assigns memory to each output, but it doesn't clean it out first. For real functions, you'll may want to loop through each element in the vector and zero it out. The most efficient way to do that is to use `memset`:

```{r, cache = TRUE}
zeroes <- cfunction(c(n_ = "integer"), '
  int n = asInteger(n_);
  SEXP out;

  PROTECT(out = allocVector(INTSXP, n));
  memset(INTEGER(out), 0, n * sizeof(int));
  UNPROTECT(1);

  return out;
')
zeroes(10);
```

### Allocation shortcuts

There are a few shortcuts for allocating matrices and 3d arrays:

* `allocMatrix(SEXPTYPE mode, int nrow, int ncol)`
* `alloc3DArray(SEXPTYPE mode, int nrow, int ncol, int nface)`

These create the object and set the needed attributes. Beware `allocList()`: it creates a pairlist, not a regular list.

`mkNamed()` simplifies the creation of named vectors. The following code is equivalent to `list(a = NULL, b = NULL, c = NULL)`:

```c
const char *names[] = {"a", "b", "c", ""};
mkNamed(VECSXP, names);
```

### Extracting C vectors

There is a helper function for each atomic vector (apart from character, see following) that allows you to index into a `SEXP` and access the C-level data structure that lives at its heart. The following example shows the use of the helper `REAL()` to inspect and modify a numeric vector:

```{r, cache = TRUE}
add_one <- cfunction(c(x = "numeric"), "
  SEXP out;
  int n = length(x);

  PROTECT(out = allocVector(REALSXP, n));
  for (int i = 0; i < n; i++) {
    REAL(out)[i] = REAL(x)[i] + 1;
  }
  UNPROTECT(1);

  return out;
")
add_one(as.numeric(1:10))
```

There are similar helpers for logical, `LOGICAL(x)`, integer, `INTEGER(x)`, complex, `COMPLEX(x)` and raw vectors `RAW(x)`. If you're working with longer vectors, there's a performance advantage to using the helper function once and saving the result in a pointer:

```{r, cache = TRUE}
add_two <- cfunction(c(x = "numeric"), "
  SEXP out;
  int n = length(x);
  double *px, *pout;

  PROTECT(out = allocVector(REALSXP, n));

  px = REAL(x);
  pout = REAL(out);
  for (int i = 0; i < n; i++) {
    pout[i] = px[i] + 2;
  }
  UNPROTECT(1);

  return out;
")
add_two(as.numeric(1:10))

library(microbenchmark)
x <- as.numeric(1:1e6)
microbenchmark(
  add_one(x),
  add_two(x)
)
```

On my computer, `add_two` is about twice as fast as `add_one` for a million element vector. This is a common idiom in R source code.

Strings and lists are more complicated because the individual elements are `SEXP`s not C-level data structures. You can use `STRING_ELT(x, i)` and `VECTOR_ELT(x, i)` to extract individual components of strings and lists respectively. To get a single C string from a element in a R character vector, use `CHAR(STRING_ELT(x, i))`. Modify components of list or character vector with `SET_VECTOR_ELT()` and `SET_STRING_ELT()`.

### Modifying strings

String vectors are a little more complicated. As discussed earlier, a string vector is a vector made up of pointers to immutable `CHARSXP`s, and it's the `CHARSXP` that contains the C string (which can be extracted using `CHAR`). The following function shows how to create a vector of fixed values:

```{r, cache = TRUE}
abc <- cfunction(NULL, '
  SEXP out;
  PROTECT(out = allocVector(STRSXP, 3));

  SET_STRING_ELT(out, 0, mkChar("a"));
  SET_STRING_ELT(out, 1, mkChar("b"));
  SET_STRING_ELT(out, 2, mkChar("c"));

  UNPROTECT(1);

  return out;
')
abc()
```

Things are a little harder if you want to modify the strings in the vector because you need to know a lot about string manipulation in C (which is hard, and harder to do right). For any problem that involves any kind of string modification, you're better off using Rcpp. 

The following function just makes a copy of a string, so you can at least see how all the pieces work together.

```{r, cache = TRUE}
copy <- cfunction(c(x = "character"), '
  SEXP out;
  int n = length(x);
  const char* letter;

  PROTECT(out = allocVector(STRSXP, n));
  for (int i = 0; i < n; i++) {
    letter = CHAR(STRING_ELT(x, i));
    SET_STRING_ELT(out, i, mkChar(letter));
  }
  UNPROTECT(1);
  
  return out;
')
copy(letters[1:10])
```

One last useful function for operating with strings: we can use `STRING_PTR()` to get a pointer to the `STRING_ELT()`s in a vector, so that we can access the `STRING_ELT()`s by de-referencing the pointer. Occasionally, this can be easier to work with. The following example illustrates how you might use `STRING_PTR()` reverse a character vector:

```{r, cache = TRUE}
reverse <- cfunction(c(x = "character"), '
  int len = length(x);
  SEXP out;
  PROTECT(out = allocVector(STRSXP, len));
  
  SEXP* out_ptr = STRING_PTR(out);
  SEXP* x_ptr = STRING_PTR(x);
  for(int i = 0; i < len; ++i) {
    out_ptr[i] = x_ptr[len - i - 1];
  }
  
  UNPROTECT(1);
  return out;
')

reverse(letters[1:10])
```

### Coercing scalars

There also a few helper functions if you want to turn the length on R vector into a C scalar:

* `asLogical(x): INTSXP -> int`
* `asInteger(x): INTSXP -> int`
* `asReal(x): REALSXP -> double`
* `CHAR(asChar(x)): STRSXP -> const char*`

And similarly it's easy to turn a C scalar into a length-one R vector:

* `ScalarLogical(x): int -> LGLSXP`
* `ScalarInteger(x): int -> INTSXP`
* `ScalarReal(x): double -> REALSXP`
* `mkString(x): const char* -> STRSXP`

These all create R-level objects, so need to be `PROTECT`ed.

## Modifying objects

You must be very careful when modifying an object that the user has passed into the function. The following function has some very unexpected behaviour:

```{r, cache = TRUE}
add_three <- cfunction(c(x = "numeric"), '
  REAL(x)[0] = REAL(x)[0] + 3;
  return x;
')
x <- 1
y <- x
add_three(x)
x
y
```

Not only has it modified the value of `x`, but it has also modified `y`!  This happens because of the way that R implements it's copy-on-modify philosophy. It does so lazily, so a complete copy only has to be made if you make a change: `x` and `y` point to the same object, and the object is only duplicated if you change either `x` or `y`.

To avoid problems like this, always `duplicate()` inputs before modifying them:

```{r, cache = TRUE}
add_four <- cfunction(c(x = "numeric"), '
  SEXP x_copy;
  PROTECT(x_copy = duplicate(x));
  REAL(x_copy)[0] = REAL(x_copy)[0] + 4;
  UNPROTECT(1);
  return x_copy;
')
x <- 1
y <- x
add_four(x)
x
y
```

In some scenarios you might know that it's ok to modify the object in place (for example you've constructed it in R for the sole purpose of modifying in C). In that scenario, you should double-check `NO_REFERENCES(x)` is true. If you're working with lists, use `shallow_duplicate()` to duplicate the list, but not its individual components.

## Pairlists and symbols

R hides a few details of the underlying datastructures it uses. In some places in R code, it looks like you're working with a list (`VECSXP`), but behind the scenes R you're actually modifying a pairlist (`LISTSXP`). These include attributes, calls and `...`.  Pairlists differ from lists in the following ways:

1. Pairlists are [linked lists](http://en.wikipedia.org/wiki/Linked_list), a 
   data structure which does not have an easy way to get to an arbitrary 
   element.

1. Pairlists have `tags`, not `names`, and tags are symbols, not strings. 

Because you can't easily index into a specified location in a pairlist, R provides a set of helper functions to move along the linked list. The basic helpers are `CAR()` which extracts the first element of the list, and `CDR()` which extracts the rest. These can be composed to get `CAAR()`, `CDAR()`, `CADDR()`, `CADDR()`, `CADDDR()`. As well as the getters, R also provides `SETCAR()`, `SETCDR()` etc.

```{r, cache = TRUE}
car <- cfunction(c(x = "ANY"), 'return CAR(x);')
cdr <- cfunction(c(x = "ANY"), 'return CDR(x);')
cadr <- cfunction(c(x = "ANY"), 'return CADR(x);')

x <- quote(f(a = 1, b = 2))
car(x)
cdr(x)
car(cdr(x))
cadr(x)
```

Similarly, you can loop through all elements of a pairlist as follows:

```{r, cache = TRUE}
count <- cfunction(c(x = "ANY"), '
  SEXP el, nxt;
  int i = 0;

  for(nxt = x; nxt != R_NilValue; el = CAR(nxt), nxt = CDR(nxt)) {
    i++;
  }
  return ScalarInteger(i);
')
count(quote(f(a, b, c)))
count(quote(f()))
```

You can make new pairlists with `CONS()` and new calls with `LCONS()`. A pairlist is always terminated with `R_NilValue`.

```{r, cache = TRUE}
new_call <- cfunction(NULL, '
  SEXP out;

  out = LCONS(install("+"), LCONS(
    ScalarReal(10), LCONS(
      ScalarReal(5), R_NilValue)
    )
  );
  return out;
')
new_call();
```

`TAG()` and `SET_TAG()` allow you to get and set the tag (aka name) associated with an element of a pairlist. The tag should be a symbol. To create a symbol (the equivalent of `as.symbol()` or `as.name()` in R), use `install()`. 

Attributes are also pairlists behind the scenes, but come with the helper functions `setAttrib()` and `getAttrib()` to make access a little easier:

```{r, cache = TRUE}
set_attr <- cfunction(c(obj = "ANY", attr = "character", value = "ANY"), '
  const char* attr_s = CHAR(asChar(attr));

  duplicate(obj);
  setAttrib(obj, install(attr_s), value);
  return obj;
')
x <- 1:10
set_attr(x, "a", 1)
```

There are some (confusingly named) shortcuts for common setting operations: `classgets()`, `namesgets()`, `dimgets()` and `dimnamesgets()` are the internal versions of the default methods of `class<-`, `names<-`, `dim<-` and `dimnames<-`. 

## Missing and non-finite values

Each atomic vector has a special constant for getting or setting missing values:

* `INTSXP`: `NA_INTEGER`
* `LGLSXP`: `NA_LOGICAL`
* `STRSXP`s: `NA_STRING`
  
Missing values are somewhat more complicated for `REALSXP` because `NAs` are a special type of NaN, and there are special values for positive and negative infinity. Use `ISNA()`, `ISNAN()`, and `R_FINITE()` macros to check for missing, NaN or non-finite values.  Use the constants `NA_REAL`, `R_NaN`, `R_PosInf` and `R_NegInf` to set those values.

We can combine all these together to implement a primitive version of `is.NA()`:

```{r, cache = TRUE}
is_na <- cfunction(c(x = "ANY"), '
  SEXP out;
  int n = length(x);

  PROTECT(out = allocVector(LGLSXP, n));

  for (int i = 0; i < n; i++) {
    switch(TYPEOF(x)) {
      case LGLSXP:
        LOGICAL(out)[i] = (LOGICAL(x)[i] == NA_LOGICAL);
        break;
      case INTSXP:
        LOGICAL(out)[i] = (INTEGER(x)[i] == NA_INTEGER);
        break;
      case REALSXP:
        LOGICAL(out)[i] = ISNA(REAL(x)[i]);
        break;
      case STRSXP:
        LOGICAL(out)[i] = (STRING_ELT(x, i) == NA_STRING);
        break;
      default:
        LOGICAL(out)[i] = NA_LOGICAL;
    }
  }
  UNPROTECT(1);

  return out;
')
is_na(c(NA, 1L))
is_na(c(NA, 1))
is_na(c(NA, "a"))
is_na(c(NA, TRUE))
```

It's worth noting that R's `base::is.na()` returns `TRUE` for both `NA` and `NaN`s in a numeric vector, as opposed to the C `ISNA()` macro, which returns `TRUE` only for `NA_REAL`s.

There are a few other special values:

* R-level `NULL` is C-level `R_NilValue`
* An unbound argument is represented by `R_UnboundValue`
* A missing argument is represented by `R_MissingArg`

## Checking types in C

If the user provides unexpected input to your function (e.g. a list instead of a numeric vector), it's very easy to crash R. For this reason, it's a good idea to write a wrapper function that checks arguments are of the correct type, or coerces them if necessary. It's usually easier to do this at the R level. For example, going back to our first example of C code, we might rename it to `add_` and then write a wrapper function to check the inputs are ok:

```{r, cache = TRUE}
add_ <- cfunction(signature(a = "integer", b = "integer"), "
  SEXP result;

  PROTECT(result = allocVector(REALSXP, 1));
  REAL(result)[0] = asReal(a) + asReal(b);
  UNPROTECT(1);

  return result;
")
add <- function(a, b) {
  stopifnot(is.numeric(a), is.numeric(b), length(a) == 1, length(b) == 1)
  add_(a, b)
}
```

Or if we wanted to be more accepting of diverse inputs:

```{r, cache = TRUE}
add <- function(a, b) {
  a <- as.numeric(a)
  b <- as.numeric(b)

  if (length(a) > 1) warning("Only first element of a used")
  if (length(b) > 1) warning("Only first element of b used")
  
  add_(a, b)
}
```

To coerce objects at the C level, use `PROTECT(new = coerceVector(old, SEXPTYPE))`. This will return an error if the `SEXP` can not be converted to the desired type. 

To check if an object is of a specified type, you can use `TYPEOF`, which returns a `SEXPTYPE`:

```{r, cache = TRUE}
is_numeric <- cfunction(c("x" = "ANY"), "
  return ScalarLogical(TYPEOF(x) == REALSXP);
")
is_numeric(7)
is_numeric("a")
```

Or you can use one of the many helper functions. They all return 0 for FALSE and 1 for TRUE:

* For atomic vectors: `isInteger()`, `isReal()`, `isComplex()`, `isLogical()`,
  `isString()`.

* For combinations of atomic vectors: `isNumeric()` (integer, logical, real), 
  `isNumber()` (integer, logical, real, complex), `isVectorAtomic()` 
  (logical, interger, numeric, complex, string, raw)

* Matrices (`isMatrix()`) and arrays (`isArray()`)

* For other more esoteric object: `isEnvironment()`, `isExpression()`, 
 `isList()` (a pair list), `isNewList()` (a list), `isSymbol()`, `isNull()`,
 `isObject()` (S4 objects), `isVector()` (atomic vectors, lists, expressions).

Note that some of these functions behave differently to the R-level functions with similar names. For example `isVector()` is true for atomic vectors, lists and expressions, where `is.vector()` is returns `TRUE` only if its input has no attributes apart from names.

## Finding the C source code for a function

In many R functions you'll find code like `.Internal(mean(x))` or `.Primitive("sum")`. That means that most of the function is implemented at the C-level. There are two steps to finding the corresponding C source code:

1. Open [src/main/names.c](https://github.com/wch/r-source/blob/trunk/src/main/names.c) 
   and search for the name of the function.  You'll find an entry that tells 
   you the name of the function (which always starts with `do_`).

1. Next, search the R source code for that name. To make it easier to find 
   where it's defined (rather than everywhere it's used), you can add `(SEXP`.  
   e.g. to find the source code for `findInterval`, search for 
   `do_findinterval(SEXP`.
