# Code generation

```{r setup, include = FALSE}
source("common.R")
library(pryr)
```

## Calls {#calls}

### Modifying a call

You can add, modify, and delete elements of the call with the standard replacement operators, `$<-` and `[[<-`: \index{calls|modifying}

```{r}
y <- quote(read.csv("important.csv", row.names = FALSE))
y$row.names <- TRUE
y$col.names <- FALSE
y

y[[2]] <- quote(paste0(filename, ".csv"))
y[[4]] <- NULL
y

y$sep <- ","
y
```

Calls also support the `[` method. But use it with care. Removing the first element is unlikely to create a useful call.

```{r}
x <- quote(read.csv("important.csv", row.names = FALSE))

x[-3] # remove the second argument
x[-1] # remove the function name - but it's still a call!
x
```

If you want a list of the unevaluated arguments (expressions), use explicit coercion:

```{r}
# A list of the unevaluated arguments
as.list(x[-1])
```

Generally speaking, because R's function calling semantics are so flexible, getting or setting arguments by position is dangerous. For example, even though the values at each position are different, the following three calls all have the same effect:

```{r}
m1 <- quote(read.delim("data.txt", sep = "|"))
m2 <- quote(read.delim(s = "|", "data.txt"))
m3 <- quote(read.delim(file = "data.txt", , "|"))
```

To work around this problem, pryr provides `standardise_call()`. It uses the base `match.call()` function to convert all positional arguments to named arguments: \indexc{standardise\_call()} \indexc{match.call()}

```{r}
standardise_call(m1)
standardise_call(m2)
standardise_call(m3)
```

### Creating a call from its components

To create a new call from its components, you can use `call()` or `as.call()`. The first argument to `call()` is a string which gives a function name. The other arguments are expressions that represent the arguments of the call. \indexc{call()} \indexc{as.call()}

```{r}
call(":", 1, 10)
call("mean", quote(1:10), na.rm = TRUE)
```

`as.call()` is a minor variant of `call()` that takes a single list as input. The first element is a name or call. The subsequent elements are the arguments.

```{r}
as.call(list(quote(mean), quote(1:10)))
as.call(list(quote(adder(10)), 20))
```

### Exercises

1.  The following two calls look the same, but are actually different:

    ```{r}
    (a <- call("mean", 1:10))
    (b <- call("mean", quote(1:10)))
    identical(a, b)
    ```

    What's the difference? Which one should you prefer?

1.  Implement a pure R version of `do.call()`.

1.  Concatenating a call and an expression with `c()` creates a list. Implement
    `concat()` so that the following code works to combine a call and
    an additional argument.

    ```{r, eval = FALSE}
    concat(quote(f), a = 1, b = quote(mean(a)))
    #> f(a = 1, b = mean(a))
    ```

1.  Since `list()`s don't belong in expressions, we could create a more
    convenient call constructor that automatically combines lists into the
    arguments. Implement `make_call()` so that the following code works.

    ```{r, eval = FALSE}
    make_call(quote(mean), list(quote(x), na.rm = TRUE))
    #> mean(x, na.rm = TRUE)
    make_call(quote(mean), quote(x), na.rm = TRUE)
    #> mean(x, na.rm = TRUE)
    ```

1.  How does `mode<-` work? How does it use `call()`?

1.  Read the source for `pryr::standardise_call()`. How does it work?
    Why is `is.primitive()` needed?

1.  `standardise_call()` doesn't work so well for the following calls.
    Why?

    ```{r}
    standardise_call(quote(mean(1:10, na.rm = TRUE)))
    standardise_call(quote(mean(n = T, 1:10)))
    standardise_call(quote(mean(x = 1:10, , TRUE)))
    ```

1.  Read the documentation for `pryr::modify_call()`. How do you think
    it works? Read the source code.

1.  Use `ast()` and experimentation to figure out the three arguments in an
    `if()` call. Which components are required? What are the arguments to
    the `for()` and `while()` calls?

## Capturing the current call {#capturing-call}

```{r, eval = FALSE, echo = FALSE}
std <- c("package:base", "package:utils", "package:stats")
names(find_uses(std, "sys.call"))
names(find_uses(std, "match.call"))
```

Many base R functions use the current call: the expression that caused the current function to be run. There are two ways to capture a current call: \indexc{calls|capturing current}

* `sys.call()` captures exactly what the user typed. \indexc{sys.call()}

* `match.call()` makes a call that only uses named arguments. It's like
  automatically calling `pryr::standardise_call()` on the result of
  `sys.call()` \indexc{match.call()}

The following example illustrates the difference between the two:

```{r}
f <- function(abc = 1, def = 2, ghi = 3) {
  list(sys = sys.call(), match = match.call())
}
f(d = 2, 2)
```

Modelling functions often use `match.call()` to capture the call used to create the model. This makes it possible to `update()` a model, re-fitting the model after modifying some of original arguments. Here's an example of `update()` in action: \indexc{update()}

```{r}
mod <- lm(mpg ~ wt, data = mtcars)
update(mod, formula = . ~ . + cyl)
```

How does `update()` work? We can rewrite it using some tools from pryr to focus on the essence of the algorithm.

```{r}
update_call <- function (object, formula., ...) {
  call <- object$call

  # Use update.formula to deal with formulas like . ~ .
  if (!missing(formula.)) {
    call$formula <- update.formula(formula(object), formula.)
  }

  modify_call(call, dots(...))
}
update_model <- function(object, formula., ...) {
  call <- update_call(object, formula., ...)
  eval(call, parent.frame())
}
update_model(mod, formula = . ~ . + cyl)
```

The original `update()` has an `evaluate` argument that controls whether the function returns the call or the result. But I think it's better, on principle, that a function returns only one type of object, rather than different types depending on the function's arguments.

This rewrite also allows us to fix a small bug in `update()`: it re-evaluates the call in the global environment, when what we really want is to re-evaluate it in the environment where the model was originally fit --- in the formula.

```{r, error = TRUE}
f <- function() {
  n <- 3
  lm(mpg ~ poly(wt, n), data = mtcars)
}
mod <- f()
update(mod, data = mtcars)

update_model <- function(object, formula., ...) {
  call <- update_call(object, formula., ...)
  eval(call, environment(formula(object)))
}
update_model(mod, data = mtcars)
```

This is an important principle to remember: if you want to re-run code captured with `match.call()`, you also need to capture the environment in which it was evaluated, usually the `parent.frame()`. The downside to this is that capturing the environment also means capturing any large objects which happen to be in that environment, which prevents their memory from being released. This topic is explored in more detail in [garbage collection](#gc). \index{environments|capturing}

Some base R functions use `match.call()` where it's not necessary. For example,  `write.csv()` captures the call to `write.csv()` and mangles it to call `write.table()` instead:

```{r}
write.csv <- function(...) {
  Call <- match.call(expand.dots = TRUE)
  for (arg in c("append", "col.names", "sep", "dec", "qmethod")) {
    if (!is.null(Call[[arg]])) {
      warning(gettextf("attempt to set '%s' ignored", arg))
    }
  }
  rn <- eval.parent(Call$row.names)
  Call$append <- NULL
  Call$col.names <- if (is.logical(rn) && !rn) TRUE else NA
  Call$sep <- ","
  Call$dec <- "."
  Call$qmethod <- "double"
  Call[[1L]] <- as.name("write.table")
  eval.parent(Call)
}
```

To fix this, we could implement `write.csv()` using regular function call semantics:

```{r}
write.csv <- function(x, file = "", sep = ",", qmethod = "double", 
                      ...) {
  write.table(x = x, file = file, sep = sep, qmethod = qmethod, 
    ...)
}
```

This is much easier to understand: it's just calling `write.table()` with different defaults. This also fixes a subtle bug in the original `write.csv()`: `write.csv(mtcars, row = FALSE)` raises an error, but `write.csv(mtcars, row.names = FALSE)` does not. The lesson here is that it's always better to solve a problem with the simplest tool possible.

### Exercises

1.  Compare and contrast `update_model()` with `update.default()`.

1.  Why doesn't `write.csv(mtcars, "mtcars.csv", row = FALSE)` work?
    What property of argument matching has the original author forgotten?

1.  Rewrite `update.formula()` to use R code instead of C code.

1.  Sometimes it's necessary to uncover the function that called the
    function that called the current function (i.e., the grandparent, not
    the parent). How can you use `sys.call()` or `match.call()` to find
    this function?

## Walking the AST with recursive functions {#ast-funs}

It's easy to modify a single call with `substitute()` or `pryr::modify_call()`. For more complicated tasks we need to work directly with the AST. The base `codetools` package provides some useful motivating examples of how we can do this: \index{recursion!over ASTs}

* `findGlobals()` locates all global variables used by a function. This
  can be useful if you want to check that your function doesn't inadvertently
  rely on variables defined in their parent environment.

* `checkUsage()` checks for a range of common problems including
  unused local variables, unused parameters, and the use of partial
  argument matching.

To write functions like `findGlobals()` and `checkUsage()`, we'll need a new tool. Because expressions have a tree structure, using a recursive function would be the natural choice. The key to doing that is getting the recursion right. This means making sure that you know what the base case is and figuring out how to combine the results from the recursive case. For calls, there are two base cases (atomic vectors and names) and two recursive cases (calls and pairlists). This means that a function for working with expressions will look like:

```{r, eval = FALSE}
recurse_call <- function(x) {
  if (is.atomic(x)) {
    # Return a value
  } else if (is.name(x)) {
    # Return a value
  } else if (is.call(x)) {
    # Call recurse_call recursively
  } else if (is.pairlist(x)) {
    # Call recurse_call recursively
  } else {
    # User supplied incorrect input
    stop("Don't know how to handle type ", typeof(x), 
      call. = FALSE)
  }
}
```

### Finding F and T

We'll start simple with a function that determines whether a function uses the logical abbreviations `T` and `F`. Using `T` and `F` is generally considered to be poor coding practice, and is something that `R CMD check` will warn about. Let's first compare the AST for `T` vs. `TRUE`:

```{r}
ast(TRUE)
ast(T)
```

`TRUE` is parsed as a logical vector of length one, while `T` is parsed as a name. This tells us how to write our base cases for the recursive function: while an atomic vector will never be a logical abbreviation, a name might, so we'll need to test for both `T` and `F`. The recursive cases can be combined because they do the same thing in both cases: they recursively call `logical_abbr()` on each element of the object. \indexc{logical\_abbr}

```{r}
logical_abbr <- function(x) {
  if (is.atomic(x)) {
    FALSE
  } else if (is.name(x)) {
    identical(x, quote(T)) || identical(x, quote(F))
  } else if (is.call(x) || is.pairlist(x)) {
    for (i in seq_along(x)) {
      if (logical_abbr(x[[i]])) return(TRUE)
    }
    FALSE
  } else {
    stop("Don't know how to handle type ", typeof(x), 
      call. = FALSE)
  }
}

logical_abbr(quote(TRUE))
logical_abbr(quote(T))
logical_abbr(quote(mean(x, na.rm = T)))
logical_abbr(quote(function(x, na.rm = T) FALSE))
```

### Finding all variables created by assignment

`logical_abbr()` is very simple: it only returns a single `TRUE` or `FALSE`. The next task, listing all variables created by assignment, is a little more complicated. We'll start simply, and then make the function progressively more rigorous. \indexc{find\_assign()}

Again, we start by looking at the AST for assignment:

```{r}
ast(x <- 10)
```

Assignment is a call where the first element is the name `<-`, the second is the object the name is assigned to, and the third is the value to be assigned. This makes the base cases simple: constants and names don't create assignments, so they return `NULL`. The recursive cases aren't too hard either. We `lapply()` over pairlists and over calls to functions other than `<-`.

```{r}
find_assign <- function(x) {
  if (is.atomic(x) || is.name(x)) {
    NULL
  } else if (is.call(x)) {
    if (identical(x[[1]], quote(`<-`))) {
      x[[2]]
    } else {
      lapply(x, find_assign)
    }
  } else if (is.pairlist(x)) {
    lapply(x, find_assign)
  } else {
    stop("Don't know how to handle type ", typeof(x), 
      call. = FALSE)
  }
}
find_assign(quote(a <- 1))
find_assign(quote({
  a <- 1
  b <- 2
}))
```

This function works for these simple cases, but the output is rather verbose and includes some extraneous `NULL`s. Instead of returning a list, let's keep it simple and use a character vector. We'll also test it with two slightly more complicated examples:

```{r}
find_assign2 <- function(x) {
  if (is.atomic(x) || is.name(x)) {
    character()
  } else if (is.call(x)) {
    if (identical(x[[1]], quote(`<-`))) {
      as.character(x[[2]])
    } else {
      unlist(lapply(x, find_assign2))
    }
  } else if (is.pairlist(x)) {
    unlist(lapply(x, find_assign2))
  } else {
    stop("Don't know how to handle type ", typeof(x), 
      call. = FALSE)
  }
}

find_assign2(quote({
  a <- 1
  b <- 2
  a <- 3
}))

find_assign2(quote({
  system.time(x <- print(y <- 5))
}))
```

This is better, but we have two problems: dealing with repeated names and neglecting assignments inside other assignments. The fix for the first problem is easy. We need to wrap `unique()` around the recursive case to remove duplicate assignments. The fix for the second problem is a bit more tricky. We also need to recurse when the call is to `<-`. `find_assign3()` implements both strategies:

```{r}
find_assign3 <- function(x) {
  if (is.atomic(x) || is.name(x)) {
    character()
  } else if (is.call(x)) {
    if (identical(x[[1]], quote(`<-`))) {
      lhs <- as.character(x[[2]])
    } else {
      lhs <- character()
    }

    unique(c(lhs, unlist(lapply(x, find_assign3))))
  } else if (is.pairlist(x)) {
    unique(unlist(lapply(x, find_assign3)))
  } else {
    stop("Don't know how to handle type ", typeof(x), 
      call. = FALSE)
  }
}

find_assign3(quote({
  a <- 1
  b <- 2
  a <- 3
}))

find_assign3(quote({
  system.time(x <- print(y <- 5))
}))
```

We also need to test subassignment:

```{r}
find_assign3(quote({
  l <- list()
  l$a <- 5
  names(l) <- "b"
}))
```

We only want assignment of the object itself, not assignment that modifies a property of the object. Drawing the tree for the quoted object will help us see what condition to test for. The second element of the call to `<-` should be a name, not another call.

```{r}
ast(l$a <- 5)
ast(names(l) <- "b")
```

Now we have a complete version:

```{r}
find_assign4 <- function(x) {
  if (is.atomic(x) || is.name(x)) {
    character()
  } else if (is.call(x)) {
    if (identical(x[[1]], quote(`<-`)) && is.name(x[[2]])) {
      lhs <- as.character(x[[2]])
    } else {
      lhs <- character()
    }

    unique(c(lhs, unlist(lapply(x, find_assign4))))
  } else if (is.pairlist(x)) {
    unique(unlist(lapply(x, find_assign4)))
  } else {
    stop("Don't know how to handle type ", typeof(x), 
      call. = FALSE)
  }
}

find_assign4(quote({
  l <- list()
  l$a <- 5
  names(l) <- "b"
}))
```

While the complete version of this function is quite complicated, it's important to remember we wrote it by working our way up by writing simple component parts.

### Modifying the call tree {#modifying-code}

The next step up in complexity is returning a modified call tree, like what you get with `bquote()`. `bquote()` is a slightly more flexible form of quote: it allows you to optionally quote and unquote some parts of an expression (it's similar to the backtick operator in Lisp). Everything is quoted, _unless_ it's encapsulated in `.()` in which case it's evaluated and the result is inserted: \index{bquote()}

```{r}
a <- 1
b <- 3
bquote(a + b)
bquote(a + .(b))
bquote(.(a) + .(b))
bquote(.(a + b))
```

This provides a fairly easy way to control what gets evaluated and when. How does `bquote()` work? Below, I've rewritten `bquote()` to use the same style as our other functions: it expects input to be quoted already, and makes the base and recursive cases more explicit:

```{r}
bquote2 <- function (x, where = parent.frame()) {
  if (is.atomic(x) || is.name(x)) {
    # Leave unchanged
    x
  } else if (is.call(x)) {
    if (identical(x[[1]], quote(.))) {
      # Call to .(), so evaluate
      eval(x[[2]], where)
    } else {
      # Otherwise apply recursively, turning result back into call
      as.call(lapply(x, bquote2, where = where))
    }
  } else if (is.pairlist(x)) {
    as.pairlist(lapply(x, bquote2, where = where))
  } else {
    # User supplied incorrect input
    stop("Don't know how to handle type ", typeof(x), 
      call. = FALSE)
  }
}

x <- 1
y <- 2
bquote2(quote(x == .(x)))
bquote2(quote(function(x = .(x)) {
  x + .(y)
}))
```

The main difference between this and the previous recursive functions is that after we process each element of calls and pairlists, we need to coerce them back to their original types.

Note that functions that modify the source tree are most useful for creating expressions that are used at run-time, rather than those that are saved back to the original source file. This is because all non-code information is lost:

```{r}
bquote2(quote(function(x = .(x)) {
  # This is a comment
  x +  # funky spacing
    .(y)
}))
```

These tools are somewhat similar to Lisp macros, as discussed in [Programmer's Niche: Macros in R](http://www.r-project.org/doc/Rnews/Rnews_2001-3.pdf#page=10) by Thomas Lumley. However, macros are run at compile-time, which doesn't have any meaning in R, and always return expressions. They're also somewhat like Lisp [fexprs](http://en.wikipedia.org/wiki/Fexpr). A fexpr is a function where the arguments are not evaluated by default. The terms macro and fexpr are useful to know when looking for useful techniques from other languages. \index{macros} \index{fexprs}

### Exercises

1.  Why does `logical_abbr()` use a for loop instead of a functional
    like `lapply()`?

1.  `logical_abbr()` works when given quoted objects, but doesn't work when
    given an existing function, as in the example below. Why not? How could
    you modify `logical_abbr()` to work with functions? Think about what
    components make up a function.

    ```{r, eval = FALSE}
    f <- function(x = TRUE) {
      g(x + T)
    }
    logical_abbr(f)
    ```

1.  Write a function called `ast_type()` that returns either "constant",
    "name", "call", or "pairlist". Rewrite `logical_abbr()`, `find_assign()`,
    and `bquote2()` to use this function with `switch()` instead of nested if
    statements.

1.  Write a function that extracts all calls to a function. Compare your
    function to `pryr::fun_calls()`.

1.  Write a wrapper around `bquote2()` that does non-standard evaluation
    so that you don't need to explicitly `quote()` the input.

1.  Compare `bquote2()` to `bquote()`. There is a subtle bug in `bquote()`:
    it won't replace calls to functions with no arguments. Why?

    ```{r}
    bquote(.(x)(), list(x = quote(f)))
    bquote(.(x)(1), list(x = quote(f)))
    ```

1.  Improve the base `recurse_call()` template to also work with lists of
    functions and expressions (e.g., as from `parse(path_to_file))`.
