# Conditions

```{r, include = FALSE}
source("common.R")
```

## Introduction

The job of the __condition__ system is to alert the user to problems and give them tools to handle them. In R, conditions encompass errors (`stop()`), warnings (`warning()`), and messages (`message()`), and other custom classes. It's important to understand the condition system because in your own code you'll need to both __signal__ conditions to the user, and to __handle__ conditions thrown by other functions.

R offers an exceptionally powerful condition handling system based on ideas from Common Lisp, but it's currently not very well documented or often used. This chapter will introduce you to the most important basics, but if you want to learn more, I recommend the following two sources:

* [_A prototype of a condition system for R_][prototype] by Robert Gentleman 
  and Luke Tierney. This describes an early version of R's condition system. 
  While the implementation has changed somewhat since this document was 
  written, it provides a good overview of how the pieces fit together, and 
  some motivation for its design.

* [_Beyond exception handling: conditions and restarts_][beyond-handling]
  by Peter Seibel. This describes exception handling in Lisp, which happens 
  to be very similar to R's approach. It provides useful motivation and 
  more sophisticated examples. I have provided an R translation of the 
  chapter at <http://adv-r.had.co.nz/beyond-exception-handling.html>.

### Overview {-}

* Discuss the details of signalling conditions
* Show the basic tools for ignoring conditions
* Dive into condition handlers, including the details of condition objects.
* Show a bunch of smaller applications.

### Quiz {-}

Want to skip this chapter? Go for it, if you can answer the questions below. Find the answers at the end of the chapter in [answers](#conditions-answers).

1. What are the four most important types of condition?

1. What function do you use to ignore errors in block of code?

1. Why might you want to create an error with a custom S3 class?

### Prerequisites

```{r setup}
library(rlang)

# Wrappers while we wait for rlang
cnd_type <- function(cnd) {
  if (inherits(cnd, "error")) {
    "error"
  } else if (inherits(cnd, "warning")) {
    "warning"
  } else if (inherits(cnd, "message")) {
    "message"
  } else {
    "condition"
  }
}

cnd_signal <- function(cnd) {
  switch(cnd_type(cnd), 
    error = cnd_abort(cnd),
    warning = cnd_warn(cnd),
    message = cnd_inform(cnd),
    condition = rlang::cnd_signal(cnd, .mufflable = TRUE)
  )
}

cnd_muffle <- function(cnd) {
  switch(cnd_type(cnd),
    error = abort("Can not muffle an error"),
    warning = invokeRestart("muffleWarning"),
    message = invokeRestart("muffleMessage"),
    condition = invokeRestart("muffle")
  )
}

abort <- function(.msg, .type = NULL, ...) {
  cnd <- error_cnd(.type = .type, ..., .msg = .msg, call = NULL)
  stop(cnd)
}

# Waiting for lobstr update
cst <- function() print(rlang::calltrace(globalenv()))

```

## Signalling conditions
\index{errors!throwing}
\index{conditions!signalling}

There are three conditions that you can signal in code: errors, warnings, and messages. Errors are the most severe; they indicate that there is no way for a function to continue and force execution to terminate. Messages are the mildest; they are way of informing the user that some action has been performed. Warnings fall somewhat in between, and typically indicate that something has gone wrong but the function has been able to recover in some way. There is a final condition that can only be generated interactively: an interrupt, which indicates that the user has "interrupted" execution by pressing Escape, Ctrl + Break, or Ctrl + C (depending on the platform).

Conditions are usually displayed prominently, in a bold font or coloured red, depending on the R interface. You can tell them apart because errors always start with "Error", warnings with "Warning message", and messages with nothing.

```{r, error = TRUE}
stop("This is what an error looks like")

warning("This is what a warning looks like")

message("This is what a message looks like")
```

### Errors

In base R, errors are signalled, or __thrown__, by `stop()`:

```{r, error = TRUE}
f <- function() g()
g <- function() h()
h <- function() stop("This is an error!")

f()
```

By default, the error message includes the call, but this is typically not useful (and recapitulates information that you can easily get from `traceback()`), so I think it's good practice to use `call. = FALSE`:

```{r, error = TRUE}
h <- function() stop("This is an error!", call. = FALSE)
f()
```

The rlang equivalent, `rlang::abort()`, does not capture the call by default:

```{r, error = TRUE}
h <- function() abort("This is an error!")
f()
```

We'll use of `abort()` throughout this chapter, but we won't get to its most compelling feature, the ability to add additional metadata to the condition object, until we're near the end of the chapter.

The best error messages tell you what is wrong, and point you in the right direction for fixing the problem. Writing good error messages is extremely hard because usually the error occurs because the funtion caller has an incorrect mental model of the function. As a developer it's hard to predict how the mental model might be incorrect, especially when you've just written the function and the details are fresh in your mind. Often writing a good error message takes some iteration - you have to wait until someone sees it in the wild, and then you can talk to them to figure out what went wrong in their prediction of your function's behaviour. There is some evolving advice of writing good error messages in the tidyverse style guide: <http://style.tidyverse.org/error-messages.html>.

### Warnings

Warnings are weaker than errors: they signal that something has gone wrong, but the code has been able to recover and continue. They are generated by `warning()`. Unlike errors, you can have multiple warnings from a single function call:

```{r}
f <- function() {
  cat("1\n")
  warning("W1")
  cat("2\n")
  warning("W2")
  cat("3\n")
  warning("W3")
}
```

By default, warnings are cached and printed only when control returns to the top level:

```{r, eval = FALSE}
f()
#> 1
#> 2
#> 3
#> Warning messages:
#> 1: In f() : W1
#> 2: In f() : W2
#> 3: In f() : W3
```

You can control this behaviour with the `warn` option:

* To have warnings to appear immediately, set `options(warn = 1)`. 

* To turn warnings into errors, set `options(warn = 2)`. This is usually
  the easiest way to debug a warning, as once it's an error you can
  use tools like `traceback()` to find the source. 

* Restore the default behaviour with `option(warn = 0)`.

Like `stop()`, `warning()` also has a call argument. It is slightly more useful (since warnings are often more distant from their source), but I still generally suppress it with `call. = FALSE`. The rlang wrapper, `rlang::warn()`, also suppresses by default.

Warnings occupy a somewhat awkward limnal state between messages ("you should know about this") and errors ("you must fix this"). You should be cautious with your use of `warnings()`: warnings are easy to miss if there's a lot of other output, and you don't want your function to recover too easily from clearly incorrect input. In my opinion, base R tends to overuse warnings, and many warnings in base R would be better off as clear errors. For example, take `read.csv()`, which uses the the `file()` function. The file function simple warns if the file exists. That means that when you try and read a file that does not exist, you get both a warning and an error:

```{r, error = TRUE}
read.csv("blah.csv")
```

There are a few cases where warnings are clearly useful:

* When deprecating a function. A deprecated function still works, but you want
  to ensure users know that they need to transition to a new approach. A 
  deprecation warning should always help the user do the right thing.

* When you are reasonably certain you can recover from a problem.
  If you were 100% certain that you could fix the problem, you wouldn't need 
  any message; if you were uncertain that you could correctly fix the issue, 
  you'd throw an error.

### Messages

Messages are informational; use them to inform the user of important actions that you function has taken on their behalf. The key principle is don't try and be silently helpful. 

* Determining a function default that requires a lot of computation.
  For example, 
  
Generally any function that produces a message should have some way to suppress it, like `quiet = TRUE`. The user can always use `suppressMessages()`, but this the nuclear option as it suppresses all messages regardless of source.

Generally, you should use `message()` rather than `cat()` or `print()` for informing the user about actions that your function has taken. This is useful, for example, if you've had to do non-trivial computation to determine the default value of an argument, and you want to let the user know exactly what you've done. Messages are a side-channel. Functions that are primarily called for their printed output (like `print()` or `str()` methods), should use `cat()` to write directly to the console. 

Messages are also important when developing packages: when displaying messages during package startup (i.e. in `.onAttach()`), you must use `packageStartupMessage()`.

### Exercises

## Ignoring conditions 
\indexc{try()} \indexc{suppressWarnings()} \indexc{suppressMessages()}  

The simplest way of handling conditions in R is to simply ignore them. These are the bluntest instruments, but because they require little knowledge of the condition system, they're a good place to start.

`try()` allows execution to continue even after an error has occurred. Normally if you run a function that throws an error, it terminates immediately and doesn't return a value: 

```{r, error = TRUE}
f1 <- function(x) {
  log(x)
  10
}
f1("x")
```

However, if you wrap the statement that creates the error in `try()`, the error message will be printed but execution will continue:

```{r, eval = FALSE}
f2 <- function(x) {
  try(log(x))
  10
}
f2("a")
#> Error in log(x) : non-numeric argument to mathematical function
#> [1] 10
```

(You can suppress the message with `try(..., silent = TRUE)`.)

A useful `try()` pattern is to do assignment inside: this lets you define a default value to be used if the code does not succeed.

```{r, eval = FALSE}
default <- NULL
try(default <- read.csv("possibly-bad-input.csv"), silent = TRUE)
```

It is possible, but not recommended, to save the result of `try()` and perform different actions based on whether or not the result has class `try-error`. Instead, it is better to use `tryCatch()`, which we'll get to shortly.

There are two functions that are analagous to `try()` for warnings and messages: `suppressWarnings()` and `suppressMessages()`. These allow you to suppress all warnings and messages generated by a block of code.

```{r}
suppressWarnings({
  warning("Uhoh!")
})

suppressMessages({
  message("Hello there")
})
```

These functions are heavy handed: you can't use them to suppress a single warning that you know about, while allowing other warnings that you don't know about to pass through. We'll come back to that challenge later in the chapter.

### Exercises



## Handling conditions
\index{errors!catching}
\index{conditions!handling}

`tryCatch()` and `withCallingHandlers()` are general tools for handling conditions. They allows you to map conditions to __handlers__, functions that are called with the condition as an argument. `tryCatch()` and `withCallingHandlers()` differ in the type of handlers they create:

*  `tryCatch()` defines __exiting__ handlers; after the condition is captured 
   control returns to the context where `tryCatch()` was called. This makes 
   `tryCatch()` most suitable for working with errors and interrupts, as these 
   have to exit the code anyway.
  
*  `withCallingHandlers()` defines __in-place__ handlers; after the condition
   is captued control returns to the context where the condition was signalled.
   This makes it most suitable for working with non-error conditions.

But before we can learn about these handlers, we need to talk a little bit about condition __objects__.  In simple usage, you never need to think about these objects, but they become explicit when you start working with the handlers.

### Condition objects
\index{conditions!objects}

So far we've signalled conditions, but haven't looked at the objects created as part of that process. The easiest way to see what's going on is to capture a condition object with `rlang::catch_cnd()`:

```{r}
library(rlang)
c <- catch_cnd(abort("An error"))
str(c)

typeof(c)
attr(c, "class")
```

Conditions are S3 objects (which you'll learn more about [S3]). Fortunately, conditions are very simple objects so you don't need to know anything about S3 to work with them: a condition is just a list with a special `class` attribute (that attribute is what makes it an S3 object). Every condition object must have two components in the list: the `message` (the text display to a user), and the `call` (which as described above, we don't use).

```{r}
c$message
c$call
```

Conditions can contain other elements, which we'll discuss shortly in in [custom conditions].

### Exiting handlers
\indexc{tryCatch()} \index{handlers!exiting}

Each condition has some default behaviour: errors stop execution and return to the top-level, warnings are captured, and messages are display. `tryCatch()` allows us to temporarily override the default behaviour and do something else. 

The basic form of `tryCatch()` is shown below. The named arguments set up handlers to be called when the unnamed argument, the `expr` to evaluate, is executed. The name is usually one of `error`, `warning`, `message`, or `interrupt` (the components of the condition class), and the function will be called with a single object, the condition.

```{r}
tryCatch(
  error = function(c) 10,
  stop("This is an error!")
)
```

If no conditions are signalled, or the signalled condition does not match the handler name, the code executes normally:

```{r}
tryCatch(
  error = function(c) 10,
  1 + 1
)

tryCatch(
  error = function(c) 10,
  {
    message("Hi!")
    1 + 1
  }
)
```

The handlers set up by `tryCatch()` are called __exiting__ handlers because after the condition is signal, control passes to the handler and never returns to the original code, effectively meaning that the code "exits":

```{r}
tryCatch(
  message = function(c) "There",
  {
    message("Here")
    stop("This code is never run!")
  }
)
```

The argument to the handler is the condition object (hence, by convention, I use the name `c`). This is only moderately useful for the base conditions because they only have `message` and `call` fields. As we'll see shortly, it's more useful when you make your own custom conditions.

```{r}
tryCatch(
  error = function(c) c$message,
  stop("This is an error")
)
```

`tryCatch()` has one other argument: `finally`. It specifies a block of code (not a function) to run regardless of whether the initial expression succeeds or fails. This can be useful for clean up (e.g., deleting files, closing connections). This is functionally equivalent to using `on.exit()` (and indeed that's how it's implemented) but it can wrap smaller chunks of code than an entire function. \indexc{on.exit()}

### In-place handlers
\index{handlers!in-place}

The handlers set up by `tryCatch()` are called exiting, because they cause code to exit once the condition has been caught. By contrast, the handlers set up by `withCallingHandler()` are __in-place__: code execution will continue normally once the handler returns. This tends to make `withCallingHandlers()` a more natural pairing with the non-error conditions.

```{r}
tryCatch(
  message = function(c) cat("Caught a message!\n"), 
  {
    message("Someone there?")
    message("Why, yes!")
  }
)

withCallingHandlers(
  message = function(c) cat("Caught a message!\n"), 
  {
    message("Someone there?")
    message("Why, yes!")
  }
)
```

The return value of an in-place handler is ignored; they are only useful for their side-effects.

Handlers apply to the wrapped code, not code in the handlers. That means you don't need to worry about getting caught in an infinite loop:

```{r}
withCallingHandlers(
  message = function(c) message("Second message"),
  message("First message")
)
```

By default, a condition will continue to propogate to parent handlers, all the way up to the default handler (or an exiting handler, if provided):

```{r}
# Bubbles all the way up to default handler which generates the message
withCallingHandlers(
  message = function(c) cat("Level 2\n"),
  withCallingHandlers(
    message = function(c) cat("Level 1\n"),
    message("Hello")
  )
)

# Bubbles up to tryCatch
tryCatch(
  message = function(c) cat("Level 2\n"),
  withCallingHandlers(
    message = function(c) cat("Level 1\n"),
    message("Hello")
  )
)
```

If you want to prevent the condition "bubbling up" but still run the rest of the code in the block, you need to explicitly muffle it with `rlang::cnd_muffle()`:

```{r}
withCallingHandlers(
  message = function(c) {
    cat("Level 2\n")
    cnd_muffle(c)
  },
  withCallingHandlers(
    message = function(c) cat("Level 1\n"),
    message("Hello")
  )
)

withCallingHandlers(
  message = function(c) cat("Level 2\n"),
  withCallingHandlers(
    message = function(c) {
      cat("Level 1\n")
      cnd_muffle(c)
    },
    message("Hello")
  )
)
```

### Calling environments

To complete the section, there are some subtle differences between the calling environment of exiting and in-place handlers. This generally is not important, unless you need to capture call stacks. Generally, you can skip this section without worrying about it.

* `withCallingHandlers()`: handlers are called in the context of the call that 
  signalled the condition
  
* `tryCatch()`: handlers are called in the context the call to `tryCatch()`. 

We can see this most easily by using `lobstr::cst()`

```{r, error = TRUE}
f <- function() g()
g <- function() h()
h <- function() stop("!")

tryCatch(
  f(), 
  error = function(e) cst()
)

withCallingHandlers(
  f(), 
  error = function(e) cst()
)
```

### Exercises

1.  Predict the results of evaluating the following code

    ```{r, eval = FALSE}
    show_condition <- function(code) {
      tryCatch(
        error = function(c) "error",
        warning = function(c) "warning",
        message = function(c) "message",
        {
          code
          NULL
        }
      )
    }
    
    show_condition(stop("!"))
    show_condition(10)
    show_condition(warning("?!"))
    show_condition({
      10
      message("?")
      warning("?!")
    })
    ```

1.  Read the source code for `catch_cnd()` and explain how it works.

1.  How could you rewrite `show_condition()` to use a single handler.

## Custom conditions
\index{conditions!custom}

One of the challenges of error handling in R is that, to date, most functions on generate the default conditions which consistent only of a `message` and `call`. This makes detecting a specific error challenging because you have to at the text of the error message. This is error prone, not only because the message might change over time, but also because messages can be translated into other languages. 

Fortunately, however, R has a powerful but little used feature: the ability to use custom condition objects. It is somewhat fiddly to create custom conditions in base R, but rlang makes it very easy: in `rlang::abort()` you can supply a custom `type` and additional metadata. 

Chicken and egg situation.

Currently, the primary motivation for creating your own conditions is to make it easier to test your own code, because you are no longer forced to use string comparisons of the error message. Over time, however, as more people learn about and master the condition system, custom conditions will make it easier for the user to take different actions for different types of errors. For example, "expected" errors (like a model failing to converge for some input datasets) can be silently ignored, while unexpected errors (like no disk space available) can be propagated to the user.

### Motivation

```{r, error = TRUE}
log(letters)
log(1:10, base = letters)
```

We want something a little more informative.

```{r}
log <- function(x, base = exp(1)) {
  if (!is.numeric(x)) {
    stop("`x` must be numeric; not ", typeof(x), call. = FALSE)
  }
  if (!is.numeric(base) || length(base) != 1) {
    stop("`base` must be a single number", call. = FALSE)
  }

  base::log(x, base = base)
}
```

```{r, error = TRUE}
log(letters)
log(1:10, base = letters)
```

This is a big improvement from the interactive point of view - the error messages are much more likely to yield a correct fix. However, from the programming point of a view, it's not a big win - all the data is jammed into a string. This makes it hard to program with; and importantly it also makes our code harder to test.

### Signalling

So let's build some infrastructure to improve this problem.

Let's illustrate this idea with an example. Let's provide a helper for a common type of issue: supplying the wrong type on input to a function. There are three key arguments: 

* `arg`: the name of the argument
* `must`: a description of what the argument must be 
* `not`: a description of what the argument actually is

The pattern is fairly simple. We create a nice error message for the user, using `glue::glue()`, and store metadata for the developer. 

This is a little over-generalised for the example at hand, but it reflects common patterns that I've seen across other functions.

```{r}
abort_bad_argument <- function(arg, must, not = NULL) {
  msg <- glue::glue("`{arg}` must {must}")
  if (!is.null(not)) {
    msg <- glue::glue("{msg}; not {not}")
  }
  abort(msg, "error_bad_argument", arg = arg, must = must, not = not)
}
```

This means we still get a nice

```{r, error = TRUE}
abort_bad_argument("x", must = "be numeric")
```

(Note that you can define a method for the `conditionMessage()` generic instead of generating a message at creation time. This is usually of limited utility because to fulfill the specification of the condition object you must always provide a text `message`.)

With this in hand, we can now rewrite `my_log()`

```{r}
log <- function(x, base = exp(1)) {
  if (!is.numeric(x)) {
    abort_bad_argument("x", must = "be numeric", not = typeof(x))
  }
  if (!is.numeric(base) || length(base) != 1) {
    abort_bad_argument("base", must = "be a single number")
  }

  log(x)
}
```

The code is not much shorter, but I think more semantically meaningful.

This yields the same interactive error messages as before:

```{r, error = TRUE}
log(letters)
log(1:10, base = letters)
```

But importantly yields better structured objects that are easier to program with.

### Handling

```{r}
cnd <- catch_cnd(my_log(1:10, base = "x"))
str(cnd)
```

```{r}
cnd <- catch_cnd(my_log("a"))
str(cnd)
```

Note that when using `tryCatch()` with multiple handlers and custom classes, the first handler to match any class in the signal's class hierarchy is called, not the best match. For this reason, you need to make sure to put the most specific handlers first:

```{r}
tryCatch(my_log("a"),
  error = function(c) "???",
  error_bad_argument = function(c) "bad_argument"
)

tryCatch(my_log("a"),
  error_bad_argument = function(c) "bad_argument",
  error = function(c) "???"
)
```

## Applications {#condition-applications}

What can you do with thse tools? The following section exposes some come use cases.

### Catching a condition

```{r}
# Captures error object
c <- catch_cnd(stop("An error"))
c
str(c)

# Captures first condition
c <- catch_cnd({
  warning("First")
  warning("Second")
})
c

# No condition, so returns NULL
catch_cnd(1 + 2)
```

### Replacement value

You can use `tryCatch()` to implement `try()`. A simple implementation is shown below. `base::try()` is more complicated in order to make the error message look more like what you'd see if `tryCatch()` wasn't used. Note the use of `conditionMessage()` to extract the message associated with the original error.

```{r}
fail_with <- function(expr, value = NULL) {
  tryCatch(expr, error = function(c) value)
}
```

```{r}
try2 <- function(code, silent = FALSE) {
  tryCatch(code, error = function(c) {
    msg <- conditionMessage(c)
    if (!silent) {
      message(msg)
    }
    structure(msg, class = "try-error")
  })
}

try2(1)

try2(stop("Hi"))

try2(stop("Hi"), silent = TRUE)
```

```{r}
does_error <- function(expr) {
  tryCatch({
    expr
    FALSE
  }, error = function(c) TRUE)
}
```

```{r}
safety <- function(expr) {
  tryCatch(
    {
      list(result = expr, error = NULL)
    }, 
    function(e) {
      list(result = NULL, error = e)
    }
  )
}
```



### Resignal

As well as returning default values when a condition is signalled, handlers can be used to make more informative error messages. For example, by modifying the message stored in the error condition object, the following function wraps `read.csv()` to add the file name to any errors:

```{r, error = TRUE, warning = FALSE}
read.csv2 <- function(file, ...) {
  tryCatch(read.csv(file, ...), error = function(c) {
    message <- paste0(c$message, " (in ", file, ")")
    abort(message)
  })
}
read.csv("code/dummy.csv")
read.csv2("code/dummy.csv")
```

Update to use whatever `rethrow()` becomes.

### Record

This is what the evaluate package does. It powers knitr. (A little more complicated because it also has to handle output which uses a different system.)

### Return early

```{r}
try_parse_eval <- function(x, env = globalenv()) {
  expr <- tryCatch(parse(text = text), error = function(e) NULL)
  if (is.null(expr)) {
    return(NULL)
  }

  res <- tryCatch(eval(expr, env), error = function(e) NULL)
  if (is.null(res)) {
    return(res)
  }
  
  ...
}

try_parse_eval <- function(x, env = globalenv()) {
  expr <- tryCatch(parse(text = text), error = function(e) return_from(NULL))
  res <- tryCatch(eval(expr, env), error = function(e) return_from(NULL))
  ...
}
```

### Performance {#cond-perf}

```{r, eval = FALSE}
f1 <- function(n) {
  for (i in seq_len(n)) {
    cat("Hi")
  }
}

f2 <- function(n) {
  for (i in seq_len(n)) {
    message("Hi")
  }
}



bench::mark(
  signalCondition(cnd),
  handled = withCallingHandlers(signalCondition(cnd), silent = function(c) {})
)


bench::mark(
  cat("Hello\n"),
  message("Hello"), 
  handler = withCallingHandlers(message("Hello"), message = function(e) {})
)
```

### Muffle

Due to the way that restarts are implemented in R, the ability to muffle, or ignore a condition (so it doesn't bubble up to other handlers) is defined by the function that signals the condition.  `message()` and `warning()` automatically setup muffle handlers, but `signalCondition()` does not.

`cnd_signal()` ensures that a muffler is always set up. `cnd_muffle(c)` always picks the right muffler depending on the class of the condition.

Log messages to disk example.

```{r}
write_line <- function(path, ...) {
  cat(..., "\n", file = path, append = TRUE, sep = "")
}

log_messages <- function(expr, path) {
  
  withCallingHandlers(expr,
    message = function(c) {
      write_line(path, "[MESSAGE] ", c$message)
      cnd_muffle(c)
    })
}
```


### Exercises

1.  Compare the following two implementations of `message2error()`. What is the
    main advantage of `withCallingHandlers()` in this scenario? (Hint: look
    carefully at the traceback.)

    ```{r}
    message2error <- function(code) {
      withCallingHandlers(code, message = function(e) stop(e))
    }
    message2error <- function(code) {
      tryCatch(code, message = function(e) stop(e))
    }
    ```

1.  Why is catching interrupts dangerous? Run this code to find out.

    ```{r, eval = FALSE}
    bottles_of_beer <- function(i = 99) {
      message("There are ", i, " bottles of beer on the wall, ", i, " bottles of beer.")
      while(i > 0) {
        tryCatch(
          Sys.sleep(1),
          interrupt = function(err) {
            i <<- i - 1
            if (i > 0) {
              message(
                "Take one down, pass it around, ", i, 
                " bottle", if (i > 1) "s", " of beer on the wall."
              )
            }
          }
        )
      }
      message("No more bottles of beer on the wall, no more bottles of beer.")
    }
    ```



## Quiz answers {#conditions-answers}

1. You could use `try()` or `tryCatch()`.

1. Because you can then capture specific types of error with `tryCatch()`,
   rather than relying on the comparison of error strings, which is risky,
   especially when messages are translated.

[prototype]: http://homepage.stat.uiowa.edu/~luke/R/exceptions/simpcond.html
[beyond-handling]: http://www.gigamonkeys.com/book/beyond-exception-handling-conditions-and-restarts.html
