---
title: Data structures
layout: default
output: bookdown::html_chapter
---

# Data structures

This chapter summarises the most important data structures in base R. I assume you've used many (if not all) of them before, but you may not have thought deeply about how they are interrelated. In this brief overview, I won't discuss individual types in depth. Instead, I'll show you how they fit together as a whole. If you need more details, you can find them in R's documentation.

R's base data structures can be organised by their dimensionality (1d, 2d, or nd) and whether they're homogeneous (all contents must be of the same type) or heterogeneous (the contents can be of different types):

|    | Homogeneous   | Heterogeneous |
|----|---------------|---------------|
| 1d | Atomic vector | List          |
| 2d | Matrix        | Data frame    |
| nd | Array         |               |

Almost all other objects are built upon these foundations, and in [OO field guide](#oo-field-guide) you'll see how more complicated objects are built of these simple pieces. Note that R has 0-dimensional, or scalar types. Individual numbers or strings, which you might think would be scalars, are actually vectors of length one. 

Given an object, the best way to understand what data structures it's composed of is to use `str()`. `str()` is short for structure and it gives a compact, human readable description of any R data structure.

##### Quiz

Take this short quiz to determine if you need to read this chapter. If the answers quickly come to mind, you can comfortably skip this chapter. You can check your answers in [answers](#data-structure-answers).

1. What are the three properties of a vector? (apart from its contents).

1. What are the four common types of atomic vectors? What are the two 
   rare types?

1. What are attributes? How do you get and set them?

1. How is a list different from a vector? How is a matrix different from 
   a data frame?

1. Can you have a list that is a matrix? Can a data frame have a column 
   that is a list?

##### Outline

The chapter starts by describing R's 1d structures (atomic vectors and lists), then detours to discuss attributes (R's flexible metadata specification) and factors, before returning to discuss high-d structures (matrices, arrays and data frames).

## Vectors

The basic data structure in R is the vector, which comes in two flavours: atomic vectors and lists. They differ in their content: an atomic vector has elements of the same type, list can have elements of different types. As well as their content, vectors have three properties: `typeof()` (what it is), `length()` (how long it is) and `attributes()` (additional arbitrary metadata).

NB: `is.vector()` does not test if an object is a vector. Instead it returns `TRUE` only if the object is a vector with no attributes apart from names. Use `is.atomic(x) || is.list(x)` to test if an object is actually a vector.

### Atomic vectors

Atomic vectors can be logical, integer, double (often called numeric), or character, or less commonly complex or raw. Atomic vectors are usually created with `c()`, short for combine:

```{r}
num_var <- c(1, 2.5, 4.5)
# With the L suffix, you get an integer rather than a double
int_var <- c(1L, 6L, 10L)
# Use TRUE and FALSE (or T and F) to create logical vectors
log_var <- c(TRUE, FALSE, T, F)
chr_var <- c("these are", "some strings")
```

Atomic vectors are flat, even if you nest `c()`'s:

```{r}
c(1, c(2, c(3, 4)))
# the same as
c(1, 2, 3, 4)
```

Missing values are specified with `NA`, which is a logical vector of length 1. `NA` will always be coerced to the correct type with `c()`, or you can create NA's of specific types with `NA_real_` (double), `NA_integer_` and `NA_character_`.

#### Types and tests

Given a vector, you can determine its type with `typeof()`, or check if it's a specific type with an "is" function: `is.character()`, `is.double()`, `is.integer()`, `is.logical()`, or, more generally, `is.atomic()`.

```{r}
int_var <- c(1L, 6L, 10L)
typeof(int_var)
is.integer(int_var)

num_var <- c(1, 2.5, 4.5)
typeof(num_var)
is.double(num_var)
```

NB: `is.numeric()` is a general test for the "numberliness" of a vector and returns `TRUE` for both integer and double vectors. It is not a specific test for double vectors, which are often called numeric.

```{r}
is.numeric(int_var)
is.numeric(num_var)
```

#### Coercion

An atomic vector can only be of one type, so when you attempt to combine different types they will be __coerced__ to the most flexible type. Types from least to most flexible are: logical, integer, double and character.

```{r}
str(c("a", 1))
```

When a logical vector is coerced to an integer or double, `TRUE` becomes 1 and `FALSE` becomes 0. This is very useful in conjunction with `sum()` and `mean()`

```{r}
as.numeric(c(FALSE, FALSE, TRUE))
# Total number of TRUEs
sum(mtcars$cyl == 4)
# Proportion of TRUEs
mean(mtcars$cyl == 4)
```

Coercion often happens automatically. Most mathematical functions (`+`, `log`, `abs`, etc.) will coerce to a double or integer, and most logical operations (`&`, `|`, `any`, etc) will coerce to a logical. You will usually get a warning message if the coercion might lose information. If confusion is likely, explicitly coerce with `as.character()`, `as.double()`, `as.integer()`, or `as.logical()`. 

### Lists

Lists are different from atomic vectors because they can contain any other type of object, including lists. You construct them using `list()` instead of `c()`.

```{r}
x <- list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))
str(x)
```

Lists are sometimes called __recursive__ vectors, because a list can contain other lists. This makes them fundamentally different from atomic vectors.

```{r}
x <- list(list(list(list())))
str(x)
is.recursive(x)
```

`c()` will combine several lists into one. If given a combination of atomic vectors and lists, `c()` will coerce the vectors to list before combining them. Compare the results of `list()` and `c()`:

```{r}
x <- list(list(1, 2), c(3, 4))
y <- c(list(1, 2), c(3, 4))
str(x)
str(y)
```

The `typeof()` a list is `list`, you can test for a list with `is.list()` and coerce to a list with `as.list()`.

Lists are used to build up many of the more complicated data structures in R. For example, both data frames (described in [data frames](#data-frames)) and linear models objects (as produced by `lm()`) are lists:

```{r}
is.list(mtcars)

mod <- lm(mpg ~ wt, data = mtcars)
is.list(mod)
```

You can turn a list into an atomic vector using `unlist()`. If the list is of mixed type, it uses the same coercion rules as `c()`, 

### Exercises

1. What are the six types of atomic vectors? How does a list differ from an
   atomic vector?

1. What makes `is.vector()` and `is.numeric()` fundamentally different to
   `is.list()` and `is.character()`?

1. Test your knowledge of vector coercion rules by predicting the output of
   the following uses of `c()`:

    ```{r, eval=FALSE}
    c(1, FALSE)
    c("a", 1)
    c(list(1), "a")
    c(TRUE, 1L)
    ```

1. Why is `1 == "1"` true? Why is `-1 < FALSE` true? Why is `"one" < 2` false?

1. Why is the default (and shortest) `NA` a logical vector? What's special
   about logical vectors?

## Attributes

All objects can have arbitrary additional attributes. These can be thought of as a named list (with unique names). Attributes can be accessed individually with `attr()` or all at once (as a list) with `attributes()`.

```{r}
y <- 1:10
attr(y, "my_attribute") <- "This is a vector"
attr(y, "my_attribute")
str(attributes(y))
```

The `structure()` function returns a new object with modified attributes:

```{r}
structure(1:10, my_attribute = "This is a vector")
```

By default, most attributes are lost when modifying a vector:

```{r}
attributes(y[1])
attributes(sum(y))
```

The exceptions are the most common attributes:

* `names()`, character vector of element names.

* `class()`, used to implement the S3 object system, described [S3](#s3).
  
* `dim()`, used to turn vectors into high-dimensional structures.

You should get and set these attributes with their accessor functions: use `names(x)`, `class(x)` and `dim(x)`, not `attr(x, "names")`, `attr(x, "class")`, and `attr(x, "dim")`.

#### Names {#vector-names}

You can name a vector in three ways:

* During creation: `x <- c(a = 1, b = 2, c = 3)`.

* By modifying an existing vector: `x <- 1:3; names(x) <- c("a", "b", "c")`.

* By creating a modified vector: `x <- setNames(1:3, c("a", "b", "c"))`.

Names should be unique, because character subsetting (see [subsetting](#subsetting)), the most important reason to use names, will only return the first match.

Not all elements of a vector need to have a name. If any names are missing, `names()` will return an empty string for those elements. If all names are missing, `names()` will return NULL.

```{r}
y <- c(a = 1, 2, 3)
names(y)

z <- c(1, 2, 3)
names(z)
```

You can create a new vector without names using `unname(x)`, or remove names in place with `names(x) <- NULL`.

### Factors

A factor is a vector that can contain only predefined values. It is R's structure for dealing with qualitative data. Factors are built on top of integer vectors using an S3 class, as described in [S3](#s3). Factors have two key attributes: their `class()`, "factor", which controls their behaviour; and their `levels()`, the set of allowed values.

```{r}
x <- factor(c("a", "b", "b", "a"))
x
class(x)
levels(x)

# You can't use values that are not in the levels
x[2] <- "c"
x

# NB: you can't combine factors
c(factor("a"), factor("b"))
```

While factors look (and often behave) like character vectors, they are actually integers under the hood. You need to be careful when treating them like strings. Some string methods (like `gsub()` and `grepl()`) will coerce factors to strings, while others (like `nchar()`) will throw an error, and still others (like `c()`) will use the underlying integer IDs. For this reason, it's usually best to explicitly convert factors to strings when modifying their levels.

Factors are useful when you know the possible values a variable may take, even if you don't see all values in a given dataset. Using a factor instead of a character vector makes it obvious when some groups contain no observations:

```{r}
sex_char <- c("m", "m", "m")
sex_factor <- factor(sex_char, levels = c("m", "f"))

table(sex_char)
table(sex_factor)
```

Sometimes when a data frame is read directly from a file, a column you had thought would produce a numeric vector instead produces a factor, with the numbers appearing in the levels. This is caused by a non-numeric value in the column, often a missing value encoded in a special way like `.` or `-`. To remedy the situation you will need to coerce the vector from a factor to character, and then from character to numeric. (Be sure to check for missing values after this process.) Of course, a much better plan is to discover and fix what caused the problem in the first place; using the `na.strings` argument to `read.csv()` is often a good place to start.

```{r}
# Reading in "text" instead of from a file here:
z <- read.csv(text="value\n12\n1\n.\n9")
typeof(z$value)
as.numeric(z$value)
# Oops, that's not right: 3 2 1 4 are the levels of a factor, 
# not the values we read in!
class(z$value)
# We can fix it now:
as.numeric(as.character(z$value))
# Or change how we read it in:
z <- read.csv(text="value\n12\n1\n.\n9", na.strings=".")
typeof(z$value)
class(z$value)
z$value
# Perfect! :)
```

Unfortunately, most data loading functions in R automatically convert character vectors to factors. This is suboptimal, because there's no way for those functions to know the set of all possible levels or their optimal order. Instead, use the argument `stringsAsFactors = FALSE` to suppress this behaviour, and then manually convert character vectors to factors using your knowledge of the data. A global option (`options(stringsAsFactors = FALSE)`) is available to control this behaviour, but it's not recommended - it makes it harder to share your code, and it may have unexpected consequences when combined with other code (either from packages, or code that you're `source()`ing). Global options make code harder to understand, because they increase the number of lines you need to read to understand what a function is doing. 

In early versions of R, there was a memory advantage to using factors instead of character vectors, but this is no longer the case.

Atomic vectors and lists are the building blocks for higher dimensional data structures. Atomic vectors extend to matrices and arrays, and lists are used to create data frames.

### Exercises

1.  An early draft used this code to illustrate `structure()`:

    ```{r}
    structure(1:5, comment = "my attribute")
    ```

    But when you print that object you don't see the comment attribute.
    Why? Is the attribute missing, or is there something else special about
    it? (Hint: try using help)

## Matrices and arrays

Adding a `dim()` attribute allows an atomic vector to also be treated like a multi-dimensional __array__. A special case of the array is the __matrix__, which has two dimensions. Matrices are used commonly as part of the mathematical machinery of statistics. Arrays are much rarer, but worth being aware of.

Matrices and arrays are created with `matrix()` and `array()`, or by using the assignment form of `dim()`:

```{r}
# Two scalar arguments to specify rows and columns
a <- matrix(1:6, ncol = 3, nrow = 2)
# One vector argument to describe all dimensions
b <- array(1:12, c(2, 3, 2))

# You can also modify an object in place by setting dim()
c <- 1:6
dim(c) <- c(3, 2)
c
dim(c) <- c(2, 3)
c
```

The basic properties `length()` and `names()` have high-dimensional generalisations that work with matrices and arrays:

* `length()` generalises to `nrow()` and `ncol()` for matrices, and `dim()`
  for arrays.

* `names()` generalises to `rownames()` and `colnames()` for matrices, and
  `dimnames()`, a list, for arrays.

```{r}
length(a)
nrow(a)
ncol(a)
rownames(a) <- c("A", "B")
colnames(a) <- c("a", "b", "c")
a

length(b)
dim(b)
dimnames(b) <- list(c("one", "two"), c("a", "b", "c"), c("A", "B"))
b
```

`c()` generalises to `cbind()` and `rbind()` for matrices, and to `abind()` (provided by the `abind` package) for arrays. You can transpose a matrix with `t()`; the generalised equivalent for arrays is `aperm()`.

You can test if an object is a matrix or array using `is.matrix()` and `is.array()`, or by looking at the length of the `dim()`. `as.matrix()` and `as.array()` make it easy to turn an existing vector into a matrix or array.

Vectors are not the only 1-dimensional data structure. You can have matrices with a single row or single column, or arrays with a single dimension. They may print similarly, but will behave differently. The differences aren't too important, but it's useful to know they exist in case you get strange output from a function (`tapply()` is a frequent offender). As always, use `str()` to reveal the differences.

```{r}
str(1:3)                   # 1d vector
str(matrix(1:3, ncol = 1)) # column vector
str(matrix(1:3, nrow = 1)) # row vector
str(array(1:3, 3))         # "array" vector
```

While atomic vectors are most commonly turned into matrices, the dimension attribute can also be set on lists to make list-matrices or list-arrays:

```{r}
l <- list(1:3, "a", TRUE, 1.0)
dim(l) <- c(2, 2)
l
```

These are relatively esoteric data structures, but can be useful if you want to arrange objects into a grid-like structure. For example, if you're running models on a spatio-temporal grid, it might be natural to preserve the grid structure by storing the models in a 3d array.

### Exercises

1.  If `is.matrix(x)` is `TRUE`, what will `is.array(x)` return?

1.  How would you describe the following three objects? What makes them
    different to `1:5`?

    ```{r}
    x1 <- array(1:5, c(1, 1, 5))
    x2 <- array(1:5, c(1, 5, 1))
    x3 <- array(1:5, c(5, 1, 1))
    ```

## Data frames

A data frame is the most common way of storing data in R, and if [used systematically](http://vita.had.co.nz/papers/tidy-data.pdf) make data analysis easier. Under the hood, a data frame is a list of equal-length vectors. This makes it a 2-dimensional structure, so it shares properties of both the matrix and the list.  This means that a data frame has `names()`, `colnames()` and `rownames()`, although `names()` and `colnames()` are the same thing. The `length()` of a data frame is the length of the underlying list and so is the same as `ncol()`, `nrow()` gives the number of rows.

As described in [subsetting](#subsetting), you can subset a data frame like a 1d structure (where it behaves like a list), or a 2d structure (where it behaves like a matrix).

### Creation

You create a data frame using `data.frame()`, which takes named vectors as input:

```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
str(df)
```

Beware `data.frame()`'s default behaviour which turns strings into factors. Use `stringAsFactors = FALSE` to suppress this behaviour:

```{r}
df <- data.frame(
  x = 1:3,
  y = c("a", "b", "c"),
  stringsAsFactors = FALSE)
str(df)
```

### Testing and coercion

Because `data.frame` is an S3 class, its type reflects the underlying vector used to build it: `list`. To check if an object is a data frame, use `class()` or test explicitly with `is.data.frame()`:

```{r}
typeof(df)
class(df)
is.data.frame(df)
```

You can coerce an object to a data frame with `as.data.frame()`:

* A vector will create a one-column data frame.

* A list will create one column for each element; it's an error if they're 
  not all the same length.
  
* A matrix will create a data frame with the same number of columns and rows.

### Combining data frames

You can combine data frames using `cbind()` and `rbind()`:

```{r}
cbind(df, data.frame(z = 3:1))
rbind(df, data.frame(x = 10, y = "z"))
```

When combining column-wise, the number of rows must match, but rownames are ignored. When combining row-wise, both the number and names of columns must match. Use `plyr::rbind.fill()` to combine data frames that don't have the same columns. 

It's a common mistake to try and create a data frame by `cbind()`ing vectors together. This doesn't work because `cbind()` will create a matrix unless one of the arguments is already a data frame. Instead use `data.frame()` directly:

```{r}
bad <- data.frame(cbind(a = 1:2, b = c("a", "b")))
str(bad)
good <- data.frame(a = 1:2, b = c("a", "b"),
  stringsAsFactors = FALSE)
str(good)
```

The conversion rules for `cbind()` are complicated and best avoided by ensuring all inputs are of the same type.

### Special columns

Since a data frame is a list of vectors, it is possible for a data frame to have a column that is a list:

```{r}
df <- data.frame(x = 1:3)
df$y <- list(1:2, 1:3, 1:4)
df
```

However, when a list is given to `data.frame()`, it tries to put each item of the list into its own column, so this fails:

```{r, error = TRUE}
data.frame(x = 1:3, y = list(1:2, 1:3, 1:4))
```

A workaround is to use `I()`, which causes `data.frame()` to treat the list as one unit:

```{r}
dfl <- data.frame(x = 1:3, y = I(list(1:2, 1:3, 1:4)))
str(dfl)
dfl[2, "y"]
```

`I()` adds the `AsIs` class to its input, but this can usually be safely ignored.

Similarly, it's also possible to have a column of a data frame that's a matrix or array, as long as the number of rows matches the data frame:

```{r}
dfm <- data.frame(x = 1:3, y = I(matrix(1:9, nrow = 3)))
str(dfm)
dfm[2, "y"]
```

Use list and array columns with caution: many functions that work with data frames assume that all columns are atomic vectors.

### Exercises

1.  What attributes does a data frame possess?

1.  What does `as.matrix()` do when applied to a data frame with 
    columns of different types?

1.  Can you have a data frame with 0 rows? What about 0 columns?
