# Evaluation

```{r setup, include = FALSE}
source("common.R")
library(rlang)
```

## Introduction

The opposite of quotation is evaluation. 

## Overscopes

### What to use `...` for?

Take the `grouped_mean` function we defined above - there are 3 possible ways we could use `...`:

* Pass `...` onto the `mean()` function. That would make it easy to set 
  `na.rm = TRUE`. This is easiest to implement.
  
* Allow the user to select multiple groups

* Allow the user to select multiple variables to summarise.

```{r}
grouped_mean <- function(df, group_by, summarise, ...) {}
grouped_mean <- function(df, group_by, ...) {}
grouped_mean <- function(df, ..., summarise) {}
```

What if we want to use many of these?

```{r}
library(purrr)

grouped_mean <- function(df, groups, vars, args) {

  var_means <- map(vars, function(var) expr(mean(!!var, !!!args)))
  names(var_means) <- map_chr(vars, expr_name)
  
  df %>%
    dplyr::group_by(!!!groups) %>%
    dplyr::summarise(!!!var_means)
}

grouped_mean(mtcars, exprs(vs, am), exprs(hp, drat, wt), list(na.rm = TRUE))
```

This function needs more work in order to give useful error messages.

This is very similar to the scoped verbs in dplyr (e.g. `summarise_if()`, `mutate_at()`), and dplyr provides a wrapper to make the expression look a little more natural.

```{r}
library(dplyr)
grouped_mean <- function(df, groups, vars, ...) {

  var_means <- map(vars, function(var) expr(mean(!!var, ...)))
  names(var_means) <- map_chr(vars, quo_name)
  
  df %>%
    dplyr::group_by(!!!groups) %>%
    dplyr::summarise(!!!var_means)
}

grouped_mean(mtcars, vars(vs, am), vars(hp, drat, wt), na.rm = TRUE)
```



## Base R

### Formulas

```{r}
make_model <- function(resp, preds) {
  pred_sum <- purrr::reduce(preds, ~ expr(!!.x + !!.y))
  eval_tidy(expr(!!resp ~ !!pred_sum))
}
make_model(expr(y), exprs(a, b, c))
```

### `match.call()`

Why is it dangerous?

## Case studies

### Wrapping NSE

Standard NSE for modelling functions. How can you capture the `subset` or `weight` argument in a function? Need to use `expr()`. 

```{r, error = TRUE, fig.keep = "none"}
library(lattice)
xyplot(mpg ~ disp, data = mtcars)

x <- quote(mpg)
y <- quote(disp)
xyplot(x ~ y, data = mtcars)
```


### Promises

These functions work because internally R represents function arguments with a special type of object called a __promise__. A promise captures the expression needed to compute the value and the environment in which to compute it. You're not normally aware of promises because the first time you access a promise its code is evaluated in its environment, yielding a value. \index{promises}

Promises are hard to work with because they are quantum - attempting to look at them in R changes their behaviour.

You can see both the expression and it's environment if you capture with a quosure, rather than an expression.

```{r}
capture <- function(x) {
  enquo(x)
}

f <- function(x) {
  x <- 10
  capture(x + 1)
}

x <- 1
f1 <- capture(x + 1)
f2 <- f(x + 1)

get_expr(f1)
get_expr(f2)

get_env(f1)
get_env(f2)

eval_tidy(f1)
eval_tidy(f2)
```

We'll come back to quosures in the next chapter: because they capture the environment they're most useful for non-standard evaluation.


## Non-standard evaluation in subset {#subset}

While printing out the code supplied to an argument value can be useful, we can actually do more with the unevaluated code. Take `subset()`, for example. It's a useful interactive shortcut for subsetting data frames: instead of repeating the name of data frame many times, you can save some typing: \indexc{subset()}

```{r}
sample_df <- data.frame(a = 1:5, b = 5:1, c = c(5, 3, 1, 4, 1))

subset(sample_df, a >= 4)
# equivalent to:
# sample_df[sample_df$a >= 4, ]

subset(sample_df, b == c)
# equivalent to:
# sample_df[sample_df$b == sample_df$c, ]
```

`subset()` is special because it implements different scoping rules: the expressions `a >= 4` and `b == c` are evaluated in the specified data frame rather than in the current or global environments. This is the essence of non-standard evaluation.

How does `subset()` work? We've already seen how to capture an argument's expression rather than its result, so we just need to figure out how to evaluate that expression in the right context. Specifically, we want `x` to be interpreted as `sample_df$x`, not `globalenv()$x`. To do this, we need `eval()`. This function takes an expression and evaluates it in the specified environment. \indexc{eval()}

Before we can explore `eval()`, we need one more useful function: `quote()`. It captures an unevaluated expression like `substitute()`, but doesn't do any of the advanced transformations that can make `substitute()` confusing. `quote()` always returns its input as is: \indexc{quote()} \index{quoting}

```{r}
quote(1:10)
quote(x)
quote(x + y^2)
```

We need `quote()` to experiment with `eval()` because `eval()`'s first argument is an expression. So if you only provide one argument, it will evaluate the expression in the current environment. This makes `eval(quote(x))` exactly equivalent to `x`, regardless of what `x` is:

```{r, error = TRUE}
eval(quote(x <- 1))
eval(quote(x))

eval(quote(y))
```

`quote()` and `eval()` are opposites. In the example below, each `eval()` peels off one layer of `quote()`'s.

```{r}
quote(2 + 2)
eval(quote(2 + 2))

quote(quote(2 + 2))
eval(quote(quote(2 + 2)))
eval(eval(quote(quote(2 + 2))))
```

`eval()`'s second argument specifies the environment in which the code is executed:

```{r}
x <- 10
eval(quote(x))

e <- new.env()
e$x <- 20
eval(quote(x), e)
```

Because lists and data frames bind names to values in a similar way to environments, `eval()`'s second argument need not be limited to an environment: it can also be a list or a data frame. 

```{r}
eval(quote(x), list(x = 30))
eval(quote(x), data.frame(x = 40))
```

This gives us one part of `subset()`:

```{r}
eval(quote(a >= 4), sample_df)
eval(quote(b == c), sample_df)
```

A common mistake when using `eval()` is to forget to quote the first argument. Compare the results below:

```{r, error = TRUE}
a <- 10
eval(quote(a), sample_df)
eval(a, sample_df)

eval(quote(b), sample_df)
eval(b, sample_df)
```
```{r, echo = FALSE}
rm(a)
```

We can use `eval()` and `substitute()` to write `subset()`. We first capture the call representing the condition, then we evaluate it in the context of the data frame and, finally, we use the result for subsetting:

```{r}
subset2 <- function(x, condition) {
  condition_call <- substitute(condition)
  r <- eval(condition_call, x)
  x[r, ]
}
subset2(sample_df, a >= 4)
```

### Exercises

1.  Predict the results of the following lines of code:

    ```{r, eval = FALSE}
    eval(quote(eval(quote(eval(quote(2 + 2))))))
    eval(eval(quote(eval(quote(eval(quote(2 + 2)))))))
    quote(eval(quote(eval(quote(eval(quote(2 + 2)))))))
    ```

1.  `subset2()` has a bug if you use it with a single column data frame.
    What should the following code return? How can you modify `subset2()`
    so it returns the correct type of object?

    ```{r}
    sample_df2 <- data.frame(x = 1:10)
    subset2(sample_df2, x > 8)
    ```

1.  The real subset function (`subset.data.frame()`) removes missing
    values in the condition. Modify `subset2()` to do the same: drop the 
    offending rows.

1.  What happens if you use `quote()` instead of `substitute()` inside of
    `subset2()`?

1.  The third argument in `subset()` allows you to select variables. It
    treats variable names as if they were positions. This allows you to do 
    things like `subset(mtcars, , -cyl)` to drop the cylinder variable, or
    `subset(mtcars, , disp:drat)` to select all the variables between `disp`
    and `drat`. How does this work? I've made this easier to understand by
    extracting it out into its own function.

    ```{r, eval = FALSE}
    select <- function(df, vars) {
      vars <- substitute(vars)
      var_pos <- setNames(as.list(seq_along(df)), names(df))
      pos <- eval(vars, var_pos)
      df[, pos, drop = FALSE]
    }
    select(mtcars, -cyl)
    ```

1.  What does `evalq()` do? Use it to reduce the amount of typing for the
    examples above that use both `eval()` and `quote()`.

1.  Write an equivalent to `get()` using `as.name()` and `eval()`. Write an
    equivalent to `assign()` using `as.name()`, `substitute()`, and `eval()`.
    (Don't worry about the multiple ways of choosing an environment; assume
    that the user supplies it explicitly.)

## Scoping issues {#scoping-issues}

It certainly looks like our `subset2()` function works. But since we're working with expressions instead of values, we need to test things more extensively. For example, the following applications of `subset2()` should all return the same value because the only difference between them is the name of a variable: \index{lexical scoping}

```{r, error = TRUE}
y <- 4
x <- 4
condition <- 4
condition_call <- 4

subset2(sample_df, a == 4)
subset2(sample_df, a == y)
subset2(sample_df, a == x)
subset2(sample_df, a == condition)
subset2(sample_df, a == condition_call)
```

What went wrong? You can get a hint from the variable names I've chosen: they are all names of variables defined inside `subset2()`. If `eval()` can't find the variable inside the data frame (its second argument), it looks in the environment of `subset2()`. That's obviously not what we want, so we need some way to tell `eval()` where to look if it can't find the variables in the data frame.

The key is the third argument to `eval()`: `enclos`. This allows us to specify a parent (or enclosing) environment for objects that don't have one (like lists and data frames). If the binding is not found in `env`, `eval()` will next look in `enclos`, and then in the parents of `enclos`. `enclos` is ignored if `env` is a real environment. We want to look for `x` in the environment from which `subset2()` was called. In R terminology this is called the __parent frame__ and is accessed with `parent.frame()`. This is an example of [dynamic scope](http://en.wikipedia.org/wiki/Scope_%28programming%29#Dynamic_scoping): the values come from the location where the function was called, not where it was defined. \indexc{parent.frame()}

With this modification our function now works:

```{r}
subset2 <- function(x, condition) {
  condition_call <- substitute(condition)
  r <- eval(condition_call, x, parent.frame())
  x[r, ]
}

x <- 4
subset2(sample_df, a == x)
```

Using `enclos` is just a shortcut for converting a list or data frame to an environment. We can get the same behaviour by using `list2env()`. It turns a list into an environment with an explicit parent: \indexc{list2env()}

```{r}
subset2a <- function(x, condition) {
  condition_call <- substitute(condition)
  env <- list2env(x, parent = parent.frame())
  r <- eval(condition_call, env)
  x[r, ]
}

x <- 5
subset2a(sample_df, a == x)
```

### Exercises

1.  What does `transform()` do? Read the documentation. How does it work?
    Read the source code for `transform.data.frame()`. What does
    `substitute(list(...))` do?

1.  What does `with()` do? How does it work? Read the source code for
    `with.default()`. What does `within()` do? How does it work? Read the
    source code for `within.data.frame()`. Why is the code so much more
    complex than `with()`?



## Capturing the current call {#capturing-call}

```{r, eval = FALSE, echo = FALSE}
std <- c("package:base", "package:utils", "package:stats")
names(find_uses(std, "sys.call"))
names(find_uses(std, "match.call"))
```

Many base R functions use the current call: the expression that caused the current function to be run. There are two ways to capture a current call: \indexc{calls|capturing current}

* `sys.call()` captures exactly what the user typed. \indexc{sys.call()}

* `match.call()` makes a call that only uses named arguments. It's like
  automatically calling `pryr::standardise_call()` on the result of
  `sys.call()` \indexc{match.call()}

The following example illustrates the difference between the two:

```{r}
f <- function(abc = 1, def = 2, ghi = 3) {
  list(sys = sys.call(), match = match.call())
}
f(d = 2, 2)
```

Modelling functions often use `match.call()` to capture the call used to create the model. This makes it possible to `update()` a model, re-fitting the model after modifying some of original arguments. Here's an example of `update()` in action: \indexc{update()}

```{r}
mod <- lm(mpg ~ wt, data = mtcars)
update(mod, formula = . ~ . + cyl)
```

How does `update()` work? We can rewrite it using some tools from pryr to focus on the essence of the algorithm.

```{r, eval = FALSE}
update_call <- function (object, formula., ...) {
  call <- object$call

  # Use update.formula to deal with formulas like . ~ .
  if (!missing(formula.)) {
    call$formula <- update.formula(formula(object), formula.)
  }

  modify_call(call, dots(...))
}
update_model <- function(object, formula., ...) {
  call <- update_call(object, formula., ...)
  eval(call, parent.frame())
}
update_model(mod, formula = . ~ . + cyl)
```

The original `update()` has an `evaluate` argument that controls whether the function returns the call or the result. But I think it's better, on principle, that a function returns only one type of object, rather than different types depending on the function's arguments.

This rewrite also allows us to fix a small bug in `update()`: it re-evaluates the call in the global environment, when what we really want is to re-evaluate it in the environment where the model was originally fit --- in the formula.

```{r, error = TRUE}
f <- function() {
  n <- 3
  lm(mpg ~ poly(wt, n), data = mtcars)
}
mod <- f()
update(mod, data = mtcars)

update_model <- function(object, formula., ...) {
  call <- update_call(object, formula., ...)
  eval(call, environment(formula(object)))
}
update_model(mod, data = mtcars)
```

This is an important principle to remember: if you want to re-run code captured with `match.call()`, you also need to capture the environment in which it was evaluated, usually the `parent.frame()`. The downside to this is that capturing the environment also means capturing any large objects which happen to be in that environment, which prevents their memory from being released. This topic is explored in more detail in [garbage collection](#gc). \index{environments|capturing}

Some base R functions use `match.call()` where it's not necessary. For example,  `write.csv()` captures the call to `write.csv()` and mangles it to call `write.table()` instead:

```{r}
write.csv <- function(...) {
  Call <- match.call(expand.dots = TRUE)
  for (arg in c("append", "col.names", "sep", "dec", "qmethod")) {
    if (!is.null(Call[[arg]])) {
      warning(gettextf("attempt to set '%s' ignored", arg))
    }
  }
  rn <- eval.parent(Call$row.names)
  Call$append <- NULL
  Call$col.names <- if (is.logical(rn) && !rn) TRUE else NA
  Call$sep <- ","
  Call$dec <- "."
  Call$qmethod <- "double"
  Call[[1L]] <- as.name("write.table")
  eval.parent(Call)
}
```

To fix this, we could implement `write.csv()` using regular function call semantics:

```{r}
write.csv <- function(x, file = "", sep = ",", qmethod = "double", 
                      ...) {
  write.table(x = x, file = file, sep = sep, qmethod = qmethod, 
    ...)
}
```

This is much easier to understand: it's just calling `write.table()` with different defaults. This also fixes a subtle bug in the original `write.csv()`: `write.csv(mtcars, row = FALSE)` raises an error, but `write.csv(mtcars, row.names = FALSE)` does not. The lesson here is that it's always better to solve a problem with the simplest tool possible.

### Exercises

1.  Compare and contrast `update_model()` with `update.default()`.

1.  Why doesn't `write.csv(mtcars, "mtcars.csv", row = FALSE)` work?
    What property of argument matching has the original author forgotten?

1.  Rewrite `update.formula()` to use R code instead of C code.

1.  Sometimes it's necessary to uncover the function that called the
    function that called the current function (i.e., the grandparent, not
    the parent). How can you use `sys.call()` or `match.call()` to find
    this function?


## Anaphoric functions

One useful application of `make_function()` is in functions like `curve()`. `curve()` allows you to plot a mathematical function without creating an explicit R function:

```{r curve-demo, fig.width = 3.5, fig.height = 2.5, small_mar = TRUE}
curve(sin(exp(4 * x)), n = 1000)
```

Here `x` is a pronoun. `x` doesn't represent a single concrete value, but is instead a placeholder that varies over the range of the plot. One way to implement `curve()` would be with `make_function()`:

```{r curve2}
curve4 <- function(expr, xlim = c(0, 1), n = 100) {
  expr <- enquo(expr)
  f <- new_function(alist(x = ), get_expr(expr), get_env(env))

  x <- seq(xlim[1], xlim[2], length = n)
  y <- f(x)

  plot(x, y, type = "l", ylab = deparse(substitute(expr)))
}
curve4(sin(exp(4 * x)), n = 1000)


curve3 <- function(expr, xlim = c(0, 1), n = 100) {
  expr <- enquo(expr)
  e <- rlang::expr({
    function(x) !!get_expr(expr)
  })
  f <- eval_tidy(e, env = get_env(expr))

  x <- seq(xlim[1], xlim[2], length = n)
  y <- f(x)

  plot(x, y, type = "l", ylab = deparse(substitute(expr)))
}

curve3(sin(exp(4 * x)), n = 1000)
```

Functions that use a pronoun are called [anaphoric](http://en.wikipedia.org/wiki/Anaphora_(linguistics)) functions. They are used in [Arc](http://www.arcfn.com/doc/anaphoric.html) (a lisp like language), [Perl](http://www.perlmonks.org/index.pl?node_id=666047), and [Clojure](http://amalloy.hubpages.com/hub/Unhygenic-anaphoric-Clojure-macros-for-fun-and-profit). \index{anaphoric functions} \index{functions!anaphoric}

### Exercises

1.  How are `alist(a)` and `alist(a = )` different? Think about both the
    input and the output.

1.  Read the documentation and source code for `pryr::partial()`. What does it
    do? How does it work? Read the documentation and source code for
    `pryr::unenclose()`. What does it do and how does it work?

1.  The actual implementation of `curve()` looks more like

    ```{r curve3}
    curve3 <- function(expr, xlim = c(0, 1), n = 100,
                       env = parent.frame()) {
      env2 <- new.env(parent = env)
      env2$x <- seq(xlim[1], xlim[2], length = n)

      y <- eval(substitute(expr), env2)
      plot(env2$x, y, type = "l", 
        ylab = deparse(substitute(expr)))
    }
    ```

    How does this approach differ from `curve2()` defined above?

## Sourcee

With `parse()` and `eval()`, it's possible to write a simple version of `source()`. We read in the file from disk, `parse()` it and then `eval()` each component in a specified environment. This version defaults to a new environment, so it doesn't affect existing objects. `source()` invisibly returns the result of the last expression in the file, so `simple_source()` does the same. \index{source()}

```{r}
simple_source <- function(file, envir = new.env()) {
  stopifnot(file.exists(file))
  stopifnot(is.environment(envir))

  lines <- readLines(file, warn = FALSE)
  exprs <- parse(text = lines)

  n <- length(exprs)
  if (n == 0L) return(invisible())

  for (i in seq_len(n - 1)) {
    eval(exprs[i], envir)
  }
  invisible(eval(exprs[n], envir))
}
```

The real `source()` is considerably more complicated because it can `echo` input and output, and also has many additional settings to control behaviour.

### Exercises


1.  Compare and contrast `source()` and `sys.source()`.

1.  Modify `simple_source()` so it returns the result of _every_ expression,
    not just the last one.

1.  The code generated by `simple_source()` lacks source references. Read
    the source code for `sys.source()` and the help for `srcfilecopy()`,
    then modify `simple_source()` to preserve source references. You can
    test your code by sourcing a function that contains a comment. If
    successful, when you look at the function, you'll see the comment and
    not just the source code.
