# Functions

```{r setup, include = FALSE}
source("common.R")
```

## Introduction
\index{functions}

Functions are a fundamental building block of R: to master many of the more advanced techniques in this book, you need a solid foundation in how functions work. You've probably already created many R functions, and you're know how to use them to reduce duplication in your code. The focus of this chapter is to turn your existing, informal knowledge of functions into a rigorous understanding of what functions are and how they work. You'll see some interesting tricks and techniques, but most of what you'll learn will be more important as the building blocks for more advanced techniques. 

The most important thing to understand about R is that functions are objects in their own right, a language property often called "first-class functions". You can work with them exactly the same way you work with any other type of object. This theme will be explored in depth in [functional programming](#functional-programming).

### Quiz {-}

Answer the following questions to see if you can safely skip this chapter. You can find the answers at the end of the chapter in [answers](#function-answers).

1.  What are the three components of a function?

1.  What does the following code return?

    ```{r, eval = FALSE}
    x <- 10
    f1 <- function(x) {
      function() {
        x + 10
      }
    }
    f1(1)()
    ```
    
1.  How would you more typically write this code?

    ```{r, eval = FALSE}
    `+`(1, `*`(2, 3))
    ```
    
1.  How could you make this call easier to read?

    ```{r, eval = FALSE}
    mean(, TRUE, x = c(1:10, NA))
    ```

1.  Does the following function throw an error when called? Why/why not?

    ```{r, eval = FALSE}
    f2 <- function(a, b) {
      a * 10
    }
    f2(10, stop("This is an error!"))
    ```

1.  What is an infix function? How do you write it? What's a replacement 
    function? How do you write it?

1.  What function do you use to ensure that a cleanup action occurs 
    regardless of how a function terminates?

### Outline {-}

* [Function components](#function-components) describes the three main 
  components of a function.

* [Lexical scoping](#lexical-scoping) teaches you how R finds values from 
  names, the process of lexical scoping.

* [Every operation is a function call](#all-calls) shows you that everything
  that happens in R is a result of a function call, even if it doesn't look 
  like it.

* [Function arguments](#function-arguments) discusses the three ways of 
  supplying arguments to a function, how to call a function given a list of 
  arguments, and the impact of lazy evaluation.

* [Special calls](#special-calls) describes two special types of function: 
  infix and replacement functions.
  
* [Return values](#return-values) discusses how and when functions return
  values, and how you can ensure that a function does something before it
  exits.

### Prerequisites {-}

The only package you'll need is lobstr, which as in previous chapters, we'll use to dig in to the details behind the scenes.

## Function components {#function-components}

All R functions have three parts: \index{functions!body} \index{functions!formals} \index{functions!environment}

* the `formals()`, the list of arguments which controls how you can call the function.

* the `body()`, the code inside the function.

* the `environment()`, the "map" of the location of the function's variables.

When you print a function in R, it shows you these three important components. If the environment isn't displayed, it means that the function was created in the global environment. \indexc{formals()} \indexc{body()} \index{environments!of a function}

```{r, eval = FALSE}
f <- function(x) x^2
f
#> function(x) x^2

formals(f)
#> $x
body(f)
#> x^2
environment(f)
#> <environment: R_GlobalEnv>
```

The assignment forms of `formals()`, `body()`, and `environment()` can also be used to modify functions.

Like all objects in R, functions can also possess any number of additional `attributes()`. One attribute used by base R is "srcref", short for source reference, which points to the source code used to create the function. Unlike `body()`, this contains code comments and other formatting. You can also add attributes to a function. For example, you can set the `class()` and add a custom `print()` method. \index{functions!attributes}

### Primitive functions

There is one exception to the rule that functions have three components. Primitive functions, like `sum()`, call C code directly with `.Primitive()` and contain no R code. Therefore their `formals()`, `body()`, and `environment()` are all `NULL`:  \index{primitive functions} \index{functions!primitive|see{primitive functions}} \index{.Primitive@\texttt{.Primitive()}|see{Primitive functions}}

```{r}
sum
formals(sum)
body(sum)
environment(sum)
```

Primitive functions are only found in the `base` package and since they operate at a low level, they can be more efficient (primitive replacement functions don't have to make copies) and have different rules for argument matching (e.g. `switch` and `call`).  This, however, comes at the cost of behaving differently from all other functions in R. Hence the R core team generally avoids creating them unless there is no other option.

### Exercises

1.  What function allows you to tell if an object is a function? What function
    allows you to tell if a function is a primitive function?

1.  This code makes a list of all functions in the base package. 
    
    ```{r}
    objs <- mget(ls("package:base"), inherits = TRUE)
    funs <- Filter(is.function, objs)
    ```

    Use it to answer the following questions:

    a. Which base function has the most arguments?
    
    a. How many base functions have no arguments? What's special about those
       functions?
       
    a. How could you adapt the code to find all primitive functions?

1. What are the three important components of a function?

1. When does printing a function not show what environment it was created in?

## Lexical scoping {#lexical-scoping}
 \index{scoping!lexical|see{lexical scoping}} \index{lexical scoping}
 
In [Names and values], we discussed assignment, the act of binding a name to a value. __Scoping__ is the opposite: the scoping rules determine how R finds the value bound to a name. The basic rules of scoping are quite intuitive, and you've probably already internalised them, even if you've never been explicitly taught them. For example, what will the following code return, 10 or 20?

```{r, results = "hide"}
x <- 10
f <- function() {
  x <- 20
  x
}

f()
```

In this section you'll learn the rules of scoping more formally, and learn some of the subtler details you might not have yet encounted. A deeper understanding of scoping will help you use advanced functional programming tools, and eventually, to write tools that translate R code to other languages.

R primarily uses __lexical scoping__[^dyn-scope], which looks up the values of names based on how functions were defined (not how they are called). The "lexical" in lexical scoping doesn't correspond to the usual English definition ("of or relating to words or the vocabulary of a language as distinguished from its grammar and construction") but comes from the computer science term "lexing", which is part of the process that converts code represented as text to meaningful pieces that the programming language understands.

[^dyn-scope]: Functions that automatically quote one or more arguments (sometimes called NSE functions) can override the default scoping rules to implement other varieties of scoping. You'll learn more about that in [metaprogramming].

There are four primarily rules that govern scoping in R:

* Name masking
* Functions vs. variables
* A fresh start
* Dynamic lookup

### Name masking

The following example illustrates the most basic principle of lexical scoping:

```{r, eval = FALSE}
f <- function() {
  x <- 1
  y <- 2
  c(x, y)
}
f()
rm(f)
```

If a name isn't defined inside a function, R will look one level up.

```{r, eval = FALSE}
x <- 2
g <- function() {
  y <- 1
  c(x, y)
}
g()
rm(x, g)
```

The same rules apply if a function is defined inside another function: look inside the current function, then where that function was defined, and so on, all the way up to the global environment, and then on to other loaded packages. Run the following code in your head, then confirm the output by running the R code.

```{r, results = "hide"}
x <- 1
h <- function() {
  y <- 2
  i <- function() {
    z <- 3
    c(x, y, z)
  }
  i()
}
h()
rm(x, h)
```

The same rules apply to closures, functions created by other functions. Closures will be described in more detail in [functional programming](#functional-programming); here we'll just look at how they interact with scoping. The following function, `j()`, returns a function.  What do you think this function will return when we call it? \index{closures!scoping}

```{r, eval = FALSE}
j <- function(x) {
  y <- 2
  function() {
    c(x, y)
  }
}
k <- j(1)
k()
rm(j, k)
```

This seems a little magical (how does R know what the value of `y` is after the function has been called). It works because `k` preserves the environment in which it was defined and because the environment includes the value of `y`. [Environments](#environments) gives some pointers on how you can dive in and figure out what values are stored in the environment associated with each function.

### Functions vs. variables

The same principles apply regardless of the type of associated value --- finding functions works exactly the same way as finding variables:

```{r}
l <- function(x) x + 1
m <- function() {
  l <- function(x) x * 2
  l(10)
}
m()
rm(l, m)
```

For functions, however, there is one small tweak to the rule. If you are using a name in a context where it's obvious that you want a function (e.g., `f(3)`), R will ignore objects that are not functions while it is searching. In the following example `n` takes on a different value depending on whether R is looking for a function or a variable.

```{r}
n <- function(x) x / 2
o <- function() {
  n <- 10
  n(n)
}
o()
rm(n, o)
```

However, using the same name for functions and other objects will make for confusing code, and is best avoided.

### A fresh start {#fresh-start}

What happens to the values in between invocations of a function? What will happen the first time you run this function? What will happen the second time? (If you haven't seen `exists()` before: it returns `TRUE` if there's a variable of that name, otherwise it returns `FALSE`.)

```{r, eval = FALSE}
j <- function() {
  if (!exists("a")) {
    a <- 1
  } else {
    a <- a + 1
  }
  a
}
j()
rm(j)
```

You might be surprised that it returns the same value, `1`, every time. This is because every time a function is called, a new environment is created to host execution. A function has no way to tell what happened the last time it was run; each invocation is completely independent. (We'll see some ways to get around this in [mutable state](#mutable-state).)

### Dynamic lookup

Lexical scoping determines where to look for values, not when to look for them. R looks for values when the function is run, not when it's created. This means that the output of a function can be different depending on objects outside its environment: 

```{r}
f <- function() x
x <- 15
f()

x <- 20
f()
```

You generally want to avoid this behaviour because it means the function is no longer self-contained. This is a common error --- if you make a spelling mistake in your code, you won't get an error when you create the function, and you might not even get one when you run the function, depending on what variables are defined in the global environment.

One way to detect this problem is the `findGlobals()` function from `codetools`. This function lists all the external dependencies of a function: \indexc{findGlobals()}

```{r}
f <- function() x + 1
codetools::findGlobals(f)
```

Another way to try and solve the problem would be to manually change the environment of the function to the `emptyenv()`, an environment which contains absolutely nothing:

```{r, error = TRUE}
environment(f) <- emptyenv()
f()
```

This doesn't work because R relies on lexical scoping to find _everything_, even the `+` operator. It's very difficult to make a function completely self-contained because you must always rely on functions defined in base R or other packages.

### Exercises

1. What does the following code return? Why? What does each of the three `c`'s mean?

    ```{r, eval = FALSE}
    c <- 10
    c(c = c)
    ```

2. What are the four principles that govern how R looks for values?

3. What does the following function return? Make a prediction before 
   running the code yourself.

    ```{r, results = "hide"}
    f <- function(x) {
      f <- function(x) {
        f <- function(x) {
          x ^ 2
        }
        f(x) + 1
      }
      f(x) * 2
    }
    f(10)
    ```

## Function forms {#all-calls}

> "To understand computations in R, two slogans are helpful:
>
> * Everything that exists is an object.
> * Everything that happens is a function call."
>
> --- John Chambers

While everything that happens in R is a result of a function call, not all calls look the same. There are three common forms:

* In __prefix__ form where the name of the function comes before its
  arguments, like `foofy(a, b, c)`. These constitute of the majority of
  function calls in R.

* In __infix__ form, the function come in between its arguments, like `x + y`. 
  Infix forms are used for many mathematical operators, and user-defined 
  functions that begin and end with `%`.
  
* A __replacement__ function looks like a prefix functions, but includes
  a `<-`, like `names(df) <- c("a", "b", "c")`.
  
There are also a variety of __special forms__ like `[[`, `if`, and `for`. These are built-in to R; you can only add new forms by modifying R's parser, which can only be done by R-core (and requires careful analysis to prevent introducing parsing ambiguities).

### Rewriting to prefix form
\indexc{`} \index{backticks|see{\texttt{`}}}

An interesting propety of R is every infix, replacement, or special form has an underlying function that can be called in prefix form. This means that you can rewrite any R expression to use only prefix form. The names of these underlying forms always have non-syntactic names, you to convert to prefix form you must wrap in in backticks, `` ` ``. The following example shows three pairs of equivalent calls, for an infix form, replacement form, and a special form.

```{r, eval = FALSE}
x + y
`+`(x, y)

names(df) <- c("x", "y", "z")
`names<-`(df, c("x", "y", "z"))

for(i in 1:10) print(i)
`for`(i, 1:10, print(i))
```

This transformation is performed automatically behind the scenes, then R uses the usual rules to find the function given its name. This means that you can override the behaviour of any infix function or special  form. For example, if you're ever feeling particularly evil, run the following code while your friend is away from their computer. It will introduce a particularly pernicious bug: 10% of the time, 1 will be added to any numeric calculation inside of parentheses.

```{r}
`(` <- function(e1) {
  if (is.numeric(e1) && runif(1) < 0.1) {
    e1 + 1
  } else {
    e1
  }
}
replicate(50, (1 + 2))
rm("(")
```

Of course, overriding built-in functions like this is a bad idea, but it is possible to apply it only selected code blocks. This provides a clean and elegant approach to writing domain specific languages and translators to other languages, as you'll learn about in [metaprogramming].

A technique more often useful in practice is this allows us to use more functions with functional programming tools. For example, could use `sapply()` to add 3 to every element of a list by first defining a function `add()`, like this: \indexc{sapply()}

```{r}
add <- function(x, y) x + y
sapply(1:10, add, 3)
```

But we can also get the same effect using the built-in `+` function.

```{r}
sapply(1:5, `+`, 3)
```

We'll explore this idea in detail later on in [functionals].

### Prefix form

The prefix form is the most common form in R code, and indeed in the majority of programming languages. R is a little special however, in that you can specific arguments by position, complete name, or partial name. Arguments are matched first by exact name, then partialy (by unique prefix), and finally by position. \index{functions!arguments}

```{r, error = TRUE}
f <- function(abcdef, bcde1, bcde2) {
  list(a = abcdef, b1 = bcde1, b2 = bcde2)
}
str(f(1, 2, 3))
str(f(2, 3, abcdef = 1))

# Can abbreviate long argument names:
str(f(2, 3, a = 1))

# But this doesn't work because abbreviation is ambiguous
str(f(1, 3, b = 1))
```

Generally, you only want to use positional matching for the first one or two arguments; they will be the most commonly used, and most readers will know what they are. Avoid using positional matching for less commonly used arguments, and never use partial matching. Named arguments should always come after unnamed arguments. The tidyverse style guide, <http://style.tidyverse.org/syntax.html#argument-names>, has more advice on good function calling style.

### Infix functions
\index{functions!infix} 
\index{infix functions} 
\indexc{\%\%}
 
R comes with a number of built-in infix operators: `:, ::, :::, $, @, ^, *, /, +, -, >, >=, <, <=, ==, !=, !, &, &&, |, ||, ~, <-, <<-`. You can also create your own infix functions must start and end with `%`, and base R uses this to additionally define `%%`, `%*%`, `%/%`, `%in%`, `%o%`, and `%x%` functions.

Defining your own infix function is simple. You create a two argument function and bind it to a name that starts and ends with `%`:

```{r}
`%+%` <- function(a, b) paste0(a, b)
"new " %+% "string"
```

Note that while prefix functions can have any number of arguments, infix functions must have exactly two. 

The names of infix functions are more flexible than regular R functions: they can contain any sequence of characters, except "%". You will need to escape any special characters in the string used to define the function, but not when you call it:

```{r}
`% %` <- function(a, b) paste(a, b)
`%'%` <- function(a, b) paste(a, b)
`%/\\%` <- function(a, b) paste(a, b)

"a" % % "b"
"a" %'% "b"
"a" %/\% "b"
```

R's default precedence rules mean that infix operators are composed from left to right:

```{r}
`%-%` <- function(a, b) paste0("(", a, " %-% ", b, ")")
"a" %-% "b" %-% "c"
```

Note that there are two functions that are usually infix, but can be called with a single argument: `+` and `-`. 

```{r}
-1
+10
```

### Replacement functions {#replacement-functions}
\index{replacement functions} 
\index{functions!replacement}
\index{assignment!replacement functions}

Replacement functions act like they modify their arguments in place, and have the special name `xxx<-`. They typically have two arguments (`x` and `value`) and they must return the modified object. For example, the following function allows you to modify the second element of a vector: 

```{r}
`second<-` <- function(x, value) {
  x[2] <- value
  x
}
```

You can of course call this using the prefix form, but it's not very useful

```{r}
x <- 1:4
`second<-`(x, 100L)
x
```

Instead, replacement functions are used by placing the function call on the LHS of `<-`: 

```{r}
x <- 1:10
second(x) <- 5L
x
```

I say they "act" like they modify their arguments in place, because, as discussed in [Modify-in-place], they actually create a modified copy. We can see that by using `tracemem()`:

```{r, eval = FALSE}
x <- 1:10
tracemem(x)
#> <0x7ffae71bd880>

second(x) <- 6L
#> tracemem[0x7ffae71bd880 -> 0x7ffae61b5480]: 
#> tracemem[0x7ffae61b5480 -> 0x7ffae73f0408]: second<- 
```

If you want to supply additional arguments, they go in between `x` and `value`:

```{r}
`modify<-` <- function(x, position, value) {
  x[position] <- value
  x
}
modify(x, 1) <- 10
x
```

When you call `modify(x, 1) <- 10`, behind the scenes R turns it into:

```{r, eval = FALSE}
x <- `modify<-`(x, 1, 10)
```

This means you can't do things like:

```{r, eval = FALSE}
modify(get("x"), 1) <- 10
#> Error: target of assignment expands to non-language object
```

because that gets turned into the invalid code:

```{r, eval = FALSE}
get("x") <- `modify<-`(get("x"), 1, 10)
#> Error: target of assignment expands to non-language object
```

It’s often useful to combine replacement and subsetting:

```{r}
x <- c(a = 1, b = 2, c = 3)
names(x)

names(x)[2] <- "two"
names(x)
```

This works because the expression `names(x)[2] <- "two"` is evaluated as if you had written:

```{r}
`*tmp*` <- x
x <- `names<-`(`*tmp*`, `[<-`(names(`*tmp*`), 2, "two"))
rm(`*tmp*`)
```

(Yes, it really does create a local variable named *tmp*, which is removed afterwards.)

### Special forms

Finally, there are a bunch of language features that are usually written in special ways, but also have prefix forms. These include parentheses:

* `(x)` (`` `(`(x) ``)
* `{y}` (`` `{`(y) ``).

The subsetting operators:

* `x[i]` (`` `[`(x, i) ``) 
* `x[[i]]` (`` `[`(x, i) ``)

And the tools of control flow:

* `if (cond) true` (`` `if`(cond, true) ``)
* `if (cond) true else false` (`` `if`(cond, true, false) ``)
* `for(var in seq) action` (`` `for`(var, seq, action) ``)
* `while(cond) action` (`` `while`(cond, action)  ``)
* `repeat expr` (`` `repeat`(expr) ``)
* `next` (`` `next`() ``)
* `break` (`` `break`() ``)

Finally, the most complex is the "function" function:

* `function(arg1, arg2) {body}` (`` `function`(alist(arg1, arg2), body, env))

Knowing the name of the function that underlies the special form is useful for getting documentation. `?(` is a syntax error; `` ?`(` `` will give you the documentation for parentheses.

Note that all special forms are implemented as primitive functions (i.e. in C); this means printing these functions is not informative:

```{r}
`for`
```

### Exercises

1. Rewrite the following code snippets into prefix form:

    ```{r, eval = FALSE}
    1 + 2 + 3
    
    1 + (2 + 3)
    
    if (length(x) <= 5) x[[5]] else x[[n]]
    ```

1. Create a replacement function that modifies a random location in a vector.

1. Write your own version of `+` that will paste its inputs together if
   they are character vectors but behaves as usual otherwise. In other words,
   make this code work:
   
    ```{r, eval = FALSE}
    1 + 2
    #> [1] 3
    
    "a" + "b"
    #> [1] "ab"
    ```

1. Create a list of all the replacement functions found in the base package. 
   Which ones are primitive functions? (Hint use `apropros()`)

1. What are valid names for user-created infix functions?

1. Create an infix `xor()` operator.

1. Create infix versions of the set functions `intersect()`, `union()`, 
   `setdiff()`. You might call them `%n%`, `%u%`, and `%/%` to match 
   conventions from mathematics.

## Function arguments {#function-arguments}

### Lazy evaluation {#lazy-evaluation}
\index{lazy evaluation} 
\index{functions!lazy evaluation}

In R, function arguments are __lazily evaluated__, which means that they're only evaluated if they're accessed. For example, this code doesn't generate an error because `x` is never used:

```{r}
f <- function(x) {
  10
}
f(stop("This is an error!"))
```

One place that this laziness is useful is in if statements. For example, in the code block below, the second competent of the if statement (`x > 0`) will only be evaluated if the first (`!is.null(x)`) is true. If it wasn't, the statement would return an error because `NULL > 0` is a logical vector of length 0 and not a valid input to `if`.

```{r, eval = FALSE}
x <- NULL
if (!is.null(x) && x > 0) {

}
```

### Forcing evaluation
\indexc{force()}

If you want to ensure that an argument is __eagerly__ evaluated, use `force()`: 

```{r, error = TRUE}
f <- function(x) {
  force(x)
  10
}
f(stop("This is an error!"))
```

The definition of `force()` is simple:

```{r}
force
```

However, it's better to use `force(x)` than just `x`, even though they do the same thing: `force()` provides a clear signal that you're deliberately forcing evaluation.

It is usually not necessary to force evaluation. It's primary usage is in higher order functions which return functions. Take this relative simple function. It takes a single argument, and returns a function that when called returns that function.

```{r}
g <- function(x) {
  function() {
    x
  }
}
```

There's a subtle bug in the definition of this function: the value of `x` is captured not when you call `g()`, but when you call the function that `g()` returns:

```{r}
x <- 10
g1 <- g(x)

x <- 20
g1()
```

And even more confusingly this only happens once. Once you have called the function returned by `g()`, the value is locked in:

```{r}
x <- 30
g1()
```

We can avoid this problem by forcing `x`:

```{r}
g <- function(x) {
  force(x)
  
  function() {
    x
  }
}

x <- 10
g1 <- g(x)

x <- 20
g1()
```

This is most important in functional programming, particularly when creating functional operators. In older versions of R, this was also a problem if you created functionals using an apply function, but as of R 3.2.0 and later, the base apply functions automatically force their inputs. 

### Promises
\index{promises}
\index{thunks|see{promises}}

Lazy evaluation is powered by a data structure called a __promise__, or (less commonly) a thunk. A promise has two components: 

* The __expression__, like `x + y` which gives rise to the delayed computation.

* The __environment__ where the expression should be evaluated.

You'll learn more about these ideas in [expressions], and [environment], and then see their fullest expression in [translation].

The first time a promise is accessed the expression is evaluated in the environment where it was created. This value is cached, so that subsequent access to the evaluated promise does not recompute the value. The following code only computes `runif(1)` once:

```{r}
j <- function(x) { 
  c(x, x, x)  
}

j(runif(1))
```

You can also create promises "by hand" using `delayedAssign()`:

```{r}
delayedAssign("x", {print("Executing code"); runif(1)})
x
x
```

We'll come back to this idea again in [advanced bindings].

### Default arguments
\index{functions!default values}

Another important property of function arguments is their default value. These are used if the caller does not supply a value:

```{r}
f <- function(a = 1, b = 2) {
  c(a, b)
}
f()
```

Since arguments in R are evaluated lazily, the default value can be defined in terms of other arguments, or even in terms of variables defined later in the function:

```{r}
g <- function(x = 1, y = x * 2, z = a + b) {
  a <- 10
  b <- 100
  
  c(x, y, z)
}

g()
```

While base R functions use this technique frequently, I don't recommend it as it tends to make code harder to understand as it relies on exactly when an argument is evaluated.

The evaluation environment is slightly different for default and user supplied arguments, as default arguments are evaluated inside the function. This means that seemingly identical calls can yield different results. It's easiest to see this with an extreme example:

```{r}
f <- function(x = ls()) {
  a <- 1
  x
}

# ls() evaluated inside f:
f()

# ls() evaluated in global environment:
f(ls())
```

### Missing arguments
\indexc{missing()}

If an argument has a default, you can determine if the value comes from the user or the defaultwith `missing()`:

```{r}
i <- function(x = 10) {
  list(missing(x), x)
}
str(i())
str(i(10))
```

I think `missing()` is best used sparingly. While you can use it for optionalal arguments, I think it's bad practice as it makes it harder to determine which arguments are required. Instead, I think it's better to use a default value of `NULL`. rlang provides `%||%` to make this even easier:

```{r}
`%||%` <- function(x, y) {
  if (is.null(x)) {
    y
  } else {
    x
  }
}

NULL %||% 10
10 %||% 20
```

Lazy evaluation means that you can put an expensive computation on the RHS, and it will only be run if the user did not provide a default argument:

```{r}
my_fun <- function(x, y, z = NULL) {
  z <- z %||% default_z(x, y)
}
```

Compare this with the equivalent code using a missing argument. The main problem is that when you look at the function definition, there's no way to tell that `z` is optional.

```{r}
my_fun <- function(x, y, z) {
  if (missing(z)) {
    z <- default_z(x, y)
  }
}
```

### `...` {#fun-dot-dot-dot}
\indexc{...}

There is a special argument called `...`. If a function has this argument, it can take an number arguments as it match any arguments not otherwise matched. If you use `...` in the function arguments, you can use it inside your function to pass those arguments on to another function:

```{r}
f <- function(x, ...) {
  g(...)
}

g <- function(y, z) {
  y + 2 * z
}

f(x = 1, y = 2, z = 3)
```

It's possible (but rarely useful) to refer to elements of `...` by their position, using a special form:

```{r}
f <- function(...) {
  list(first = ..1, third = ..3)
}
str(f(1, 2, 3))
```

If you want to capture the values of `...` in a form that you can compute upon, it's typically more useful to put them in a list: 

```{r}
f <- function(...) {
  list(...)
}
str(f(a = 1, b = 2))
```

(See also `rlang::list2()` for support for splicing and to silently ignore trailing commas)

`...` is useful if you want to collect arguments to call another function, but you don't want to prespecify their possible names. There are two primary uses of `...`:

*   If your function takes a function as an argument, you want some way to 
    pass on additional arguments to that function. For example, here `lapply()`
    uses `...` to pass `na.rm` on to `mean()`:
    
    ```{r}
    x <- list(c(1, 3, NA), c(4, NA, 6))
    str(lapply(x, mean, na.rm = TRUE))
    ```
    
*   If your function is an S3 generic, you need some way to allow methods to 
    take arbitrary extra arguments. For example, one relatively sophisticated 
    user of `...` is the base `plot()` function, which has argument `x`, `y` 
    and `...` . To understand what `...` does for a given function we need to 
    read the help: "Arguments to be passed to methods, such as graphical
    parameters". Most simple invocations of `plot()` end up calling
    `plot.default()` which has many more arguments, but also has `...`.  
    Again, reading the documentation reveals that `...` accepts "other 
    graphical parameters", which are listed in the help for `par()`.  This 
    allows us to write code like:
    
    ```{r, eval = FALSE}
    plot(1:5, col = "red")
    plot(1:5, cex = 5, pch = 20)
    ```

Using `...` comes with two downsides:

*   When you use it to pass arguments on to another function, you have to 
    carefully explain to the user where those arguments go. This makes it
    hard to understand the what you can do with functions like `lapply()` and 
    `plot()`
    
*   Any misspelled arguments will not raise an error.  This makes it easy for 
    typos to go unnoticed:

    ```{r}
    sum(1, 2, NA, na.mr = TRUE)
    ```

`...` is a powerful tool, but be aware of the downsides.

### Exercises

1.  Clarify the following list of odd function calls:

    ```{r, eval = FALSE}
    x <- sample(replace = TRUE, 20, x = c(1:10, NA))
    y <- runif(min = 0, max = 1, 20)
    cor(m = "k", y = y, u = "p", x = x)
    ```

1.  What does this function return? Why? Which principle does it illustrate?
  
    ```{r, results = "hide"}    
    f1 <- function(x = {y <- 1; 2}, y = 0) {
      x + y
    }
    f1()
    ```

1.  What does this function return? Why? Which principle does it illustrate?

    ```{r, results = "hide"}
    f2 <- function(x = z) {
      z <- 100
      x
    }
    f2()
    ```
  
1.  In `hist()`, the default value of `xlim` is `range(breaks)`, the default 
    value for `breaks` is `"Sturges"`, and 

    ```{r}
    range("Sturges")
    ```
    
    Explain how `hist()` works to get a correct `xlim` value.
    
1.  Explain why this function works. Why is it confusing?

    ```{r}
    show_time <- function(x = stop("Error!")) {
      stop <- function(...) Sys.time()
      print(x)
    }
    show_time()
    ```

## Return values {#return-values}

The last expression evaluated in a function becomes the return value, the result of invoking the function. \index{functions!return value}

```{r}
f <- function(x) {
  if (x < 10) {
    0
  } else {
    10
  }
}
f(5)
f(15)
```

Generally, I think it's good style to reserve the use of an explicit `return()` for when you are returning early, such as for an error, or a simple case of the function. This style of programming can also reduce the level of indentation, and generally make functions easier to understand because you can reason about them locally. \indexc{return()}

```{r}
f <- function(x, y) {
  if (!x) return(y)

  # complicated processing here
}
```

Functions can return only a single object. But this is not a limitation because you can return a list containing any number of objects.

### Purity

The functions that are the easiest to understand and reason about are pure functions: functions that always map the same input to the same output and have no other impact on the workspace. In other words, pure functions have no __side effects__: they don't affect the state of the world in any way apart from the value they return. \index{pure functions}

Most base R functions are pure, with a few notable exceptions:

* `library()` which loads a package, and hence modifies the search path.

* `setwd()`, `Sys.setenv()`, `Sys.setlocale()` which change the working 
  directory, environment variables, and the locale, respectively.

* `plot()` and friends which produce graphical output.

* `write()`, `write.csv()`, `saveRDS()`, etc. which save output to disk.

* `options()` and `par()` which modify global settings.

* S4 related functions which modify global tables of classes and methods.

* Random number generators which produce different numbers each time you 
  run them.

It's generally a good idea to minimise the use of side effects, and where possible, to minimise the footprint of side effects by separating pure from impure functions. Pure functions are easier to test (because all you need to worry about are the input values and the output), and are less likely to work differently on different versions of R or on different platforms. For example, this is one of the motivating principles of ggplot2: most operations work on an object that represents a plot, and only the final `print` or `plot` call has the side effect of actually drawing the plot.

### Invisible values

Functions can return `invisible` values, which are not printed out by default when you call the function. \indexc{invisible()} \index{functions!invisible results}

```{r}
f1 <- function() 1
f2 <- function() invisible(1)

f1()
f2()
f1() == 1
f2() == 1
```

You can force an invisible value to be displayed by wrapping it in parentheses:

```{r}
(f2())
```

The most common function that returns invisibly is `<-`: \index{assignment}

```{r}
a <- 2
(a <- 2)
```

This is what makes it possible to assign one value to multiple variables:

```{r}
a <- b <- c <- d <- 2
```

because this is parsed as:

```{r}
(a <- (b <- (c <- (d <- 2))))
```

### On exit {#on-exit}

As well as returning a value, functions can set up other triggers to occur when the function is finished using `on.exit()`. This is often used as a way to guarantee that changes to the global state are restored when the function exits. The code in `on.exit()` is run regardless of how the function exits, whether with an explicit (early) return, an error, or simply reaching the end of the function body. \indexc{on.exit()}

```{r}
in_dir <- function(dir, code) {
  old <- setwd(dir)
  on.exit(setwd(old))

  force(code)
}
getwd()
in_dir("~", getwd())
```

The basic pattern is simple:

* We first set the directory to a new location, capturing the current location 
  from the output of `setwd()`.

* We then use `on.exit()` to ensure that the working directory is returned to 
  the previous value regardless of how the function exits.

* Finally, we explicitly force evaluation of the code. (We don't actually need 
  `force()` here, but it makes it clear to readers what we're doing.)

**Caution**: If you're using multiple `on.exit()` calls within a function, make sure to set `add = TRUE`. Unfortunately, the default in `on.exit()` is `add = FALSE`, so that every time you run it, it overwrites existing exit expressions. Because of the way `on.exit()` is implemented, it's not possible to create a variant with `add = TRUE`, so you must be careful when using it.

### Exercises

1.  How does the `chdir` parameter of `source()` compare to `in_dir()`? Why 
    might you prefer one approach to the other?

1.  What function undoes the action of `library()`? How do you save and restore
    the values of `options()` and `par()`?

1.  Write a function that opens a graphics device, runs the supplied code, and 
    closes the graphics device (always, regardless of whether or not the plotting 
    code worked).

1.  We can use `on.exit()` to implement a simple version of `capture.output()`.

    ```{r}
    capture.output2 <- function(code) {
      temp <- tempfile()
      on.exit(file.remove(temp), add = TRUE)

      sink(temp)
      on.exit(sink(), add = TRUE)

      force(code)
      readLines(temp)
    }
    capture.output2(cat("a", "b", "c", sep = "\n"))
    ```

    Compare `capture.output()` to `capture.output2()`. How do the functions 
    differ? What features have I removed to make the key ideas easier to see? 
    How have I rewritten the key ideas to be easier to understand?


## Quiz answers {#function-answers}

1.  The three components of a function are its body, arguments, and environment.

1.  `f1(1)()` returns 11.

1.  You'd normally write it in infix style: `1 + (2 * 3)`.

1.  Rewriting the call to `mean(c(1:10, NA), na.rm = TRUE)` is easier to
    understand.
    
1.  No, it does not throw an error because the second argument is never used 
    so it's never evaluated.

1.  See [infix](#infix-functions) and 
    [replacement functions](#replacement-functions).

1.  You use `on.exit()`; see [on exit](#on-exit) for details.
