# Functions

```{r setup, include = FALSE}
source("common.R")
```

## Introduction
\index{functions}

If you're reading this book, I expect that you've already created many R functions, and you know how to use them to reduce duplication in your code. In this chapter, you'll learn how to turn your existing, informal knowledge into a rigorous understanding of functions. This chapter will primarily help you develop your theoretical knowledge of functions. While you'll see some interesting tricks and techniques, much of what you'll learn will be most important for understanding more advanced techniques later in the book. 

### Quiz {-}

Answer the following questions to see if you can safely skip this chapter. You can find the answers at the end of the chapter in [answers](#function-answers).

1.  What are the three components of a function?

1.  What does the following code return?

    ```{r, eval = FALSE}
    x <- 10
    f1 <- function(x) {
      function() {
        x + 10
      }
    }
    f1(1)()
    ```
    
1.  How would you usually write this code?

    ```{r, eval = FALSE}
    `+`(1, `*`(2, 3))
    ```
    
1.  How could you make this call easier to read?

    ```{r, eval = FALSE}
    mean(, TRUE, x = c(1:10, NA))
    ```

1.  Does the following function throw an error when called? Why/why not?

    ```{r, eval = FALSE}
    f2 <- function(a, b) {
      a * 10
    }
    f2(10, stop("This is an error!"))
    ```

1.  What is an infix function? How do you write it? What's a replacement 
    function? How do you write it?

1.  How do you ensure that cleanup action occurs regardless of how a function 
    exits?

### Outline {-}

* [Function components](#function-components) describes the three main 
  components of a function.

* [Lexical scoping](#lexical-scoping) teaches you how R finds values from 
  names, the process of lexical scoping.

* [Every operation is a function call](#all-calls) shows you that everything
  that happens in R is a result of a function call, even if it doesn't look 
  like it.

* [Function arguments](#function-arguments) discusses the three ways of 
  supplying arguments to a function, how to call a function given a list of 
  arguments, and the impact of lazy evaluation.

* [Special calls](#special-calls) describes two special types of function: 
  infix and replacement functions.
  
* [Return values](#return-values) discusses how and when functions return
  values, and how you can ensure that a function does something before it
  exits.

## Function basics

The most important thing to to understand about R is that functions are objects in their own right, a language property often called "first-class functions". Other unlike many other languages, there is no special syntax for defining a function with a given name; instead you create a function object (with `function`) and bind it to a name with `<-`:

```{r}
my_fancy_function <- function(x) {
  sin(1 / x ^ 2)
}
```

This also means that you can put functions in a list:

```{r}
funs <- list(
  half = function(x) x / 2,
  double = function(x) x * 2
)

funs$double(10)
```

In many places in R, you'll see functions called __closures__. This is a technical name for the type of functions that the R language has because R functions capture, or __enclose__, their environments.

```{r}
typeof(my_fancy_function)
```

### Function components
\index{functions!body} 
\indexc{body()} 
\index{functions!formals} 
\indexc{formals()} 
\index{functions!environment}
\index{environments!of a function}

A function has three parts: 

* The `formals()`, the list of arguments that control how you call the function.
  
* The `body()`, the code inside the function.

* The `environment()`, which determines how the function finds values 
  associated with names.

The formals and body are specified explicitly when you create a function. The environment is specified implicitly, based on _where_ you defined the function. The function environment is only printed when the function wasn't defined in the global environment, but it always exists.

```{r}
f <- function(x) {
  # A comment
  x ^ 2
}

formals(f)

body(f)

environment(f)
```

Like all objects in R, functions can also possess any number of additional `attributes()`. One attribute used by base R is "srcref", short for source reference, which points to the source code used to create the function. The srcref is used for printing because, unlike `body()`, it contains code comments and other formatting. 

```{r}
attr(f, "srcref")
```

### Primitive functions
\index{primitive functions} 
\index{functions!primitive|see{primitive functions}} 
\indexc{.Primitive}

There is one exception to the rule that functions have three components. Primitive functions, like `sum()` and `[`, call C code directly. 

```{r}
sum
`[`
```

They have type "builtin" or "special":

```{r}
typeof(sum)
typeof(`[`)
```

These functions exist primarily in C, not R, so their `formals()`, `body()`, and `environment()` are all `NULL`:  

```{r}
formals(sum)
body(sum)
environment(sum)
```

Primitive functions are only found in the base package. Primitive functions have certain performance advantages, but come at the cost of increased implementation complexity. Primitive functions are harder to write, and R-core generally avoids creating them unless there is no other option.

### Exercises

1.  What function allows you to tell if an object is a function? What function
    allows you to tell if a function is a primitive function?

1.  This code makes a list of all functions in the base package. 
    
    ```{r}
    objs <- mget(ls("package:base"), inherits = TRUE)
    funs <- Filter(is.function, objs)
    ```

    Use it to answer the following questions:

    a. Which base function has the most arguments?
    
    a. How many base functions have no arguments? What's special about those
       functions?
       
    a. How could you adapt the code to find all primitive functions?

1. What are the three important components of a function?

1. When does printing a function not show what environment it was created in?

## Lexical scoping {#lexical-scoping}
 \index{scoping!lexical|see{lexical scoping}} \index{lexical scoping}
 
In [Names and values], we discussed assignment, the act of binding a name to a value. __Scoping__ works in the opposite direction: how do you find the value associated with a given name? 

The basic rules of scoping are quite intuitive, and you've probably already internalised them, even if you've never been explicitly taught them. For example, what will the following code return, 10 or 20?[^answer1]

[^answer1]: I'll "hide" the answers to these challenges in footnotes. Try and predict the answer before you look it up; that will help you remember the correct answer better. In the case, `f()` will return `20`.

```{r, results = "hide"}
x <- 10
f <- function() {
  x <- 20
  x
}

f()
```

In this section you'll learn the rules of scoping more formally, and learn some of the subtler details you might not have yet encounted. A deeper understanding of scoping will help you use advanced functional programming tools, and eventually, to write tools that translate R code to other languages.

R uses __lexical scoping__[^dyn-scope], which looks up the values of names based on how functions were defined, not how they are called. Here "lexical" doesn't have to the usual English definition, but is instead a technical CS term, implying that the scoping rules can use parse-time structure, not run-time structure. R's lexical scoping follows four primary rules:

* Name masking
* Functions vs. variables
* A fresh start
* Dynamic lookup

[^dyn-scope]: Functions that automatically quote one or more arguments (sometimes called NSE functions) can override the default scoping rules to implement other varieties of scoping. You'll learn more about that in [metaprogramming](#meta).

### Name masking

The following example illustrates the most basic principle of lexical scoping: names in defined inside a function override names defined outside of it.

```{r}
x <- 10
y <- 20
f <- function() {
  x <- 1
  y <- 2
  c(x, y)
}
f()
```

If a name isn't defined inside a function, R will look one level up.

```{r}
x <- 2
g <- function() {
  y <- 1
  c(x, y)
}
g()
```

The same rules apply if a function is defined inside another function: look inside the current function, then where that function was defined, and so on, all the way up to the global environment, and then on to other loaded packages. 

Run the following code in your head, then confirm the output by running the R code.[^answer2]

[^answer2]: `h()` returns `c(1, 2, 3)`.

```{r, results = "hide"}
x <- 1
h <- function() {
  y <- 2
  i <- function() {
    z <- 3
    c(x, y, z)
  }
  i()
}
h()
```

The same rules apply to functions created by other functions, which we'll call __closures__. Closures will be described in more detail in [closures]; here we'll just look at how they interact with scoping. The following function, `j()`, returns a function.  What do you think this function will return when we call it?[^answer3]

[^answer3]: `k()` returns `c(10, 2)`.
\index{closures!scoping}

```{r, results = "hide"}
x <- 10
y <- 20

j <- function() {
  y <- 2
  function() {
    c(x, y)
  }
}
k <- j()
k()
```

This seems a little magical: how does R know what the value of `y` is after `j()` has returned? It works because `k` preserves the environment in which it was defined and because the environment includes the value of `y`. You'll learn more about how environments work in [Environments](#environments).

### Functions vs. variables

Because R functions are ordinary objects, the same scoping principles apply to functions:

```{r}
l <- function(x) x + 1
m <- function() {
  l <- function(x) x + 100
  l(10)
}
m()
```

The rule gets a little more complicated when a name is bound to a function and a non-function in different environments. When you use name in function call, R will ignore non-function objects while looking for the value. For example, here `n` takes on two different values:

```{r}
n <- function(x) x + 100
o <- function() {
  n <- 10
  n(n)
}
o()
```

But using the same name for two different thingswill make for confusing code, and is best avoided!

### A fresh start {#fresh-start}

What happens to the values in between invocations of a function? What will happen the first time you run this function? What will happen the second time?[^answer4] (If you haven't seen `exists()` before: it returns `TRUE` if there's a variable of that name, otherwise it returns `FALSE`.)

[^answer4]: `p()` returns `1` every time it is called.

```{r, results = "hide"}
p <- function() {
  if (!exists("a")) {
    a <- 1
  } else {
    a <- a + 1
  }
  a
}

p()
p()
```

You might be surprised that `p()` always returns the same value. That happens because a new environment is created to host its execution every time a function is called. That means a function has no way to tell what happened the last time it was run; each invocation is completely independent. (We'll see some ways to get around this in [mutable state](#mutable-state).)

### Dynamic lookup

Lexical scoping determines where to look for values, not when to look for them. R looks for values when the function is run, not when it's created. This means that the output of a function can differ depending on objects outside its environment: 

```{r}
q <- function() x
x <- 15
q()

x <- 20
q()
```

This behaviour can be quite annoying. If you make a spelling mistake in your code, you won't get an error when you create the function, and you might not even get one when you run the function, depending on what variables are defined in the global environment. 

\indexc{findGlobals()}
One way to detect this problem is to use `codetools::findGlobals()`. This function lists all the external dependencies (unbound symbols) within a function: 

```{r}
r <- function() x + 1
codetools::findGlobals(f)
```

Another way to try and solve the problem would be to manually change the environment of the function to the `emptyenv()`, an environment which contains nothing:

```{r, error = TRUE}
environment(r) <- emptyenv()
r()
```

Both of these approaches reveal why this undesirable behaviour exists: R relies on lexical scoping to find _everything_, even the `+` operator. This provides a rather beautiful simplicity to R's scoping rules.

### Exercises

1. What does the following code return? Why? Describe how each of the three
   `c`'s is interpreted.

    ```{r, eval = FALSE}
    c <- 10
    c(c = c)
    ```

1. What are the four principles that govern how R looks for values?

1. What does the following function return? Make a prediction before 
   running the code yourself.

    ```{r, results = "hide"}
    f <- function(x) {
      f <- function(x) {
        f <- function() {
          x ^ 2
        }
        f() + 1
      }
      f(x) * 2
    }
    f(10)
    ```

## Lazy evaluation {#lazy-evaluation}
\index{lazy evaluation} 
\index{functions!lazy evaluation}

In R, function arguments are __lazily evaluated__, which means that they're only evaluated if they're accessed. For example, this code doesn't generate an error because `x` is never used:

```{r}
f <- function(x) {
  10
}
f(stop("This is an error!"))
```

This is an important part of the R language because it means that you can provide potential expensive computations in function arguments, and they will only be evaluated if needed.

### Forcing evaluation
\indexc{force()}

To __eagerly__ evaluate an argument, i.e. to force it be immediately evaluated, use `force()`: 

```{r, error = TRUE}
f <- function(x) {
  force(x)
  10
}
f(stop("This is an error!"))
```

It is usually not necessary to force evaluation. It's needed primarily for certain functional programming techniques which we'll cover in detail in [function operators]. Here, I want to show you the basic issue.

Take this small but surprisingly tricky function. It takes a single argument `x`, and returns a function that returns `x` when called.

```{r}
capture1 <- function(x) {
  function() {
    x
  }
}
```

There's a subtle issue with this function: the value of `x` will be captured not when you call `capture()`, but when you call the function that `capture()` returns:

```{r}
x <- 10
g1 <- capture1(x)

x <- 20
g1()
```

Even more confusingly this only happens once: after you have called `g1()`, the value is locked in.

```{r}
x <- 30
g1()
```

This behaviour is a consequence of lazy evaluation. The `x` argument is evaluated once `g1()` is called, and then its vaue is cached. 

We can avoid the confusion by forcing `x`:

```{r}
capture2 <- function(x) {
  force(x)
  
  function() {
    x
  }
}

x <- 10
g1 <- capture2(x)

x <- 20
g1()
```

### Promises
\index{promises}
\index{thunks|see{promises}}

Lazy evaluation is powered by a data structure called a __promise__, or (less commonly) a thunk. We'll come back to this data structure in [metaprogramming](#meta) because it's one of the features of R that makes it most interesting as a programming language.

A promise has three components: 

* The expression, like `x + y` which gives rise to the delayed computation.

* The environment where the expression should be evaluated.

* The value, which is computed and cached when the promise is first accessed
  by evaluating the expression in the specified environment.

The value cache ensures that accessing the promise multiple times always returns the same value. For example, you can see in the following code that `runif(1)` is only evaluated once:

```{r}
j <- function(x) { 
  c(x, x, x)  
}

j(runif(1))
```

You can also create promises "by hand" using `delayedAssign()`:

```{r}
delayedAssign("x", {print("Executing code"); runif(1)})
x
x
```

You'll see this idea again in [advanced bindings].

### Default arguments
\index{functions!default values}

Thanks to lazy evaluation, default value can be defined in terms of other arguments, or even in terms of variables defined later in the function:

```{r}
g <- function(x = 1, y = x * 2, z = a + b) {
  a <- 10
  b <- 100
  
  c(x, y, z)
}

g()
```

Many base R functions use this technique, but I don't recommend it. It makes code harder to understand because it requires that you know exactly _when_ default arguments are evaluated in order to predict _what_ they will evaluate to.

The evaluation environment is slightly different for default and user supplied arguments, as default arguments are evaluated inside the function. This means that seemingly identical calls can yield different results. It's easiest to see this with an extreme example:

```{r, eval = FALSE}
f <- function(x = ls()) {
  a <- 1
  x
}

# ls() evaluated inside f:
f()
#> [1] "a" "x"

# ls() evaluated in global environment:
f(ls())
#> [1] "f"
```

### Missing arguments
\indexc{missing()}

If an argument has a default, you can determine if the value comes from the user or the default with `missing()`:

```{r}
i <- function(x = 10) {
  list(missing(x), x)
}
str(i())
str(i(10))
```

`missing()` is best used sparingly. Take `sample()`, for example. How many arguments are required?

```{r}
args(sample)
```
 
It looks like both `x` and `size` are required, but in fact `sample()` uses `missing()` to provide a default for `size` if it's not supplied. If I was to rewrite sample myself[^sample], I'd use an explicit `NULL` to indicate that `size` can be supplied, but it's not required:

[^sample]: Note that this only implements one way of calling `sample()`: you can also call it with a single integer, like `sample(10)`. This unfortunately makes `sample()` prone to silent errors in situations like `sample(x[i])`.

```{r}
sample <- function(x, size = NULL, replace = FALSE, prob = NULL) {
  if (is.null(size)) {
    size <- length(x)
  }
  
  x[sample.int(length(x), size, replace = replace, prob = prob)]
}
```

You can make that pattern even simpler with a small helper. The infix `%||%` function uses the LHS if it's not null, otherwise it uses the RHS:

```{r}
`%||%` <- function(lhs, rhs) {
  if (!is.null(lhs)) {
    lhs
  } else {
    rhs
  }
}

sample <- function(x, size = NULL, replace = FALSE, prob = NULL) {
  size <- size %||% length(x)
  x[sample.int(length(x), size, replace = replace, prob = prob)]
}
```

Because of lazy evaluation, you don't need to worry about unnecessary computation: the RHS of `%||%` will only be evaluated if the LHS is null.

### Exercises

1.  What important property of `&&` make `x_ok()` work?

    ```{r}
    x_ok <- function(x) {
      !is.null(x) && length(x) == 1 && x > 0
    }
    
    x_ok(NULL)
    x_ok(1)
    x_ok(1:3)
    ```

    What is different with this code? Why is this behaviour undesirable here?
    
    ```{r}
    x_ok <- function(x) {
      !is.null(x) & length(x) == 1 & x > 0
    }
    
    x_ok(NULL)
    x_ok(1)
    x_ok(1:3)
    ```

1.  The definition of `force()` is simple:

    ```{r}
    force
    ```
    
    Why is it better to `force(x)` instead of just `x`?
    
1.  What does this function return? Why? Which principle does it illustrate?

    ```{r, results = "hide"}
    f2 <- function(x = z) {
      z <- 100
      x
    }
    f2()
    ```

1.  What does this function return? Why? Which principle does it illustrate?
  
    ```{r, results = "hide"}
    y <- 10
    f1 <- function(x = {y <- 1; 2}, y = 0) {
      c(x, y)
    }
    f1()
    y
    ```
  
1.  In `hist()`, the default value of `xlim` is `range(breaks)`, the default 
    value for `breaks` is `"Sturges"`, and

    ```{r}
    range("Sturges")
    ```
    
    Explain how `hist()` works to get a correct `xlim` value.
    
1.  Explain why this function works. Why is it confusing?

    ```{r}
    show_time <- function(x = stop("Error!")) {
      stop <- function(...) Sys.time()
      print(x)
    }
    show_time()
    ```

1.  How many arguments are required when calling `library()`?

## `...`  (dot-dot-dot) {#fun-dot-dot-dot}
\indexc{...}
\index{varargs}
\index{function!variadic}
\index{ellipsis}
\index{dot-dot-dot}

Functions can have a special argument `...` (pronounced dot-dot-dot). If a function has this argument, it can take any number of additional arguments. In other programming languages, this type of argument is often called a varargs, or the function is said to be variadic. 

Inside a function, you can use `...` to pass those additional arguments on to another function:

```{r}
f <- function(x, ...) {
  g(...)
}

g <- function(y, z) {
  list(y = y, z = z)
}

str(f(x = 1, y = 2, z = 3))
```

It's possible (but rarely useful) to refer to elements of `...` by their position, using a special form:

```{r}
f <- function(...) {
  list(first = ..1, third = ..3)
}
str(f(1, 2, 3))
```

More often useful is `list(...)`, which evaluates the arguments are stores in a list:

```{r}
f <- function(...) {
  list(...)
}
str(f(a = 1, b = 2))
```

(See also `rlang::list2()` to support splicing and to silently ignore trailing commas, and `rlang::quos()` to capture the unevaluated arguments, the topic of [quasiquotation].)

There are two primary uses of `...`, both of which we'll come back to later in the book:

*   If your function takes a function as an argument, you want some way to 
    pass on additional arguments to that function. In this example, `lapply()`
    uses `...` to pass `na.rm` on to `mean()`:
    
    ```{r}
    x <- list(c(1, 3, NA), c(4, NA, 6))
    str(lapply(x, mean, na.rm = TRUE))
    ```
    
    We'll come back to this technique in Section \@ref(functional-dot-dot-dot).
    
*   If your function is an S3 generic, you need some way to allow methods to 
    take arbitrary extra arguments. For example, take the `print()` function.
    There are different options for printing types of object, so there's no
    way for the print generic to prespecify every possible argument. Instead,
    it uses `...` to allow individual methods to have different arguments:

    ```{r, eval = FALSE}
    print(factor(letters), max.levels = 4)
    
    print(y ~ x, showEnv = TRUE)
    ```
    
    We'll come back to this use of `...` in Section \@ref(s3-arguments).

Using `...` comes with two downsides:

*   When you use it to pass arguments on to another function, you have to 
    carefully explain to the user where those arguments go. This makes it
    hard to understand the what you can do with functions like `lapply()` and 
    `plot()`.
    
*   Any misspelled arguments will not raise an error.  This makes it easy for 
    typos to go unnoticed:

    ```{r}
    sum(1, 2, NA, na_rm = TRUE)
    ```

`...` is a powerful tool, but be aware of the downsides.

### Exercises

1.  Explain the following results:
    
    ```{r}
    sum(1, 2, 3)
    mean(1, 2, 3)
    
    sum(1, 2, 3, na.omit = TRUE)
    mean(1, 2, 3, na.omit = TRUE)
    ```

1.  In the following call, explain how to find the documentation for the
    named arguments in the following function call:
    
    ```{r, fig.asp = 1}
    plot(1:10, col = "red", pch = 20, xlab = "x", col.lab = "blue")
    ```
    
1.  Why does `plot(1:10, col = "red")` only colour the points, not the axes 
    or labels? Read the source code of `plot.default()` to find out.

## Exiting a function {#return-values}

Most functions exit in one of two ways[^esoterica]: either returning a value, indicating successful completion, or throwing an error, indicating failure. This section describes return values (implicit vs. explicit; visible vs. invisible), briefly discusses errors, and introduces exit handlers, which allow you to run code when a function exits, regardless of how it exits.

[^esoterica]: Functions can exit in other more esoteric ways like signalling a condition that is caught by an exiting handler, invoking a restart, or pressing "Q" in an interactive browser.

### Implicit vs. explict returns
\index{functions!return value}
\indexc{return()}

There are two ways that a function can return a value:

*   Implicitly, where the last evaluated expression becomes the return 
    value:

    ```{r}
    f <- function(x) {
      if (x < 10) {
        0
      } else {
        10
      }
    }
    f(5)
    f(15)
    ```

*   Explicitly, by calling `return()`:

    ```{r}
    f <- function(x) {
      if (x < 10) {
        return(0)
      } else {
        return(10)
      }
    }
    ```
    
### Invisible values
\indexc{invisible()} 
\index{functions!invisible results}

Most functions return visibly: calling the function in an interactive context causes the result to be automatically printed.

```{r}
f1 <- function() 1
f1()
```

However, it's also possible to return an `invisible()` value, which is not automatically printed.

```{r}
f2 <- function() invisible(1)

f2()
```

You can verify that the value exists either by explicitly printing it or by wrapping in parentheses:

```{r}
print(f2())

(f2())
```

Alternatively, use `withVisible()` to return the value and a visibility flag:

```{r}
str(withVisible(f2()))
```

\index{assignment}
The most common function that returns invisibly is `<-`: 

```{r}
a <- 2
(a <- 2)
```

And this is what makes it possible to chain assignment:

```{r}
a <- b <- c <- d <- 2
```

In general, any function called primarily for its side effects (like `<-`, `print()`, or `plot()`) should return an invisible value (typically the value of the first argument).

### Errors

If a function can not complete its assigned task, it should throw an error with `stop()`, which immediately terminates the execution of the function.

```{r, error = TRUE}
f <- function() {
  stop("I'm an error")
  return(10)
}
f()
```

Errors indicate that something has gone wrong, and force the user to handle them. Some languages (like C, go, and rust) rely on special return values to indicate problems, but in R you should always throw an error. You'll learn more about errors, and how to handle them, in [Conditions].

### Exit handlers {#on-exit}
\indexc{on.exit()}
\index{handler!exit}

Sometimes a function needs to make a temporary change to global state and you want to ensure those changes are restored when the function completes. It's painful to make sure you cleanup before any explicit return, and what happens if there's an error? Instead, you can set up an __exiting handler__ that is called when the function terminates, regardless of whether it returns a value or throws an error.

To setup an exiting handler, call `on.exit()` with the code to be run. It will execute when the function exits, regardless of what causes it to exit.

```{r, error = TRUE}
f <- function(x) {
  cat("Hello\n")
  on.exit(cat("Goodbye!\n"), add = TRUE)
  
  if (x) {
    return(10)
  } else {
    stop("Error")
  }
}

f(TRUE)

f(FALSE)
```

::: sidebar
Always set `add = TRUE` when using `on.exit()`. If you don't, each call to `on.exit()` will overwrite the previous exiting handler. Even when only registering a single handler, it's good practice to  set `add = TRUE` so that you don't get an unpleasant surprise if you later add more exit handlers
:::

`on.exit()` is important because it allows you to place clean-up actions next to actions with their cleanup operations. 

```{r}
cleanup <- function(dir, code) {
  old_dir <- setwd(dir)
  on.exit(setwd(old), add = TRUE)
  
  old_opt <- options(stringsAsFactors = FALSE)
  on.exit(options(old_opt), add = TRUE)
}
```

When coupled with lazy evaluation, this leads to a very useful pattern for running a block of code in an altered environment:

```{r}
with_dir <- function(dir, code) {
  old <- setwd(dir)
  on.exit(setwd(old), add = TRUE)

  force(code)
}

getwd()
with_dir("~", getwd())
```

See the [withr package](http://withr.r-lib.org) for a collection of functions of this nature.

### Exercises

1.  What does `load()` return? Why don't you normally see these values?

1.  What does `write.table()` return? What would be more useful?

1.  How does the `chdir` parameter of `source()` compare to `in_dir()`? Why 
    might you prefer one approach to the other?

1.  Write a function that opens a graphics device, runs the supplied code, and 
    closes the graphics device (always, regardless of whether or not the plotting 
    code worked).

1.  We can use `on.exit()` to implement a simple version of `capture.output()`.

    ```{r}
    capture.output2 <- function(code) {
      temp <- tempfile()
      on.exit(file.remove(temp), add = TRUE)

      sink(temp)
      on.exit(sink(), add = TRUE)

      force(code)
      readLines(temp)
    }
    capture.output2(cat("a", "b", "c", sep = "\n"))
    ```

    Compare `capture.output()` to `capture.output2()`. How do the functions 
    differ? What features have I removed to make the key ideas easier to see? 
    How have I rewritten the key ideas to be easier to understand?


## Function forms {#all-calls}

> "To understand computations in R, two slogans are helpful:
>
> * Everything that exists is an object.
> * Everything that happens is a function call."
>
> --- John Chambers

While everything that happens in R is a result of a function call, not all calls look the same. Function calls come in four varieties:

* In __prefix__ form, the function name comes before its arguments, like 
  `foofy(a, b, c)`. These constitute of the majority of function calls in R.

* In __infix__ form, the function name comes inbetween its arguments, like 
  `x + y`. Infix forms are used for many mathematical operators, as well as 
  user-defined functions that begin and end with `%`.
  
* A __replacement__ function assigns into what looks like a prefix function, 
  like `names(df) <- c("a", "b", "c")`.
  
* __Special forms__ like `[[`, `if`, and `for`, don't have a consistent 
  structure and provide some of the most important syntax in R.

While four forms exist, you only need to use one, because any call can be written in prefix form. I'll demonstrate this property, and then you'll learn about each of the forms in turn.

### Rewriting to prefix form
\indexc{`} 
\index{backticks|see{\texttt{`}}}

An interesting propety of R is every infix, replacement, or special form can be rewritten in prefix form. Rewriting in prefix form is useful because it helps you better understand the structure of the language, and it gives you the real name of every function. Knowing the real name of non-prefix functions is useful because it allows you to modify them for fun and profit.

The following example shows three pairs of equivalent calls, rewriting an infix form, replacement form, and a special form into prefix form.

```{r, eval = FALSE}
x + y
`+`(x, y)

names(df) <- c("x", "y", "z")
`names<-`(df, c("x", "y", "z"))

for(i in 1:10) print(i)
`for`(i, 1:10, print(i))
```

Knowing the function name of a non-prefix function allows you to override its behaviour. For example, if you're ever feeling particularly evil, run the following code while a friend is away from their computer. It will introduce a fun bug: 10% of the time, 1 will be added to any numeric calculation inside of parentheses.

```{r}
`(` <- function(e1) {
  if (is.numeric(e1) && runif(1) < 0.1) {
    e1 + 1
  } else {
    e1
  }
}
replicate(50, (1 + 2))
rm("(")
```

Of course, overriding built-in functions like this is a bad idea, but, as you'll learn about in [metaprogramming], it's possible to apply it only selected code blocks. This provides a clean and elegant approach to writing domain specific languages and translators to other languages.

A more useful technique is to use this knowledge when using functional programming tools. For example, could use `sapply()` to add 3 to every element of a list by first defining a function `add()`, like this: \indexc{sapply()}

```{r}
add <- function(x, y) x + y
sapply(1:10, add, 3)
```

But we can also get the same effect more simply by relying on the existing `+` function:

```{r}
sapply(1:5, `+`, 3)
```

We'll explore this idea in detail in [functionals].

### Prefix form
\index{functions!arguments}

The prefix form is the most common form in R code, and indeed in the majority of programming languages. Prefix calls in R are a little special because you can specify arguments in three ways:

* By position, like `help(mean)`.
* Using partial matching, like `help(to = mean)`.
* By name, like `help(topic = mean)`.

As illustrated by the following chunk, arguments are matched by exact name, then with unique prefixes, and finally by position.

```{r, error = TRUE}
f <- function(abcdef, bcde1, bcde2) {
  list(a = abcdef, b1 = bcde1, b2 = bcde2)
}
str(f(1, 2, 3))
str(f(2, 3, abcdef = 1))

# Can abbreviate long argument names:
str(f(2, 3, a = 1))

# But this doesn't work because abbreviation is ambiguous
str(f(1, 3, b = 1))
```

Generally, only use positional matching for the first one or two arguments; they will be the most commonly used, and most readers will know what they are. Avoid using positional matching for less commonly used arguments, and never use partial matching. See the tidyverse style guide, <http://style.tidyverse.org/syntax.html#argument-names>, for more advice.

### Infix functions
\index{functions!infix} 
\index{infix functions} 
\indexc{\%\%}

Infix functions are so called because the function name comes **in**between its arguments, and hence infix functions have two arguments. R comes with a number of built-in infix operators: `:`, `::`, `:::`, `$`, `@`, `^`, `*`, `/`, `+`, `-`, `>`, `>=`, `<`, `<=`, `==`, `!=`, `!`, `&`, `&&`, `|`, `||`, `~`, `<-`, and `<<-`. You can also create your own infix functions that start and end with `%`, and base R uses this to additionally define `%%`, `%*%`, `%/%`, `%in%`, `%o%`, and `%x%`.

Defining your own infix function is simple. You create a two argument function and bind it to a name that starts and ends with `%`:

```{r}
`%+%` <- function(a, b) paste0(a, b)
"new " %+% "string"
```

The names of infix functions are more flexible than regular R functions: they can contain any sequence of characters except "%". You will need to escape any special characters in the string used to define the function, but not when you call it:

```{r}
`% %` <- function(a, b) paste(a, b)
`%/\\%` <- function(a, b) paste(a, b)

"a" % % "b"
"a" %/\% "b"
```

R's default precedence rules mean that infix operators are composed from left to right:

```{r}
`%-%` <- function(a, b) paste0("(", a, " %-% ", b, ")")
"a" %-% "b" %-% "c"
```

There are two special infix functions that can be called with a single argument: `+` and `-`. 

```{r}
-1
+10
```

### Replacement functions {#replacement-functions}
\index{replacement functions} 
\index{functions!replacement}
\index{assignment!replacement functions}

Replacement functions act like they modify their arguments in place, and have the special name `xxx<-`. They typically have two arguments (`x` and `value`) and they must return the modified object. For example, the following function allows you to modify the second element of a vector: 

```{r}
`second<-` <- function(x, value) {
  x[2] <- value
  x
}
```

Replacement functions are used by placing the function call on the LHS of `<-`: 

```{r}
x <- 1:10
second(x) <- 5L
x
```

I say they "act" like they modify their arguments in place, because, as discussed in [Modify-in-place], they actually create a modified copy. We can see that by using `tracemem()`:

```{r, eval = FALSE}
x <- 1:10
tracemem(x)
#> <0x7ffae71bd880>

second(x) <- 6L
#> tracemem[0x7ffae71bd880 -> 0x7ffae61b5480]: 
#> tracemem[0x7ffae61b5480 -> 0x7ffae73f0408]: second<- 
```

If you want to supply additional arguments, they go inbetween `x` and `value`:

```{r}
`modify<-` <- function(x, position, value) {
  x[position] <- value
  x
}
modify(x, 1) <- 10
x
```

When you write `modify(x, 1) <- 10`, behind the scenes R turns it into:

```{r, eval = FALSE}
x <- `modify<-`(x, 1, 10)
```

Combining replacement with other functions requires more complex translation. For example, this:

```{r}
x <- c(a = 1, b = 2, c = 3)
names(x)

names(x)[2] <- "two"
names(x)
```

Is translated into:

```{r}
`*tmp*` <- x
x <- `names<-`(`*tmp*`, `[<-`(names(`*tmp*`), 2, "two"))
rm(`*tmp*`)
```

(Yes, it really does create a local variable named *tmp*, which is removed afterwards.)

### Special forms

Finally, there are a bunch of language features that are usually written in special ways, but also have prefix forms. These include parentheses:

* `(x)` (`` `(`(x) ``)
* `{x}` (`` `{`(x) ``).

The subsetting operators:

* `x[i]` (`` `[`(x, i) ``) 
* `x[[i]]` (`` `[`(x, i) ``)

And the tools of control flow:

* `if (cond) true` (`` `if`(cond, true) ``)
* `if (cond) true else false` (`` `if`(cond, true, false) ``)
* `for(var in seq) action` (`` `for`(var, seq, action) ``)
* `while(cond) action` (`` `while`(cond, action)  ``)
* `repeat expr` (`` `repeat`(expr) ``)
* `next` (`` `next`() ``)
* `break` (`` `break`() ``)

Finally, the most complex is the "function" function:

* `function(arg1, arg2) {body}` (`` `function`(alist(arg1, arg2), body, env)``)

Knowing the name of the function that underlies the special form is useful for getting documentation. `?(` is a syntax error; `` ?`(` `` will give you the documentation for parentheses.

Note that all special forms are implemented as primitive functions (i.e. in C); that means printing these functions is not informative:

```{r}
`for`
```

### Exercises

1. Rewrite the following code snippets into prefix form:

    ```{r, eval = FALSE}
    1 + 2 + 3
    
    1 + (2 + 3)
    
    if (length(x) <= 5) x[[5]] else x[[n]]
    ```

1.  Clarify the following list of odd function calls:

    ```{r, eval = FALSE}
    x <- sample(replace = TRUE, 20, x = c(1:10, NA))
    y <- runif(min = 0, max = 1, 20)
    cor(m = "k", y = y, u = "p", x = x)
    ```


1. Explain why the following code fails:

    ```{r, eval = FALSE}
    modify(get("x"), 1) <- 10
    #> Error: target of assignment expands to non-language object
    ```

1. Create a replacement function that modifies a random location in a vector.

1. Write your own version of `+` that will paste its inputs together if
   they are character vectors but behaves as usual otherwise. In other words,
   make this code work:
   
    ```{r, eval = FALSE}
    1 + 2
    #> [1] 3
    
    "a" + "b"
    #> [1] "ab"
    ```

1. Create a list of all the replacement functions found in the base package. 
   Which ones are primitive functions? (Hint use `apropros()`)

1. What are valid names for user-created infix functions?

1. Create an infix `xor()` operator.

1. Create infix versions of the set functions `intersect()`, `union()`, and
   `setdiff()`. You might call them `%n%`, `%u%`, and `%/%` to match 
   conventions from mathematics.


## Quiz answers {#function-answers}

1.  The three components of a function are its body, arguments, and environment.

1.  `f1(1)()` returns 11.

1.  You'd normally write it in infix style: `1 + (2 * 3)`.

1.  Rewriting the call to `mean(c(1:10, NA), na.rm = TRUE)` is easier to
    understand.
    
1.  No, it does not throw an error because the second argument is never used 
    so it's never evaluated.

1.  See [infix](#infix-functions) and 
    [replacement functions](#replacement-functions).

1.  You use `on.exit()`; see [on exit](#on-exit) for details.
