# Names and values {#names-values}

```{r setup, include = FALSE}
source("common.R")
library(pryr)
knitr::read_chunk("memory-read-delim.r")
options(scipen = 9)
```

## Introduction

<!--
Names and values
https://nedbatchelder.com/text/names1.html

Immutable values
Reference semantics
garbage collection
size of aliased objects
modification in place
call-by-value vs call-by-reference

for(x in 1:10) {}
x

-->

A solid understanding of R's memory management will help you predict how much memory you'll need for a given task and help you to make the most of the memory you have. It can even help you write faster code because accidental copies are a major cause of slow code. The goal of this chapter is to help you understand the basics of memory management in R, moving from individual objects to functions to larger blocks of code. Along the way, you'll learn about some common myths, such as that you need to call `gc()` to free up memory, or that `for` loops are always slow. \index{memory}

### Outline {-}

* [Object size](#object-size) shows you how to use `object_size()` to see 
  how much memory an object occupies, and uses that as a launching point to
  improve your understanding of how R objects are stored in memory.
  
* [Memory usage and garbage collection](#gc) introduces you to the `mem_used()`
  and `mem_change()` functions that will help you understand how R allocates 
  and frees memory.
  
* [Modification in place](#modification) introduces you to the `address()` and
  `refs()` functions so that you can understand when R modifies in place and 
  when R modifies a copy. Understanding when objects are copied is very 
  important for writing efficient R code.

### Prerequisites {-}

In this chapter, we'll use tools from the pryr and lineprof packages to understand memory usage, and a sample dataset from ggplot2. If you don't already have them, run this code to get the packages you need:

```{r, eval = FALSE}
install.packages("ggplot2")
install.packages("pryr")
install.packages("devtools")
devtools::install_github("hadley/lineprof")
```

### Sources {-}

<!-- FIXME: cite R-exts and R-ints formally -->

The details of R's memory management are not documented in a single place. Most of the information in this chapter was gleaned from a close reading of the documentation (particularly `?Memory` and `?gc`), the [memory profiling](http://cran.r-project.org/doc/manuals/R-exts.html#Profiling-R-code-for-memory-use) section of R-exts, and the [SEXPs](http://cran.r-project.org/doc/manuals/R-ints.html#SEXPs) section of R-ints. The rest I figured out by reading the C source code, performing small experiments, and asking questions on R-devel. Any mistakes are entirely mine.

## Binding


Assignment is the act of binding (or rebinding) a name to a value in an environment. It is the counterpart to scoping, the set of rules that determines how to find the value associated with a name. Compared to most languages, R has extremely flexible tools for binding names to values. In fact, you can not only bind values to names, but you can also bind expressions (promises) or even functions, so that every time you access the value associated with a name, you get something different! \index{bindings}

You've probably used regular assignment in R thousands of times. Regular assignment creates a binding between a name and an object in the current environment. Names usually consist of letters, digits, `.` and `_`, and can't begin with `_`.  If you try to use a name that doesn't follow these rules, you get an error:

```{r, eval = FALSE}
_abc <- 1
# Error: unexpected input in "_"
```

Reserved words (like `TRUE`, `NULL`, `if`, and `function`) follow the rules but are reserved by R for other purposes:

```{r, eval = FALSE}
if <- 10
#> Error: unexpected assignment in "if <-"
```

A complete list of reserved words can be found in `?Reserved`. \index{reserved names} \indexc{`} \index{non-syntactic names}

It's possible to override the usual rules and use a name with any sequence of characters by surrounding the name with backticks:

```{r, eval = FALSE}
`a + b` <- 3
`:)` <- "smile"
`    ` <- "spaces"
ls()
#  [1] "    "   ":)"     "a + b"
`:)`
#  [1] "smile"
```

::: sidebar
### Quotes {-}
You can also create non-syntactic bindings using single and double quotes instead of backticks, but I don't recommend it. The ability to use strings on the left hand side of the assignment arrow is a historical artefact, used before R supported backticks.
:::


## Object size {#object-size}

<!-- deletion size vs. serialization size -->

To understand memory usage in R, we will start with `pryr::object_size()`. This function tells you how many bytes of memory an object occupies: \index{object\_size()}

```{r}
library(pryr)
object_size(1:10)
object_size(mean)
object_size(mtcars)
```

(This function is better than the built-in `object.size()` because it accounts for shared elements within an object and includes the size of environments.)

Something interesting occurs if we use `object_size()` to systematically explore the size of an integer vector. The code below computes and plots the memory usage of integer vectors ranging in length from 0 to 50 elements. You might expect that the size of an empty vector would be zero and that memory usage would grow proportionately with length. Neither of those things are true! \index{vectors!size of}

```{r size-q, fig.height = 2.5, fig.width = 4, small_mar = TRUE}
sizes <- sapply(0:50, function(n) object_size(seq_len(n)))
plot(0:50, sizes, xlab = "Length", ylab = "Size (bytes)", 
  type = "s")
```

This isn't just an artefact of integer vectors. Every length 0 vector occupies 40 bytes of memory:

```{r}
object_size(numeric())
object_size(logical())
object_size(raw())
object_size(list())
```

Those 40 bytes are used to store four components possessed by every object in R:

* Object metadata (4 bytes). These metadata store the base type (e.g. integer) 
  and information used for debugging and memory management.

* Two pointers: one to the next object in memory and one to the previous 
  object (2 * 8 bytes). This doubly-linked list makes it easy for internal 
  R code to loop through every object in memory.

* A pointer to the attributes (8 bytes).

All vectors have three additional components: \indexc{SEXP}

* The length of the vector (4 bytes). By using only 4 bytes, you might expect 
  that R could only support vectors up to $2 ^ {4 \times 8 - 1}$ ($2 ^ {31}$, about 
  two billion) elements. But in R 3.0.0 and later, you can actually have 
  vectors up to $2 ^ {52}$ elements. [Read R-internals][long-vectors] to see how 
  support for long vectors was added without having to change the size of this 
  field. \index{long vectors} \index{atomic vectors!long}

* The "true" length of the vector (4 bytes). This is basically never used, 
  except when the object is the hash table used for an environment. In that 
  case, the true length represents the allocated space, and the length 
  represents the space currently used.

* The data (variable number of bytes). An empty vector has 0 bytes of data. Numeric vectors occupy 8 bytes for
  every element, integer vectors 4, and complex vectors 16.

If you're keeping count you'll notice that this only adds up to 36 bytes. The remaining 4 bytes are used for padding so that each component starts on an 8 byte (= 64-bit) boundary. Most cpu architectures require pointers to be aligned in this way, and even if they don't require it, accessing non-aligned pointers tends to be rather slow. (If you're interested, you can read more about it in [C structure packing](http://www.catb.org/esr/structure-packing/).)

This explains the intercept on the graph. But why does the memory size grow irregularly? To understand why, you need to know a little bit about how R requests memory from the operating system. Requesting memory (with `malloc()`) is a relatively expensive operation. Having to request memory every time a small vector is created would slow R down considerably. Instead, R asks for a big block of memory and then manages that block itself. This block is called the small vector pool and is used for vectors less than 128 bytes long. For efficiency and simplicity, it only allocates vectors that are 8, 16, 32, 48, 64, or 128 bytes long. If we adjust our previous plot to remove the 40 bytes of overhead, we can see that those values correspond to the jumps in memory use.

```{r size-a, fig.height = 2.5, fig.width = 4, small_mar = TRUE}
plot(0:50, sizes - 40, xlab = "Length", 
  ylab = "Bytes excluding overhead", type = "n")
abline(h = 0, col = "grey80")
abline(h = c(8, 16, 32, 48, 64, 128), col = "grey80")
abline(a = 0, b = 4, col = "grey90", lwd = 4)
lines(sizes - 40, type = "s")
```

Beyond 128 bytes, it no longer makes sense for R to manage vectors. After all, allocating big chunks of memory is something that operating systems are very good at. Beyond 128 bytes, R will ask for memory in multiples of 8 bytes. This ensures good alignment.

A subtlety of the size of an object is that components can be shared across multiple objects. For example, look at the following code:

```{r}
x <- 1:1e6
object_size(x)

y <- list(x, x, x)
object_size(y)
```

`y` isn't three times as big as `x` because R is smart enough to not copy `x` three times; instead it just points to the existing `x`. 

It's misleading to look at the sizes of `x` and `y` individually. If you want to know how much space they take up together, you have to supply them to the same `object_size()` call:

```{r}
object_size(x, y)
```

In this case, `x` and `y` together take up the same amount of space as `y` alone. This is not always the case. If there are no shared components, as in the following example, then you can add up the sizes of individual components to find out the total size:

```{r}
x1 <- 1:1e6
y1 <- list(1:1e6, 1:1e6, 1:1e6)

object_size(x1)
object_size(y1)
object_size(x1, y1)
object_size(x1) + object_size(y1) == object_size(x1, y1)
```

The same issue also comes up with strings, because R has a global string pool. This means that each unique string is only stored in one place, and therefore character vectors take up less memory than you might expect: \index{string pool}

```{r}
object_size("banana")
object_size(rep("banana", 10))
```

### Exercises

1.  Repeat the analysis above for numeric, logical, and complex vectors.

1.  If a data frame has one million rows, and three variables (two numeric, and 
    one integer), how much space will it take up? Work it out from theory, 
    then verify your work by creating a data frame and measuring its size.

1.  Compare the sizes of the elements in the following two lists. Each 
    contains basically the same data, but one contains vectors of small 
    strings while the other contains a single long string.

    ```{r}
    vec <- lapply(0:50, function(i) c("ba", rep("na", i)))
    str <- lapply(vec, paste0, collapse = "")
    ```

1.  Which takes up more memory: a factor (`x`) or the equivalent character 
    vector (`as.character(x)`)? Why?

1.  Explain the difference in size between `1:5` and `list(1:5)`.

## Memory usage and garbage collection {#gc}

While `object_size()` tells you the size of a single object, `pryr::mem_used()` tells you the total size of all objects in memory: \indexc{mem\_used()}

```{r}
library(pryr)
mem_used()
```

This number won't agree with the amount of memory reported by your operating system for a number of reasons:

1. It only includes objects created by R, not the R interpreter itself.

1. Both R and the operating system are lazy: they won't reclaim memory 
   until it's actually needed. R might be holding on to memory because 
   the OS hasn't yet asked for it back.

1. R counts the memory occupied by objects but there may be gaps due to 
   deleted objects. This problem is known as memory fragmentation.

`mem_change()` builds on top of `mem_used()` to tell you how memory changes during code execution. Positive numbers represent an increase in the memory used by R, and negative numbers represent a decrease. \indexc{mem\_change()}

```{r}
# Need about 4 mb to store 1 million integers
mem_change(x <- 1:1e6)
# We get that memory back when we delete it
mem_change(rm(x))
```

Even operations that don't do anything use up a little memory. This is because R is tracking the history of everything you do. You can ignore anything smaller than a couple kB.

```{r}
mem_change(NULL)
mem_change(NULL)
```

In some languages, you have to explicitly delete unused objects for their memory to be returned. R uses an alternative approach: garbage collection (or GC for short). GC automatically releases memory when an object is no longer used. It does this by tracking how many names point to each object, and when there are no names pointing to an object, it deletes that object. \index{garbage collection}

```{r, echo = FALSE}
rm(y)
```

```{r}
# Create a big object
mem_change(x <- 1:1e6)
# Also point to 1:1e6 from y
mem_change(y <- x)
# Remove x, no memory freed because y is still pointing to it
mem_change(rm(x))
# Now nothing points to it and the memory can be freed
mem_change(rm(y))
```

Despite what you might have read elsewhere, there's never any need to call `gc()` yourself. R will automatically run garbage collection whenever it needs more space; if you want to see when that is, call `gcinfo(TRUE)`. The only reason you _might_ want to call `gc()` is to ask R to return memory to the operating system. However, even that might not have any effect: older versions of Windows had no way for a program to return memory to the OS. \indexc{gc()}

GC takes care of releasing objects that are no longer used. However, you do need to be aware of possible memory leaks. A memory leak occurs when you keep pointing to an object without realising it. In R, the two main causes of memory leaks are formulas and closures because they both capture the enclosing environment. The following code illustrates the problem. In `f1()`, `1:1e6` is only referenced inside the function, so when the function completes the memory is returned and the net memory change is 0. `f2()` and `f3()` both return objects that capture environments, so that `x` is not freed when the function completes. \index{memory!leaks}

```{r}
f1 <- function() {
  x <- 1:1e6
  10
}
mem_change(x <- f1())
object_size(x)

f2 <- function() {
  x <- 1:1e6
  a ~ b
}
mem_change(y <- f2())
object_size(y)

f3 <- function() {
  x <- 1:1e6
  function() 10
}
mem_change(z <- f3())
object_size(z)
```

```{r, echo = FALSE}
rm(y, z)
```


## Modification in place {#modification}

What happens to `x` in the following code? \index{copy-on-modify!exceptions} \index{avoiding copies}

```{r}
x <- 1:10
x[5] <- 10L
x
```

There are two possibilities:

1. R modifies `x` in place.

2. R makes a copy of `x` to a new location, modifies the copy, and then uses 
   the name `x` to point to the new location.

It turns out that R can do either depending on the circumstances. In the example above, it will modify in place. But if another variable also points to `x`, then R will copy it to a new location. To explore what's going on in greater detail, we use two tools from the pryr package. Given the name of a variable, `address()` will tell us the variable's location in memory and `refs()` will tell us how many names point to that location. \indexc{address()} \indexc{refs()}

```{r, eval = FALSE}
library(pryr)
x <- 1:10
c(address(x), refs(x))
# [1] "0x103100060" "1"

y <- x
c(address(y), refs(y))
# [1] "0x103100060" "2"
```

(Note that if you're using RStudio, `refs()` will always return 2: the environment browser makes a reference to every object you create on the command line.)

`refs()` is only an estimate. It can only distinguish between one and more than one reference (future versions of R might do better). This means that `refs()` returns 2 in both of the following cases: \index{reference counting}

```{r}
x <- 1:5
y <- x
rm(y)
# Should really be 1, because we've deleted y
refs(x)

x <- 1:5
y <- x
z <- x
# Should really be 3
refs(x)
```

When `refs(x)` is 1, modification will occur in place. When `refs(x)` is 2, R will make a copy (this ensures that other pointers to the object remain unaffected). Note that in the following example, `y` keeps pointing to the same location while `x` changes.

```{r}
x <- 1:10
y <- x
c(address(x), address(y))

x[5] <- 6L
c(address(x), address(y))
```

Another useful function is `tracemem()`. It prints a message every time the traced object is copied: \indexc{tracemem()}

```{r tracemem, eval = FALSE}
x <- 1:10
# Prints the current memory location of the object
tracemem(x)
# [1] "<0x7feeaaa1c6b8>"

x[5] <- 6L

y <- x
# Prints where it has moved from and to
x[5] <- 6L
# tracemem[0x7feeaaa1c6b8 -> 0x7feeaaa1c768]:
```

For interactive use, `tracemem()` is slightly more useful than `refs()`, but because it just prints a message, it's harder to program with. I don't use it in this book because it interacts poorly with [knitr](http://yihui.name/knitr/), the tool I use to interleave text and code.

Non-primitive functions that touch the object always increment the ref count. Primitive functions usually don't. (The reasons are a little complicated, but see the R-devel thread [confused about NAMED](http://r.789695.n4.nabble.com/Confused-about-NAMED-td4103326.html).) \index{primitive functions}

```{r}
# Touching the object forces an increment
f <- function(x) x
{x <- 1:10; f(x); refs(x)}

# Sum is primitive, so no increment
{x <- 1:10; sum(x); refs(x)}

# f() and g() never evaluate x, so refs don't increment
f <- function(x) 10
g <- function(x) substitute(x)

{x <- 1:10; f(x); refs(x)}
{x <- 1:10; g(x); refs(x)}
```

Generally, provided that the object is not referred to elsewhere, any primitive replacement function will modify in place. This includes `[[<-`, `[<-`, `@<-`, `$<-`, `attr<-`, `attributes<-`, `class<-`, `dim<-`, `dimnames<-`, `names<-`, and `levels<-`. To be precise, all non-primitive functions increment refs, but a primitive function may be written in such a way that it doesn't. The rules are sufficiently complicated that there's little point in trying to memorise them. Instead, you should approach the problem practically by using `refs()` and `address()` to figure out when objects are being copied. \index{subsetting|subassignment}

While determining that copies are being made is not hard, preventing such behaviour is. If you find yourself resorting to exotic tricks to avoid copies, it may be time to rewrite your function in C++, as described in [Rcpp](#rcpp).

### Loops

For loops in R have a reputation for being slow. Often that slowness is because you're modifying a copy instead of modifying in place. Consider the following code. It subtracts the median from each column of a large data frame: \index{loops!avoiding copies}

```{r, cache = TRUE}
x <- data.frame(matrix(runif(100 * 1e4), ncol = 100))
medians <- vapply(x, median, numeric(1))

for(i in seq_along(medians)) {
  x[, i] <- x[, i] - medians[i]
}
```

You may be surprised to realise that every iteration of the loop copies the data frame. We can see that more clearly by using `address()` and `refs()` for a small sample of the loop:

```{r}
for(i in 1:5) {
  x[, i] <- x[, i] - medians[i]
  print(c(address(x), refs(x)))
}
```

For each iteration, `x` is moved to a new location so `refs(x)` is always 2. This occurs because `[<-.data.frame` is not a primitive function, so it always increments the refs. We can make the function substantially more efficient by using a list instead of a data frame. Modifying a list uses primitive functions, so the refs are not incremented and all modifications occur in place:

```{r}
y <- as.list(x)

for(i in 1:5) {
  y[[i]] <- y[[i]] - medians[i]
  print(c(address(y), refs(y)))
}
```

This behaviour was substantially more problematic prior to R 3.1.0, because every copy of the data frame was a deep copy. This made the motivating example take around 5 s, compared to 0.01 s today.

### Exercises

1.  The code below makes one duplication. Where does it occur and why? 
    (Hint: look at `refs(y)`.)

    ```{r}
    y <- as.list(x)
    for(i in seq_along(medians)) {
      y[[i]] <- y[[i]] - medians[i]
    }
    ```

1.  The implementation of `as.data.frame()` in the previous section has one 
    big downside. What is it and how could you avoid it?


[long-vectors]: http://cran.r-project.org/doc/manuals/R-ints.html#Long-vectors
