# OO field guide {#oo}

This chapter is a field guide for recognising and working with R's objects in the wild. R has three object oriented systems (plus the base types), so it can be a bit intimidating. The goal of this guide is not to make you an expert in all four systems, but to help you identify which system you're working with and to help you use it effectively. \index{object-oriented programming}

Central to any object-oriented system are the concepts of class and method. A __class__ defines the behaviour of __objects__ by describing their attributes and their relationship to other classes. The class is also used when selecting __methods__, functions that behave differently depending on the class of their input. Classes are usually organised in a hierarchy: if a method does not exist for a child, then the parent's method is used instead; the child __inherits__ behaviour from the parent.

R's three OO systems differ in how classes and methods are defined:

* __S3__ implements a style of OO programming called generic-function OO. 
  This is different from most programming languages, like Java, C++, and C#,
  which implement message-passing OO. With message-passing, messages (methods)
  are sent to objects and the object determines which function to call.
  Typically, this object has a special appearance in the method call, usually
  appearing before the name of the method/message: e.g., 
  `canvas.drawRect("blue")`. S3 is different. While computations are still
  carried out via methods, a special type of function called a 
  __generic function__ decides which method to call, e.g., 
  `drawRect(canvas, "blue")`. S3 is a very casual system. It has no 
  formal definition of classes.

* __S4__ works similarly to S3, but is more formal. There are two major
  differences to S3. S4 has formal class definitions, which describe the
  representation and inheritance for each class, and has special helper
  functions for defining generics and methods. S4 also has multiple dispatch,
  which means that generic functions can pick methods based on the class of 
  any number of arguments, not just one.

* __Reference classes__, called RC for short, are quite different from S3 
  and S4. RC implements message-passing OO, so methods belong to classes, 
  not functions. `$` is used to separate objects and methods, so method calls
  look like `canvas$drawRect("blue")`. RC objects are also mutable: they don't
  use R's usual copy-on-modify semantics, but are modified in place. This 
  makes them harder to reason about, but allows them to solve problems that 
  are difficult to solve with S3 or S4.

There's also one other system that's not quite OO, but it's important to mention here:

* __base types__, the internal C-level types that underlie the other OO 
  systems. Base types are mostly manipulated using C code, but they're 
  important to know about because they provide the building blocks for the 
  other OO systems.

The following chapters describe each system in turn, starting with base types. You'll learn how to recognise the OO system that an object belongs to, how method dispatch works, and how to create new objects, classes, generics, and methods for that system. The chapter concludes with a few remarks on when to use each system.

##### Prerequisites

You'll need the pryr package, `install.packages("pryr")`, to access useful functions for examining OO properties.

## Why OO?

The primary use of OO programming in R is for print, summary and plot methods. These methods allow us to have one generic function, e.g. `print()`, that displays the object differently depending on its type: printing a linear model is very different to printing a data frame.

## Why generic functions?

Compare pipe with method chaining.

This is a different school of though to most popular programming languages, but is a good fit to the problem of data analysis. Knowing this fact probably won't help you much in your day-to-day programming, but it will avoid some fundamental confusing if you're coming from another OO programmming language. \index{functions!generics|see{generics}} \index{S3!generics} \index{generics!S3}

```{r, eval = FALSE}
class.method(arg1, arg2)
class$method(arg1, arg2)
method(arg1, arg2, arg3)
```

(In fact this message is so powerful that I've talked to programmers who moved to R from javascript and it took them a while to figure out that they're not calling the `frame` method of the `data` object.)

## S4 {#s4}

S4 works in a similar way to S3, but it adds formality and rigour. Methods still belong to functions, not classes, but: \index{objects!S4|see{S4}} \index{S4}

* Classes have formal definitions which describe their fields and
  inheritance structures (parent classes).

* Method dispatch can be based on multiple arguments to a generic function,
  not just one.

* There is a special operator, `@`, for extracting slots (aka fields)
  from an S4 object.

All S4 related code is stored in the `methods` package. This package is always available when you're running R interactively, but may not be available when running R in batch mode. For this reason, it's a good idea to include an explicit `library(methods)` whenever you're using S4.

S4 is a rich and complex system. There's no way to explain it fully in a few pages. Here I'll focus on the key ideas underlying S4 so you can use existing S4 objects effectively. To learn more, some good references are:

* [S4 system development in Bioconductor](http://www.bioconductor.org/help/course-materials/2010/AdvancedR/S4InBioconductor.pdf)

* John Chambers' [_Software for Data Analysis_](http://amzn.com/0387759352?tag=devtools-20)

* [Martin Morgan's answers to S4 questions on stackoverflow](http://stackoverflow.com/search?tab=votes&q=user%3a547331%20%5bs4%5d%20is%3aanswe)

### Recognising objects, generic functions, and methods

Recognising S4 objects, generics, and methods is easy. You can identify an S4 object because `str()` describes it as a "formal" class, `isS4()` returns `TRUE`, and `pryr::otype()` returns "S4". S4 generics and methods are also easy to identify because they are S4 objects with well defined classes.

There aren't any S4 classes in the commonly used base packages (stats, graphics, utils, datasets, and base), so we'll start by creating an S4 object from the built-in stats4 package, which provides some S4 classes and methods associated with maximum likelihood estimation:

```{r}
library(stats4)
library(pryr)

# From example(mle)
y <- c(26, 17, 13, 12, 20, 5, 9, 8, 5, 4, 8)
nLL <- function(lambda) - sum(dpois(y, lambda, log = TRUE))
fit <- mle(nLL, start = list(lambda = 5), nobs = length(y))

# An S4 object
isS4(fit)
otype(fit)

# An S4 generic
isS4(nobs)
ftype(nobs)

# Retrieve an S4 method, described later
mle_nobs <- method_from_call(nobs(fit))
isS4(mle_nobs)
ftype(mle_nobs)
```

Use `is()` with one argument to list all classes that an object inherits from. Use `is()` with two arguments to test if an object inherits from a specific class.

```{r}
is(fit)
is(fit, "mle")
```

You can get a list of all S4 generics with `getGenerics()`, and a list of all S4 classes with `getClasses()`. This list includes shim classes for S3 classes and base types. You can list all S4 methods with `showMethods()`, optionally restricting selection either by `generic` or by `class` (or both). It's also a good idea to supply `where = search()` to restrict the search to methods available in the global environment.

### Defining classes and creating objects

In S3, you can turn any object into an object of a particular class just by setting the class attribute. S4 is much stricter: you must define the representation of a class with `setClass()`, and create a new object with `new()`. You can find the documentation for a class with a special syntax: `class?className`, e.g., `class?mle`. \index{S4!classes} \index{classes!S4}

An S4 class has three key properties:

* A __name__: an alpha-numeric class identifier. By convention, S4 class names
  use UpperCamelCase.

* A named list of __slots__ (fields), which defines slot names and
  permitted classes. For example, a person class might be represented by a
  character name and a numeric age: `list(name = "character", age = "numeric")`.
  \index{slots}

* A string giving the class it inherits from, or, in S4 terminology,
  that it __contains__. You can provide multiple classes for multiple
  inheritance, but this is an advanced technique which adds much
  complexity.

  In `slots` and `contains` you can use S4 classes, S3 classes registered
  with `setOldClass()`, or the implicit class of a base type. In `slots`
  you can also use the special class `ANY` which does not restrict the input.

S4 classes have other optional properties like a `validity` method that tests if an object is valid, and a `prototype` object that defines default slot values. See `?setClass` for more details.

The following example creates a Person class with fields name and age, and an Employee class that inherits from Person. The Employee class inherits the slots and methods from the Person, and adds an additional slot, boss. To create objects we call `new()` with the name of the class, and name-value pairs of slot values. \indexc{setClass()} \indexc{new()}

```{r}
setClass("Person",
  slots = list(name = "character", age = "numeric"))
setClass("Employee",
  slots = list(boss = "Person"),
  contains = "Person")

alice <- new("Person", name = "Alice", age = 40)
john <- new("Employee", name = "John", age = 20, boss = alice)
```

Most S4 classes also come with a constructor function with the same name as the class: if that exists, use it instead of calling `new()` directly.

To access slots of an S4 object use `@` or `slot()`: \index{subsetting!S4} \index{S4|subsetting}

```{r}
alice@age
slot(john, "boss")
```

(`@` is equivalent to `$`, and `slot()` to `[[`.)

If an S4 object contains (inherits from) an S3 class or a base type, it will have a special `.Data` slot which contains the underlying base type or S3 object: \indexc{.Data}

```{r}
setClass("RangedNumeric",
  contains = "numeric",
  slots = list(min = "numeric", max = "numeric"))
rn <- new("RangedNumeric", 1:10, min = 1, max = 10)
rn@min
rn@.Data
```

Since R is an interactive programming language, it's possible to create new classes or redefine existing classes at any time. This can be a problem when you're interactively experimenting with S4. If you modify a class, make sure you also recreate any objects of that class, otherwise you'll end up with invalid objects.

### Creating new methods and generics

S4 provides special functions for creating new generics and methods. `setGeneric()` creates a new generic or converts an existing function into a generic. `setMethod()` takes the name of the generic, the classes the method should be associated with, and a function that implements the method. For example, we could take `union()`, which usually just works on vectors, and make it work with data frames: \index{S4!generics} \index{S4!methods} \index{generics!S4} \index{methods!S4}

```{r}
setGeneric("union")
setMethod("union",
  c(x = "data.frame", y = "data.frame"),
  function(x, y) {
    unique(rbind(x, y))
  }
)
```

If you create a new generic from scratch, you need to supply a function that calls `standardGeneric()`:

```{r}
setGeneric("myGeneric", function(x) {
  standardGeneric("myGeneric")
})
```

`standardGeneric()` is the S4 equivalent to `UseMethod()`.

### Method dispatch

If an S4 generic dispatches on a single class with a single parent, then S4 method dispatch is the same as S3 dispatch. The main difference is how you set up default values: S4 uses the special class `ANY` to match any class and "missing" to match a missing argument. Like S3, S4 also has group generics, documented in `?S4groupGeneric`, and a way to call the "parent" method, `callNextMethod()`. \index{S4!method dispatch rules}

Method dispatch becomes considerably more complicated if you dispatch on multiple arguments, or if your classes use multiple inheritance. The rules are described in `?Methods`, but they are complicated and it's difficult to predict which method will be called. For this reason, I strongly recommend avoiding multiple inheritance and multiple dispatch unless absolutely necessary.

Finally, there are two methods that find which method gets called given the specification of a generic call:

```{r, eval = FALSE}
# From methods: takes generic name and class names
selectMethod("nobs", list("mle"))

# From pryr: takes an unevaluated function call
method_from_call(nobs(fit))
```

### Exercises

1.  Which S4 generic has the most methods defined for it? Which S4 class 
    has the most methods associated with it?

1.  What happens if you define a new S4 class that doesn't "contain" an 
    existing class?  (Hint: read about virtual classes in `?setClass`.)

1.  What happens if you pass an S4 object to an S3 generic? What happens 
    if you pass an S3 object to an S4 generic? (Hint: read `?setOldClass` 
    for the second case.)

## Picking a system {#picking-a-system}

Three OO systems is a lot for one language, but for most R programming, S3 suffices. In R you usually create fairly simple objects and methods for pre-existing generic functions like `print()`, `summary()`, and `plot()`. S3 is well suited to this task, and the majority of OO code that I have written in R is S3. S3 is a little quirky, but it gets the job done with a minimum of code. \index{objects!which system?}

```{r, eval = FALSE, echo = FALSE}
packageVersion("Matrix")

library(Matrix)
gs <- getGenerics("package:Matrix")
sum(gs@package == "Matrix")

length(getClasses("package:Matrix", FALSE))
```

If you are creating more complicated systems of interrelated objects, S4 may be more appropriate. A good example is the `Matrix` package by Douglas Bates and Martin Maechler. It is designed to efficiently store and compute with many different types of sparse matrices. As of version 1.1.3, it defines 102 classes and 20 generic functions. The package is well written and well commented, and the accompanying vignette (`vignette("Intro2Matrix", package = "Matrix")`) gives a good overview of the structure of the package. S4 is also used extensively by Bioconductor packages, which need to model complicated interrelationships between biological objects. Bioconductor provides many [good resources](https://www.google.com/search?q=bioconductor+s4) for learning S4. If you've mastered S3, S4 is relatively easy to pick up; the ideas are all the same, it is just more formal, more strict, and more verbose.

If you've programmed in a mainstream OO language, RC will seem very natural. But because they can introduce side effects through mutable state, they are harder to understand. For example, when you call `f(a, b)` in R you can usually assume that `a` and `b` will not be modified. But if `a` and `b` are RC objects, they might be modified in the place. Generally, when using RC objects you want to minimise side effects as much as possible, and use them only where mutable states are absolutely required. The majority of functions should still be "functional", and free of side effects. This makes code easier to reason about and easier for other R programmers to understand.
