# (PART) Object oriented programming {-}
\index{object-oriented programming}

```{r setup, include = FALSE}
source("common.R")
library(sloop)
```

# Introduction {#oo .unnumbered}

In the following five chapters you'll learn about __object oriented programming__ (OOP) in R. OOP in R is a little more challenging than in other languages because:

* There are multiple OOP systems to choose from. In this book, I'll focus 
  on the three that I believe are most important: S3, S4, and R6.

* S3 and S4 implement a OOP paradigm different to most languages popular today. 
  While the underlying ideas are the same, their expression is rather different. 
  This means that if you're familiar with OOP in another language, you need to 
  understand that R's approach is different (not better or worse, just 
  different).

* There is disagreement about the relative importance of the three systems. 
  I think S3 is most important, followed by R6, then S4. Others believe that 
  S4 is most important, followed by RC (a base alternative to R6), and S3 
  should be avoided. This means that different R communities use different 
  systems.

* There are no books focussed on the effective use of OOP in R. This book
  only helps a little because it's focussed on the underlying theory and 
  implementation, and provides relatively little advice on how to use OOP
  most effectively in R.

In R generally, functional programming is much more important than object oriented programming, because generally you solve complex problems decomposing them into functions, not objects. Nevertheless, there are important reasons to learn about each of the three systems:

* S3 allows your functions to return richer results with user-friendly display
  and programmer-friendly internals. S3 is used throughout base R, so it's 
  important to master it if you want to extend base R functions to work with 
  new types of input.

* S4 is a rigorous system that forces you to think carefully about the design.
  It's particuarly well-suited for building large systems that evolve over time 
  and will receive contributions from many programmers. This is why the
  Bioconductor project choose to build on top S4.
  
* R6 gives you a standardised way to escape R's copy-on-modify semantics. 
  This is particularly important if you want to model real-world objects 
  that can be simultaneously changed through multiple interfaces. Today,
  a common need for R6 is to model data in a web API, that may be manipulated
  outside of R.
  
This chapter will give you a rough lay of the land along with a field guide to help you identify OOP systems in the wild. The following four chapters (Base types, S3, S4, and R6) will dive into the details, starting with R's base types. These are not technically an OOP system, but they're important to understand because they're the fundamental building block of the true OOP systems.

## OOP Systems {-}

We'll begin with an brain dump of vocabulary and terminology. Don't worry if it doesn't stick as we'll come back to these ideas multiple times in the subsequent chapters.

The main reason to use OOP is is __polymorphism__ (literally: many shapes). Polymorphism means that a developer can consider a function's interface separately from its implementation, making it possible to use the same function for many types of input. This is closely related to the idea of __encapsulation__: the user doesn't need to worry about details of an object because they are encapsulated behind a standard interface. Polymorphism is what allows `summary()` to deliver different types of summary for numeric and factor variables:

```{r}
diamonds <- ggplot2::diamonds

summary(diamonds$carat)

summary(diamonds$cut)
```

You could imagine `summary()` implementing this with a series of if-else statements, but that would mean only the original author could add new implementation. An OOP system makes it possible for any developer to extend the implementation for new types of input. 

So far I've informally talked about the "type" of a "thing". OOP systems use a common vocabularly: we talk about the __class__ of an __object__. The class defines the __fields__, the data that every instance of that class possesses, and how it is related to other classes. A __method__ defines how object behaves; a method is like a function that behaves differently depending on the class of its input. In other words, a class defines what an object _is_ and methods describe what that object can _do_. Classes are usually organised in a hierarchy: if a method does not exist for a child, then the parent's method is used instead, and the child is said to __inherit__ behaviour from the parent. 

There are two main paradigms of object-oriented programming that differ in how methods are related to classes. In this book, we'll borrow the notation of _Extending R_ [@extending-R] and call the two paradigms encapsulated and functional:

*   In __encapsulated__ OOP, methods belong to objects or classes, and method 
    calls typically look like `object.method(arg1, arg2)`. This is called 
    encapsulated because the object encapsulates both data (through fields) and 
    behaviour (with methods), and is most common paradigm in today's popular 
    languages.
    
*   In __functional__ OOP, methods belong to __generic__ functions, and method 
    calls look like ordinary function calls: `generic(object, arg2, arg3)`. 
    This is called functional because it looks like a regular function call
    from the outside, and internally the components are all regular functions.

## OOP in R {-}

Base R provides three OOP systems: S3, S4, and reference classes (RC):

*   __S3__ is R's first OOP system, and is described in _Statistical Models 
    in S_ [@white-book]. S3 is an informal implementation of the functional 
    style which rellies on common conventions rather than ironclad guarantees. 
    This makes it easy to get started with, and means it's a low cost way of 
    solving many simple problems.

*   __S4__ is similar to S3, but much more formal, and was introduced in 
    _Programming with Data_ [@programming-with-data]. It requires more upfront 
    work than S3 and in return provides greater rigor and encapsulation. S4 is 
    implemented in the __methods__ package, which is attached by 
    default[^Rscript]. The only base package that uses S4 is the little known 
    stats4 package.
    
    (You might wonder if S1 and S2 exist. They don't: S3 and S4 were named 
    according to the versions of S that they accompanied. The first two 
    versions of S didn't have any OOP framework.)

*   __RC__ implements encapsulated OO. RC objects are a special type of S4 
    objects that are also mutable. Instead of using R's usual copy-on-modify 
    semantics, they are modified in place. This makes them harder to reason 
    about, but allows them to solve problems that are difficult to solve with 
    the function OOP provided by S3 and S4.

[^Rscript]: Unless you are running `Rscript`, in which case you'll need to load explicitly with `library(methods)`. This is a historical inconsistency introduced because the methods package used to take a long time to load and `Rscript` is optimised for fast command line usage. 

There are a number of other OOP systems provided by packages. Three of the most popular are:

*   __R6__ [@R6] implements encapsulated OOP like RC, but resolves some 
    important issues. In this book, you'll learn about R6 instead of RC, for
    reasons described in Chapter \@ref(R6).
    
*   __R.oo__ [@R.oo] provides some formalism on top of S3, and makes it possible to
    have mutable S3 objects.

*   __proto__ [@proto] implements another style of OOP based on the idea of 
    __prototypes__, which blur the distinctions between classes and instances 
    of classes (objects). I was briefly enamored with prototype based 
    programming [@mutatr] and used in ggplot2, but now think it's better to
    stick to the standard forms.

Apart from R6, which is widely used, the OOP systems implemented in external packages are primarily of theoretical interest. They will help you understand the spectrum of OOP better, and can make it easier to solve certain classes of problems, but they come with a big drawback: few R users know and understand them, so it is hard for others to read and contribute to your code. For that reason, I recommend that in your own code you stick with the most used: S3, S4, and R6.

## Sloop {-}

Before we go on to discuss base types, S3, S4, and R6 in more detail I want to introduce the sloop package:

```{r}
# install_github("r-lib/sloop")
library(sloop)
```

The sloop package (think "sail the seas of OOP") provides a number of helpers that fill in missing pieces in base R. The first helper to learn about is `sloop::otype()`. It makes it easy to figure out the OOP system used by a wild-caught object: 

```{r}
otype(1:10)

otype(mtcars)

mle_obj <- stats4::mle(function(x = 1) (x - 2) ^ 2)
otype(mle_obj)
```

Use this function if you want to tactically read just the chapter that applies to an object that you're working with. 
