# (PART) Object oriented programming {-}

# Introduction {#oo}

In this and following five chapters you will learn about __object oriented programming__ (OOP) in R. OOP in R is a little more challenging than in other languages, because there are multiple systems that have evolved over time. In this book, we'll focus on the three that are most important in my opinion: S3, S4, and R6. \index{object-oriented programming}

For day-to-day use of R, FP is much more important than OOP. However, there are three main reasons to learn OOP:

* Learning a little S3 allows your functions to return more complex objects 
  that have a user friendly display and programmer friendly internals.

* Investing in S4 can be helpful for building up large systems that evolve
  over many years and are contributed to by many programmers.
  
* Mastering R6 gives you a standard way to escape R's copy-on-modify semantics
  when needed.
  
This chapter will give you a rough lay of the land, and a field guide to help you identify OOP systems in the wild. The following four chapters (Base types, S3, S4, and R6) will dive into the details, starting with R's base types. These not technically an OOP system, but it's important to talk about them because they're the fundamental building block of the other OOP systems.

## OOP Systems

Central to any OOP system are the concepts of class and method. A __class__ defines the behaviour of __objects__ by describing their attributes and their relationship to other classes. The class is also used when selecting __methods__, functions that behave differently depending on the class of their input. Classes are usually organised in a hierarchy: if a method does not exist for a child, then the parent's method is used instead; the child __inherits__ behaviour from the parent. Inheritance allows you to reduce duplication by allowing more specialised object inherit behaviour from their more general kin.

Base R provides three OOP systems: S3, S4, and references classes (RC):

*   __S3__ implements a style of OOP called functional OOP. This is different 
    from most of today's popular programming languages, like Java, C++, and C#.
    The key part of encapsulated OOP is the __generic function__, a function that
    decides which method to call based on the class of an argument. S3 is 
    most commonly used OOP system in base R, and powers commonly used functions
    like `print()` and `mean()`.
  
    S3 is a very informal system. It provides no ironclad guarantees but 
    instead relies on a set of conventions. This makes it easy to get started
    with, and a low cost way of solving many simple problems.

*   __S4__ is similar to S3, but much more formal. It requires more upfront
    work and in return provides greater consistency. S4 is implemented
    in the __methods__ package, which is attached by default. The only
    package in base R to make use of S4 is stats4.
    
    In both S3 and S4 methods belong to generics, not classes. This makes them
    feel quite different to the encapsulated style of OOP found in most 
    popular languages.
    
    (You might wonder if S1 and S2 exist. They don't: S3 and S4 were named 
    according to the versions of S that they accompanied.)

*   __RC__ implements encapsulated OO, where methods belong to classes, 
    not generics. This makes method calls look quite different: 
    `canvas$drawRect("blue")`. RC objects are also mutable: they don't
    use R's usual copy-on-modify semantics, but are modified in place. This 
    makes them harder to reason about, but allows them to solve problems that 
    are difficult to solve with S3 or S4.

There are a number other OOP systems provided by packages. Two of the most popular are:

*   __R6__ implements encapsulated OOP like RC, but resolves some important 
    issues. You'll learn R6 instead of RC in this book. More on why later.
    
*   __proto__ implements another style of OOP, called prototype based. It
    blurs the distinctions between classes and instances of classes (objects).
    
There are a handful of other less popular packages. These are primarily of academic interest (and can be mind expanding to learn about). They can make it easier to solve certain classes of problems but come with a big drawback: few R users understand them.

## Field guide 

Given an existing object, how do you figure out which OOP system it comes from? It's possible, but fiddle, to work your way through the base functions `is.object()`, `isS4()`, and `inherits()`. An easier way is `sloop::otype()`. The __sloop__ package (sail the six seas of OOP in R) provides a number of small helpers to aid with OOP. 

```{r setup}
# install_github("hadley/sloop")
library(sloop)
```

`sloop::otype()` will tell you what you're dealing with:

```{r}
otype(1:10)

otype(mtcars)

mle_obj <- stats4::mle(function(x = 1) (x - 2) ^ 2)
otype(mle_obj)
```

