# (PART) Object oriented programming {-}

# OO field guide {#oo}

## What is OOP?

\index{object-oriented programming}

Central to any object-oriented system are the concepts of class and method. A __class__ defines the behaviour of __objects__ by describing their attributes and their relationship to other classes. The class is also used when selecting __methods__, functions that behave differently depending on the class of their input. Classes are usually organised in a hierarchy: if a method does not exist for a child, then the parent's method is used instead; the child __inherits__ behaviour from the parent.

## OOP Systems

Before we can discuss OOP systems, we need to talk about the __base types__. These are not an OO system in the usual sence because they are built into R and can only be modified by R-core. However, they're important to understand, because they're the building blocks upon which user OOP is build.

There are three OOP systems provided by base R:

* __S3__ implements a style of OO programming called functional OOP. 
  This is different from most programming languages, like Java, C++, and C#,
  which implement message-passing OO. With message-passing, messages (methods)
  are sent to objects and the object determines which function to call.
  Typically, this object has a special appearance in the method call, usually
  appearing before the name of the method/message: e.g., 
  `canvas.drawRect("blue")`. S3 is different. While computations are still
  carried out via methods, a special type of function called a 
  __generic function__ decides which method to call, e.g., 
  `drawRect(canvas, "blue")`. S3 is a very casual system. It has no 
  formal definition of classes.

* __S4__ works similarly to S3, but is more formal. There are two major
  differences to S3. S4 has formal class definitions, which describe the
  representation and inheritance for each class, and has special helper
  functions for defining generics and methods. S4 also has multiple dispatch,
  which means that generic functions can pick methods based on the class of 
  any number of arguments, not just one.

(You might wonder if S1 and S2 exist. They don't: S3 and S4 were named according to the version of S that they accompanied.)

* __Reference classes__, called RC for short, are quite different from S3 
  and S4. RC implements message-passing OO, so methods belong to classes, 
  not functions. `$` is used to separate objects and methods, so method calls
  look like `canvas$drawRect("blue")`. RC objects are also mutable: they don't
  use R's usual copy-on-modify semantics, but are modified in place. This 
  makes them harder to reason about, but allows them to solve problems that 
  are difficult to solve with S3 or S4.

Apart from teaching you how to recognise them, this book is not going to cover RC in any detail. Instead, you'll learn about the R6 package. R6 provides the same style of OOP as RC, but resolves some important issues (more on that in the chapter).

There are a handful of other OOP systems provided by other packages. Probably the most popular is the proto package, which provides another OOP system known of prototype based programming. Other OOP systems can make it easier to solve certain classes of problems but come with a big drawback: few R users know about them.

## Field guide 

To provide a few missing helper functions for working with OOP classes, generics, and methods, we're going to use the sloop package:

```{r setup}
# install_github("hadley/sloop")
library(sloop)
```

If you're not sure what you're dealing with, and you want to read the next chapters tactically in order to learn just what you need right now, `sloop::otype()` will help you figure out what you're dealing with:

```{r}
otype(1:10)

otype(mtcars)

mle_obj <- stats4::mle(function(x = 1) (x - 2) ^ 2)
otype(mle_obj)
```

If you get an object of type "RC", you'll have to read the chapter on R6. 
