# (PART) Object oriented programming {-}
\index{object-oriented programming}

```{r setup, include = FALSE}
source("common.R")
library(sloop)
```

# Introduction {#oo .unnumbered}

In the following five chapters you'll learn about __object oriented programming__ (OOP) in R. OOP is a little more challenging in R than in some other languages because:

* There are multiple OOP systems to choose from. In this book, I'll focus 
  on the three that I believe are most important: __S3__,  __R6__, and __S4__. 
  S3 and S4 are provided by base R. R6 is provided by the R6 package, and is 
  similar to the "Reference Classes", or __RC__ for short, from base R.

* S3 and S4 use "generic function OOP" which is rather different from the 
  "encapsulated OOP" used by most languages popular today[^julia]. We'll come 
  back to preisely what those terms mean shortly, but basically, while the 
  underlying ideas of OOP are the same, their expression is rather different. 
  This means that you can't immediately transfer your OOP skills from other
  languages.

* There is disagreement about the relative importance of the three systems. 
  I think S3 is most important, followed by R6, then S4. Others believe that 
  S4 is most important, followed by RC, and that S3 should be avoided. This 
  means that different R communities use different systems.

[^julia]: The exception is Julia, which uses a very similar style of OOP (if much more rigorous and performant) to R.

Generally in R, functional programming is much more important than object oriented programming, because you typically solve complex problems by decomposing them into simple functions, not simple objects. Nevertheless, there are important reasons to learn each of the three systems:

* S3 allows your functions to return richer results with user-friendly display
  and programmer-friendly internals. S3 is used throughout base R, so it's 
  important to master if you want to extend base R functions to work with new 
  types of input.

* R6 gives you a standardised way to escape R's from copy-on-modify semantics. 
  This is particularly important if you want to model real-world objects 
  that can be simultaneously changed through multiple interfaces. Today,
  a common need for R6 is to model data that comes from a web API, and where
  changes may be instantiated from inside or outside of R.

* S4 is a rigorous system that forces you to think carefully about program 
  design. It's particuarly well-suited for building large systems that evolve 
  over time  and will receive contributions from many programmers. This is
  why it is used by the Bioconductor project, so if you want to contribute to
  Bioconductor, you'll need to know something about S4.

The goal of this brief introdutory chapter is to give you some important vocabulary and some tools to identify OOP systems in the wild. The following four chapters (Base types, S3, R6, and S4) then dive into the details of R's OOP systems. 

The following chapters need a caveat because they focus on the mechanics of OOP, not its effective use, and if you have never done object oriented programming before, you will gain few practical skills. You might wonder why I deliberately chose not to provide coverage that is more immediately useful. I have focussed on mechanics because they need to be well described somewhere (writing these chapters required a considerable amount of reading, exploration, and synthesis on my behalf), and using OOP effectively is sufficiently complex to require book-length treatment; there's simply not enough room in Advanced R to cover it in the depth that is required.

## OOP systems {-}

Different people use OOP terms in different ways, so this section provides a quick overview of important vocabulary. The explanations are necessarily compressed, but we will come back to these ideas multiple times

The main reason to use OOP is is __polymorphism__ (literally: many shapes). Polymorphism means that a developer can consider a function's interface separately from its implementation, making it possible to use the same function for many types of input. This is closely related to the idea of __encapsulation__: the user doesn't need to worry about details of an object because they are encapsulated behind a standard interface. Polymorphism is what allows `summary()` to produce different outputs for numeric and factor variables:

```{r}
diamonds <- ggplot2::diamonds

summary(diamonds$carat)

summary(diamonds$cut)
```

You could imagine `summary()` implementing this with a series of if-else statements, but that would mean only the original author could add new implementations. An OOP system makes it possible for any developer to extend the implementation for new types of input. 

So far I've informally talked about the "type" of a "thing". OOP systems use a common vocabularly: we talk about the __class__ of an __object__. The class defines the __fields__, which are the data that every instance of that class possesses.  A class also defines how it is related to other classes, so that (for example) a sparse matrix can be a specialized kind of matrix.

A __method__ defines how object behaves; it's like a function that behaves differently depending on the class of its input. In other words, a class defines what an object _is_ and methods describe what that object can _do_. Classes are usually organised in a hierarchy: if a method does not exist for a child, then the parent's method is used instead, and the child is said to __inherit__ behaviour from the parent. 

The two main paradigms of object-oriented programming differ in how methods are related to classes. In this book, we'll borrow the terminology of _Extending R_ [@extending-R] and call these paradigms encapsulated and functional:

*   In __encapsulated__ OOP, methods belong to objects or classes, and method 
    calls typically look like `object.method(arg1, arg2)`. This is called 
    encapsulated because the object encapsulates both data (through fields) and 
    behaviour (with methods), and is most common paradigm in today's popular 
    languages.
    
*   In __functional__ OOP, methods belong to __generic__ functions, and method 
    calls look like ordinary function calls: `generic(object, arg2, arg3)`. 
    This is called functional because it looks like a regular function call
    from the outside, and internally the components are all regular functions.

In either object system, it's the process of __method dispatch__ that goes from a high-level function call (whether it's `obj.meth()` or `meth(obj)`) to find a specific implementation.

## OOP in R {-}

Base R provides three OOP systems: S3, S4, and reference classes (RC):

*   __S3__ is R's first OOP system, and is described in _Statistical Models 
    in S_ [@white-book]. S3 is an informal implementation of the functional 
    style which relies on common conventions rather than ironclad guarantees. 
    This makes it easy to get started with, and makes it a low cost way of 
    solving many simple problems, but it isn't sturdy enough to handle all
    of the complexities of very large or very complex software designs.

*   __S4__ is similar to S3, but much more formal, and was introduced in 
    _Programming with Data_ [@programming-with-data]. It requires more upfront 
    work than S3, but in return provides greater rigor and encapsulation. S4 is 
    implemented in the __methods__ package, which is attached by 
    default[^Rscript]. The only base package that uses S4 is the little known 
    stats4 package.
    
    (You might wonder if S1 and S2 exist. They don't: S3 and S4 were named 
    according to the versions of S that they accompanied. The first two 
    versions of S didn't have any OOP framework.)

*   __RC__ implements encapsulated OO. RC objects are a special type of S4 
    objects that are also __mutable__, i.e., instead of using R's usual copy-on-modify 
    semantics, they can be modified in place. This makes them harder to reason 
    about, but allows them to solve problems that are difficult to solve in
    the functional OOP of S3 and S4.

[^Rscript]: Unless you are running `Rscript`, in which case you'll need to load explicitly with `library(methods)`. This is a historical inconsistency introduced because the methods package used to take a long time to load and `Rscript` is optimised for fast command line usage. 

A number of other OOP systems are provided by packages, including:

*   __R6__ [@R6] implements encapsulated OOP like RC, but resolves some 
    important issues. In this book, you'll learn about R6 instead of RC, for
    reasons described in Chapter \@ref(R6).
    
*   __R.oo__ [@R.oo] provides some formalism on top of S3, and makes it possible to
    have mutable S3 objects.

*   __proto__ [@proto] implements another style of OOP based on the idea of 
    __prototypes__, which blur the distinctions between classes and instances 
    of classes (objects). I was briefly enamored with prototype based 
    programming [@mutatr] and used in ggplot2, but now think it's better to
    stick to the standard forms.

Apart from R6, which is widely used, these systems are primarily of theoretical interest. They do have their strengths, but few R users know and understand them, so it is hard for others to read and contribute to your code. I therefore recommend that you stick with S3, S4, and R6.

## Sloop {-}

Before we go on I want to introduce the sloop package:

```{r}
# install_github("r-lib/sloop")
library(sloop)
```

The sloop package (think "sail the seas of OOP") provides a number of helpers that fill in missing pieces in base R. The first of these is `sloop::otype()`. It makes it easy to figure out the OOP system used by a wild-caught object: 

```{r}
otype(1:10)

otype(mtcars)

mle_obj <- stats4::mle(function(x = 1) (x - 2) ^ 2)
otype(mle_obj)
```

Use this function if you want to tactically read only the chapter that applies to a current problem.
