# RC

Reference classes (or RC for short) are the newest OO system in R. They were introduced in version 2.12. They are fundamentally different to S3 and S4 because: \index{RC} \index{reference classes|see{RC}} \index{objects!RC|see{RC}}

* RC methods belong to objects, not functions

* RC objects are mutable: the usual R copy-on-modify semantics do not apply

These properties make RC objects behave more like objects do in most other programming languages, e.g., Python, Ruby, Java, and C#. Reference classes are implemented using R code: they are a special S4 class that wraps around an environment.

### Defining classes and creating objects

Since there aren't any reference classes provided by the base R packages, we'll start by creating one. RC classes are best used for describing stateful objects, objects that change over time, so we'll create a simple class to model a bank account. \index{RC!classes} \index{classes!RC}

Creating a new RC class is similar to creating a new S4 class, but you use `setRefClass()` instead of `setClass()`. The first, and only required argument, is an alphanumeric __name__. While you can use `new()` to create new RC objects, it's good style to use the object returned by `setRefClass()` to generate new objects. (You can also do that with S4 classes, but it's less common.) \indexc{setRefClass()}

```{r}
Account <- setRefClass("Account")
Account$new()
```

`setRefClass()` also accepts a list of name-class pairs that define class __fields__ (equivalent to S4 slots). Additional named arguments passed to `new()` will set initial values of the fields. You can get and set field values with `$`: \index{fields}

```{r}
Account <- setRefClass("Account",
  fields = list(balance = "numeric"))

a <- Account$new(balance = 100)
a$balance
a$balance <- 200
a$balance
```

Instead of supplying a class name for the field, you can provide a single argument function which will act as an accessor method. This allows you to add custom behaviour when getting or setting a field. See `?setRefClass` for more details.

Note that RC objects are __mutable__, i.e., they have reference semantics, and are not copied-on-modify: \index{copy-on-modify!exceptions}

```{r}
b <- a
b$balance
a$balance <- 0
b$balance
```

For this reason, RC objects come with a `copy()` method that allow you to make a copy of the object:

```{r}
c <- a$copy()
c$balance
a$balance <- 100
c$balance
```

An object is not very useful without some behaviour defined by __methods__. RC methods are associated with a class and can modify its fields in place. In the following example, note that you access the value of fields with their name, and modify them with `<<-`. You'll learn more about `<<-` in [Environments](#binding). \index{RC!methods} \index{methods!RC} \indexc{<<-}

```{r}
Account <- setRefClass("Account",
  fields = list(balance = "numeric"),
  methods = list(
    withdraw = function(x) {
      balance <<- balance - x
    },
    deposit = function(x) {
      balance <<- balance + x
    }
  )
)
```

You call an RC method in the same way as you access a field:

```{r}
a <- Account$new(balance = 100)
a$deposit(100)
a$balance
```

The final important argument to `setRefClass()` is `contains`. This is the name of the parent RC class to inherit behaviour from. The following example creates a new type of bank account that returns an error preventing the balance from going below 0.

```{r, error = TRUE}
NoOverdraft <- setRefClass("NoOverdraft",
  contains = "Account",
  methods = list(
    withdraw = function(x) {
      if (balance < x) stop("Not enough money")
      balance <<- balance - x
    }
  )
)
accountJohn <- NoOverdraft$new(balance = 100)
accountJohn$deposit(50)
accountJohn$balance
accountJohn$withdraw(200)
```

All reference classes eventually inherit from `envRefClass`. It provides useful methods like `copy()` (shown above), `callSuper()` (to call the parent field), `field()` (to get the value of a field given its name), `export()` (equivalent to `as()`), and `show()` (overridden to control printing). See the inheritance section in `setRefClass()` for more details.


### Recognising objects and methods

You can recognise RC objects because they are S4 objects (`isS4(x)`) that inherit from "refClass" (`is(x, "refClass")`). `pryr::otype()` will return "RC".  RC methods are also S4 objects, with class `refMethodDef`.

### Method dispatch

Method dispatch is very simple in RC because methods are associated with classes, not functions. When you call `x$f()`, R will look for a method f in the class of x, then in its parent, then its parent's parent, and so on. From within a method, you can call the parent method directly with `callSuper(...)`. \index{RC!method dispatch rules}

### Exercises

1.  Use a field function to prevent the account balance from being directly
    manipulated. (Hint: create a "hidden" `.balance` field, and read the 
    help for the fields argument in `setRefClass()`.)

1.  I claimed that there aren't any RC classes in base R, but that was a 
    bit of a simplification. Use `getClasses()` and find which classes 
    `extend()` from `envRefClass`. What are the classes used for? (Hint: 
    recall how to look up the documentation for a class.)
