# R6

This chapter describes the R6 object system. Unlike S3 and S4, it provides encapsulated OO, which means that:

* R6 methods belong to objects, not generics.

* R6 objects are mutable: the usual copy-on-modify semantics do not apply.

These properties make R6 objects behave more like objects do in other popular programming languages, like python, ruby and java. Reference classes are implemented using R code: they are a special S3 class that wraps around an environment.

Unlike S3 and S4, R6 is not built in to R and you'll need to load a package to use it:

```{r setup}
library(R6)
```

R6 is very similar to a built-in OO system called __reference classes__, or RC for short. RC is the newest OO system in R and was introduced in R 2.12.0 in 2010. In this book, I'm going to cover R6 instead of RC for a few reasons:

* R6 is much simpler. Both R6 and RC are built on top of environments, but
  while R6 uses S3, RC uses S4. R6 is only ~500 lines of code (and ~1700 lines
  of tests!)

* RC relies on R's regular environment scoping rules to get and set field
  values. This has a couple of downsides: it makes cross-package inheritance 
  more complex, and it makes it harder to see which variables are fields.

* RC is quite slow. Generally, method dispatch will not be a bottleneck in your
  code, but R6 is much faster than RC. Switching from RC to R6 had a substantial
  impact on the performance of the shiny pacakge.

* It will only take a small amount of additional effort if you want to learn
  RC once you have mastered R6.

## Classes and methods

Because R6 implements encapsulated OO, you create a class and its method at the same time. Here's a simple example:

```{r}
Accumulator <- R6Class("Accumulator", public = list(
  sum = 0,
  add = function(x) {
    self$sum <- self$sum + x 
    invisible(self)
  })
)

x <- Accumulator$new() 
x$add(4) 
x$sum
```

Note that `add()` is called primarily for its side-effect of updating `self$sum`. In R6, it's often useful to invisibly return `self`. This makes it possible to chain together multiple method calls.

```{r}
x$add(10)$add(10)
x$sum
```

Usually when you modify an object in R, a copy is created and the original object is left unchanged. This is not so with R6.

```{r}
y1 <- Accumulator$new() 
y2 <- y1

y1$add(10)
y2$sum
```

### Initializer

Like S4, can override default behaviour of `new()` with `initialize()` method. Looks different , but same idea.

### print

## Inheritance

`super`

