# R6

This chapter describes the R6 object system. Unlike S3 and S4, it provides encapsulated OO, which means that:

* R6 methods belong to objects, not generics.

* R6 objects are mutable: the usual copy-on-modify semantics do not apply.

These properties make R6 objects behave more like objects do in other popular programming languages, like python, ruby and java. Reference classes are implemented using R code: they are a special S3 class that wraps around an environment.

Unlike S3 and S4, R6 is not built in to R and you'll need to load a package to use it:

```{r setup}
library(R6)
```

R6 is very similar to a built-in OO system called __reference classes__, or RC for short. RC is the newest OO system in R and was introduced in R 2.12.0 in 2010. In this book, I'm going to cover R6 instead of RC for a few reasons:

* R6 is much simpler. Both R6 and RC are built on top of environments, but
  while R6 uses S3, RC uses S4. R6 is only ~500 lines of code (and ~1700 lines
  of tests!)

* RC relies on R's regular environment scoping rules to get and set field
  values. This has a couple of downsides: it makes cross-package inheritance 
  more complex, and it makes it harder to see which variables are fields.

* RC is quite slow. Generally, method dispatch will not be a bottleneck in your
  code, but R6 is much faster than RC. Switching from RC to R6 had a substantial
  impact on the performance of the shiny pacakge.

* It will only take a small amount of additional effort if you want to learn
  RC once you have mastered R6.

More details about R6 at `vignette("Introduction", package = "R6")`

## Classes and methods

Because R6 implements encapsulated OO, you create a class and its method at the same time. The one function you need to know about is `R6::R6Class`, which creates new classes. The following example shows the two most important arguments:

* The first argument is the `classname`. It's not strictly needed, but it
  improves error messages and makes it possible to also use R6 objects
  with S3 generics. By convention, R6 classes use UpperCamelCase.
  
* The second argument, `public`, supplies a list of methods (functions) and 
  fields (anything else) that make up the object. By convention, methods
  and field use snake case.

```{r}
Accumulator <- R6Class("Accumulator", list(
  sum = 0,
  add = function(x = 1) {
    self$sum <- self$sum + x 
    invisible(self)
  })
)
```

You assign the result of `R6Class` into an object with the same name as the class. This object is an R6 object that defines the class:

```{r}
class(Accumulator)
```

And you use it to generate objects from the class by calling the `new()` method. Because the method belongs to the object, you use `$` to access it.

```{r}
x <- Accumulator$new() 
```

You can then call methods on that object, or access fields.

```{r}
x$add(4) 
x$sum
```

By default, fields are public which menas that you can set or get any field value. We'll see later how to use private fields to make it more clear that a field isn't for public consumption.

Note that `add()` is called primarily for its side-effect of updating `self$sum`. In R6, it's often useful to invisibly return `self`. This makes it possible to chain together multiple method calls.

```{r}
x$add(10)$add(10)
x$sum
```

### Initializer

Like S4, can override default behaviour of `new()` with `initialize()` method. Looks different , but same idea.

```{r}
Person <- R6Class("Person", list(
  age = NA,
  name = NULL,
  initialize = function(name, age = NA) {
    self$name <- name
    self$age <- age
  }
))

hadley <- Person$new("Hadley")
```

### Print

```{r}
Person <- R6Class("Person", list(
  age = NA,
  name = NULL,
  initialize = function(name, age = NA) {
    self$name <- name
    self$age <- age
  },
  print = function(...) {
    cat("Person: \n")
    cat("  Name: ", self$name, "\n", sep = "")
    cat("  Age:  ", self$age, "\n", sep = "")
  }
))

hadley2 <- Person$new("Hadley")
hadley2
```

This also illustrates an important aspect of R6. Methods are bound to individual objects. This means that the `hadley` object I created previously does not get this new method. Indeed, from the perspective of R6, there is no relationship between `hadley` and `hadley2`.

```{r}
hadley
```

### Private methods

It's also possible to define "private" fields and methods, elements of the class that can only be accessed from within the class, not from outside of it. This distinction tends to be more important in other programming langauges where you are creating complex networks of objects, and want to make it as clear as possible what it's ok for others to access. Anything that's private can be more easily refactored because you know others aren't relying on it.

```{r}
Person <- R6Class("Person", 
  public = list(
    initialize = function(name, age = NA) {
      private$name <- name
      private$age <- age
    },
    print = function(...) {
      cat("Person: \n")
      cat("  Name: ", private$name, "\n", sep = "")
      cat("  Age:  ", private$age, "\n", sep = "")
    }
  ),
  private = list(
    age = NA,
    name = NULL
  )
)

hadley3 <- Person$new("Hadley")
hadley3$name
```


## Reference semantics

Usually when you modify an object in R, a copy is created and the original object is left unchanged. This is not so with R6.

```{r}
y1 <- Accumulator$new() 
y2 <- y1

y1$add(10)
y2$sum
```

### Finalizer

### R6 fields

Beware of setting a default value to an R6 class - it will be shared across all instances of the object. Probably not what you were expecting!

## Functional vs encapsulated

Why is reasoning harder?

Syntax differencdes

## Inheritance

`super`


