# R6

This chapter describes the R6 object system. Unlike S3 and S4, it provides encapsulated OO, which means that:

* R6 methods belong to objects, not generics.

* R6 objects are mutable: the usual copy-on-modify semantics do not apply.

These properties make R6 objects behave more like objects in other popular programming languages like python, ruby and java. Unlike S3 and S4, R6 is not built in to R and you'll need to load a package to use it:

```{r setup}
library(R6)
```

R6 is very similar to a built-in OO system called __reference classes__, or RC for short. RC is the newest OO system in R: it was introduced in R 2.12.0 in 2010. In this book, I'm going to teach you R6 instead of RC for four reasons:

* R6 is much simpler. Both R6 and RC are built on top of environments, but
  while R6 uses S3, RC uses S4. R6 is only ~500 lines of R code (and ~1700 
  lines of tests!). We're not going to discuss the implemetation in depth
  here, but if you've mastered the contents of this book, you should be able
  to read the source code and figure out how it works.

* RC relies on R's regular environment scoping rules to get (`field`) and set
  field (`field <<- value`) values. R6 uses a special object to get
  (`self$field`) and set (`self$field <- vaue`) values. The R6 approach
  makes it easier to reason about code (because you can easily distinguish
  fields from local variables) and makes cross-package inheritance simpler.

* R6 is much faster than RC. Generally, method dispatch is not a bottleneck 
  in code, but the performance of RC is quite bad. Switching from RC to R6
  yieleded substantial performance in shiny. More details on the performance
  of R6 vs RC are available in `vignette("Performance", "R6")`.

* Because the ideas that underlie R6 and RC are similar, it will only require 
  a small amount of additional effort to learn R6 if you need to.

## Classes and methods

Unlike S3 and S4, when using R6 a single function call creates both the class and its methods. Indeed, there's actually only one function in R6 that you will ever use: `R6::R6Class()`. The following example shows the two most important arguments:

* The first argument is the `classname`. It's not strictly needed, but it
  improves error messages and makes it possible to also use R6 objects
  with S3 generics. By convention, R6 classes use UpperCamelCase.
  
* The second argument, `public`, supplies a list of methods (functions) and 
  fields (anything else) that make up the public interface of the object. 
  By convention, methods and fields use snake case. Methods can access
  the methods and fields of the current object via `self$`.

```{r}
Accumulator <- R6Class("Accumulator", list(
  sum = 0,
  add = function(x = 1) {
    self$sum <- self$sum + x 
    invisible(self)
  })
)
```

You should always assign the result of `R6Class()` into an variable with the same name as the class. This is an R6 object that defines the class:

```{r}
Accumulator
```

Generate a new object from the class by calling the `new()` method. Methods belong to R6 objects so you use `$` to access `new()`:

```{r}
x <- Accumulator$new() 
```

You can then call methods and access fields with `$`:

```{r}
x$add(4) 
x$sum
```

In this class, the fields and methods are public which means that you can get or set the value of any field. Later, we'll see how to use private fields and methods to prevent casual access tothe internals of your class.

To make it clear when we're are talking about fields and methods as opposed to variables and functions, when referring to them in text, we'll prefix with `$`. For example, the `Accumulate` class has field `$sum` and method `$add()`.

### Method chaining

`$add()` is called primarily for its side-effect of updating `$sum`. Side-effect R6 methods should always return `self` invisibly. This returns the "current" object and makes it possible to chain together multiple method calls:

```{r}
x$add(10)$add(10)$sum
```

For long chains, or you can spread the call over mulitple lines:

```{r}
x$
  add(10)$
  add(10)$
  sum
```

This technique is called __method chaining__ and is commonly used in encapsulated OO languages (like Python and JavaScript) to create fluent interfaces. Method chaining achieves similar goals to the pipe operator (`%>%`) in R. The primary advantage of method chaining is that you get useful autocomplete; the primary disadvantage is that only the creator of the class can add new methods.

### Important methods

There are two important methods that will be defined for most classes: `$initialize()` and `$print()`. You don't have to provide them, but it's a good idea to because they will make your class easier to use.

Similarly to S4, `$initialize()` overrides the default behaviour of `$new()`. For example, the following code defines an R6 Person class, similar to the S4 equivalent in [S4]. Unlike S4, R6 provides no checks for object type by default. `$initialize()` is a good place to check that `name` and `age` are the correct types.

```{r}
Person <- R6Class("Person", list(
  name = NULL,
  age = NA,
  initialize = function(name, age = NA) {
    stopifnot(is.character(name), length(name) == 1)
    stopifnot(is.numeric(age), length(age) == 1)
    
    self$name <- name
    self$age <- age
  }
))

hadley <- Person$new("Hadley", age = 37)
```

If you have more expensive validation requirements, implement in a separate `validate()` method.

Defining `$print()` allows you to override the default printing behaviour. As with any R6 method called for its side effects, `$print()` should return `invisible(self)`.

```{r}
Person <- R6Class("Person", list(
  name = NULL,
  age = NA,
  initialize = function(name, age = NA) {
    self$name <- name
    self$age <- age
  },
  print = function(...) {
    cat("Person: \n")
    cat("  Name: ", self$name, "\n", sep = "")
    cat("  Age:  ", self$age, "\n", sep = "")
    invisible(self)
  }
))

hadley2 <- Person$new("Hadley")
hadley2
```

This code illustrates an important aspect of R6. Because methods are bound to individual objects, the previously created `hadley` does not get this new method:

```{r}
hadley
```

Indeed, from the perspective of R6, there is no relationship between `hadley` and `hadley2`. This can make interactive experimentation with R6 confusing. If you can't figure out why your code changes don't appear to be working, make sure you've re-constructed R6 objects with the new class.

With R6, there's a useful alternative to providing `$print()`. You can instead provide `$format()`, which should return a character vector. This will automatically be used by both `print()` and `format()` S3 generics. 

```{r}
Person <- R6Class("Person", list(
  age = NA,
  name = NULL,
  initialize = function(name, age = NA) {
    self$name <- name
    self$age <- age
  },
  format = function(...) {
    # The first `paste0()` is not necessary but it lines up
    # with the subsequent lines making it easier to see how
    # it will print
    c(
      paste0("Person:"),
      paste0("  Name: ", self$name),
      paste0("  Age:  ", self$age)
    )
  }
))

hadley3 <- Person$new("Hadley")
format(hadley3)
hadley3
```

### Private fields and methods

With R6 you can define "private" fields and methods, elements that can be only  accessed from within the class, not from outside. There are two things that you need to know to take advantage of private elements:

* The third argument to `R6Class()`, `private`, works in the same way
  as the second argument, `public`.
  
* Fields and methods defined in `private` are available within the methods
  with `private$` instead of `self$`. You can not access private fields or
  methods outside of the class.
  
To make this concrete, we could make `$age` and `$name` fields of the Person class private. 

```{r}
Person <- R6Class("Person", 
  public = list(
    initialize = function(name, age = NA) {
      private$name <- name
      private$age <- age
    },
    print = function(...) {
      cat("Person: \n")
      cat("  Name: ", private$name, "\n", sep = "")
      cat("  Age:  ", private$age, "\n", sep = "")
    }
  ),
  private = list(
    age = NA,
    name = NULL
  )
)

hadley4 <- Person$new("Hadley")
hadley4$name
```

The distinction between public and private fields is more important when you create complex networks of classes, and you want to make it as clear as possible what it's ok for others to access. Anything that's private can be more easily refactored because you know others aren't relying on it. Private methods tend to be more important in other programming langauges compared to R because this is not what you are generally using R for.

### Inheritance

To inherit behaviour from an existing class, provide the class generator to the the `inherit` argument:

```{r}
AccumulatorChatty <- R6Class("AccumulatorChatty", 
  inherit = Accumulator,
  public = list(
    add = function(x = 1) {
      cat("Adding ", x, "\n", sep = "")
      super$add(x = x)
    }
  )
)

x2 <- AccumulatorChatty$new()
x2$add(10)$add(1)$sum
```

Note that `$add()` overrides the implementation in the superclass, but we can access the previous implementation through `super$`.


### Introspection

Because R6 is implement on top of (and designed to be a compatible with) S3, you can see the class hierarchy using `class()`:

```{r}
class(hadley4)
```

All R6 objects inherit from the base "R6" class which provides common behaviour, including the an S3 `print()` method which calls the R6 `$print()` method discussed above.

You can list all methods and fields with `names()`:

```{r}
names(hadley4)
```

There's one method that we haven't defined: `$clone()`. It's provided by provded by the R6 base class and we'll come back to it in the next section when we discuss the reference semantics that R6 objects possess.

There's no way to list private or fields methods from outside the class. They're private for a reason!

## Reference semantics

One of the big differences between R6 and most other objects in R is that they have reference semantics. This is because they are S3 objects built on top of environments:

```{r}
typeof(x2)
```

The main consequence of reference semantics is that objects are not copied when modified:

```{r}
y1 <- Accumulator$new() 
y2 <- y1

y1$add(10)
y2$sum
```

You can manually copy an R6 object by calling `$clone()`.

```{r}
y1 <- Accumulator$new() 
y2 <- y1$clone()

y1$add(10)
y2$sum
```

### Reasoning

Generally, reference semantics makes code harder to reason about. For example, take a very simple call like `z <- f(x, y)`. For most objects, if `f()` can't modify your `x` and `y` objects: it will always create its own copies. However, if `x` and `y` are R6 objects, they might be modified. Generally, you want to avoid side-effects in functions in your functions so use R6 with care. 

That said, this behaviour is one of the primary reasons to choose R6 instead of S3. It can be useful to pass an object into a function and modify it in place so that you don't need to return it (Threading state). For example, ggplot2 uses R6 objects to implement scales. Scales need to get trained for every facet on every layer. Carefully threading an S3 object through all the calls and returning the layer is a lot of work. That said, using R6 has introduced fairly subtle bugs because calls that should've been independent no longer were. Fixing these bugs required careful placement of calls to `$clone()`.

### Finalizer

Another useful property of reference semantics is that it makes sense to think about when an R6 object is __finalised__, or when it's deleted. This doesn't make sense for S3 and S4 objects because the copy-on-modify semantics mean that there may be many iterations of the same object that get discarded. For example, in the following code, there are actually two factor objects. The second is created when the levels are modifying, leaving the first to get destroyed at the next garbage collection.

```{r}
x <- factor(c("a", "b", "c"))
levels(x) <- c("c", "b", "a")
```

Since R6 objects are not copied-on-modify it makes more they will only get deleted once, and it can make sense to define a finalise method. This usually plays a similar role to `on.exit()` allowing you to clean up any resources created by the class. 

For example, the following class wraps up a temporary file, automatically deleting it when the class is finalised.

```{r}
TemporaryFile <- R6Class("TemporaryFile", list(
  path = NULL,
  initialize = function() {
    self$path <- tempfile()
  },
  finalize = function() {
    message("Cleaning up ", self$path)
    unlink(self$path)
  }
))

tf <- TemporaryFile$new()
```

The the finalise method will by the first garbage collection after the object has been removed. Generally, this will happen it when it happens, but it can occassionally be useful to force a run with an explicit call to `gc()`.

```{r}
rm(tf)
invisible(gc())
```

### R6 fields

Another consequence of reference semantics can crop up where you don't expect it. Beware of setting a default value to an R6 class - it will be shared across all instances of the object. This is because `$new()` is only called once when you defined the class, not each time you call new.

```{r}
TemporaryDatabase <- R6Class("TemporaryDatabase", list(
  con = NULL,
  file = TemporaryFile$new(),
  initialize = function() {
    DBI::dbConnect(RSQLite::SQLite(), path = file$path)
  }
))

db_a <- TemporaryDatabase$new()
db_b <- TemporaryDatabase$new()

db_a$file$path == db_b$file$path
```

You can fix this by creating the object in the `initialize()`:

```{r}
TemporaryDatabase <- R6Class("TemporaryDatabase", list(
  con = NULL,
  file = NULL,
  initialize = function() {
    self$file <- TemporaryFile$new()
    DBI::dbConnect(RSQLite::SQLite(), path = file$path)
  }
))

db_a <- TemporaryDatabase$new()
db_b <- TemporaryDatabase$new()

db_a$file$path == db_b$file$path
```

## Learning more

If you'd like to learn more about R6, the best place to start is the vignettes included in the package. You can list them by calling `browseVignettes(package = "R6")`.
