# S3 {#s3}

S3 is R's first and simplest OO system. It is the only OO system used in the base and stats packages, and it's the most commonly used system in CRAN packages. S3 is informal and ad hoc, but it has a certain elegance in its minimalism: you can't take away any part of it and still have a useful OO system. Because of these reasons, S3 should be your default choice for OO programming: you should use it unless you have a compelling reason otherwise.\index{S3} \index{objects!S3|see{S3}}

S3 is a very flexible system: it allows you to do a lot things that are quite ill-advised. To use S3 safely and efficiently, you need to impose constraints yourself. Those constraints will be a focus of this chapter.

## Basics {#s3-basics}

An S3 object is built on top of a base type (typically a vector) and must have the "class" attribute set. For example, take the factor. It's built on top of an integer vector, and the value of the class attribute is "factor". It stores information about the "levels" in another attribute.

```{r}
f <- factor("a")

typeof(f)
attributes(f)
```

S3 objects differ in behaviour from the underlying base type because of __generic functions__, or generics for short. A generic behaves differently depending on the class of one of its arguments (almost always the first). You can see this difference with the most important generic function: `print()`.

```{r}
print(f)
print(unclass(f))
```

`unclass()` strips the class attribute from its input, so is a useful tool for seeing what special behaviour an S3 class adds. Be aware when using `str()`: some S3 classes provide a custom `str()` method which may attempt to hide the underlying reality. For example, take the `POSIXlt` class, which is one of the two classes used to represent date-time data:

```{r}
time <- strptime("2017-01-01", "%Y-%m-%d")
str(time)
str(unclass(time))
```

Generics behave differently for different classes because generics have __methods__. A method is a function that implements the generic behaviour for a specific class. The generic doesn't actually do any work: it's job is to find the right method and pass on its arguments. Remember that S3 is from the generic-functions school of OO, so that methods belong to the generic, not the the object or the class. 

You can recognise S3 methods by their names, which look like `generic.class()`. For example, the Date method for the `mean()` generic is called `mean.Date()`, and the factor method for `print()` is called `print.factor()`. This is the reason that most modern style guides discourage the use of `.` in function names: it makes them look like S3 methods. For example, is `t.test()` the `t` method for `test` objects? Similarly, the use of `.` in class names can also be confusing: is `print.data.frame()` the `print()` method for `data.frames`, or the `print.data()` method for `frames`?  

You can sometimes find the source code for an S3 method by typing its name. This will work for S3 methods in the base package and your own code, but will not work with most packages because S3 methods are not exported. Instead, you can use `getS3method()`, which will work regardless of where the method lives:

```{r}
getS3method("mean", "Date")
```

### Exercises

1.  The most important S3 objects in base R are factors, data.frames,
    and date/times. You've already seen the attributes and base type 
    that factors are build on. What base types and attributes are
    the others built on?

1.  Describe the difference in behaviour in these two calls.

    ```{r}
    some_days <- as.Date("2017-01-31") + sample(10, 5)
    
    mean(some_days)
    mean(unclass(some_days))
    ```

1.  Draw a Venn diagram illustrating the relationship between
    functions, generics, and methods.

1.  What does the `is.data.frame.data.frame()` method do? Why is
    it confusing?

1.  What does the following code return? What base type is built on?
    What attributes does it use?
    
    ```{r}
    x <- ecdf(rpois(100, 10))
    x
    ```

## Classes

S3 is a simple and ad hoc system; it has no formal definition of a class. To make an object an instance of a class, you just take an existing object and set the class attribute. You can do that during creation with `structure()`, or after the fact with `class<-()`: \index{S3!classes} \index{classes!S3}

```{r}
# Create and assign class in one step
foo <- structure(list(), class = "foo")

# Create, then set class
foo <- list()
class(foo) <- "foo"
```

You can determine the class of any object using `class(x)`, and see if an object inherits from a specific class using `inherits(x, "classname")`. \index{attributes!class}

```{r}
class(foo)
inherits(foo, "foo")
```

Class names can be any string, but I recommend using only lower case letters and `_`. Avoid `.`. Opinion is mixed whether to use underscores (`my_class`) or CamelCase (`MyClass`) for multi-word class names. Pick one convention and stick with it.

S3 has no checks for correctness. This means you can change the class of existing objects:

```{r, error = TRUE}
# Create a linear model
mod <- lm(log(mpg) ~ log(disp), data = mtcars)
class(mod)
print(mod)

# Turn it into a data frame (?!)
class(mod) <- "data.frame"

# Unsurprisingly this doesn't work very well
print(mod)
```

If you've used other OO languages, this might make you feel queasy. But surprisingly, this flexibility causes few problems: while you _can_ change the type of an object, you never _should_. R doesn't protect you from yourself: you can easily shoot yourself in the foot. As long as you don't aim the gun at your foot and pull the trigger, you won't have a problem.

### Constructors

Since S3 doesn't check that your object is valid (i.e. it has the right attributes of the right types), it's up to you to adopt a convention to protect yourself. Do so with a __constructor__ which extracts out object creation code into a single place. The job of the constructor is enforce consistency. It ensures that whenever you create an S3 object of a specific class it is built on the same base type with the same attributes. 

Base R generally does not use this convention, so we'll demonstrate constructors by filling in some missing functions. (If your code works a lot with base objects that don't have a constructor, you might consider writing one yourself, just to keep your code consistent). In base R, the simplest useful class is Date: it's just a double with a class attribute.

```{r}
new_Date <- function(x) {
  stopifnot(is.double(x))
  structure(x, class = "Date")
}

new_Date(c(-1, 0, 1))
```

Constructors should always be called `new_class_name()`. Here we have the slightly awkward `new_Date()`, because the existing base class uses a capital letter. I recommend using snake case class names to avoid this problem. 

Generally, constructors will be used by developers (i.e. you). That means they can be quite simple, and you don't need to optimise the error messages for user friendliness. If you expect others to create these objects, you should also create a helpful function, called `class_name()`, and you may want to consider a coercion function called `as_class_name()`.

A more complicated example is `POSIXct`, which is used to represent date-times. It is again built on a double, but has an attribute that specifies the time zone which must be a length 1 character vector. The arguments to the constructor should match the attributes of the created object.

```{r}
new_POSIXct <- function(x, tzone = "") {
  stopifnot(is.double(x))
  stopifnot(is.character(tzone), length(tzone) == 1)
  
  structure(x, 
    class = c("POSIXct", "POSIXt"),
    tzone = tzone
  )
}

new_POSIXct(1)
new_POSIXct(1, tzone = "UTC")
```

(Note that we set the class to a vector; we'll come back to that in [Inheritance])

### Validators

More complicated classes will require more complicated checks for validity. Take factors, for example. The constructor function should ensure that you have an object of the correct structure:

```{r}
new_factor <- function(x, levels) {
  stopifnot(is.integer(x))
  stopifnot(is.character(levels))
  
  structure(
    x,
    levels = levels,
    class = "factor"
  )
}
```

But it's possible to use this to create invalid factors, because we don't ensure that the `x` and `levels` are compatible:

```{r, error = TRUE}
new_factor(1:5, "a")
new_factor(0:1, "a")
```

Rather than encumbering the constructor with complicated checks, it's better to put them in a separate function.

```{r, error = TRUE}
validate_factor <- function(x) {
  values <- unclass(x)
  levels <- attr(x, "levels")
  
  if (!all(!is.na(values) & values > 0)) {
    stop(
      "All `x` values must be non-missing and greater than zero",
      call. = FALSE
    )
  }
  
  if (length(levels) < max(values)) {
    stop(
      "There must at least as many `levels` as possible values in `x`",
      call. = FALSE
    )
  }
  
  x
}

validate_factor(new_factor(1:5, "a"))
validate_factor(new_factor(0:1, "a"))
```

(This function is called primarily for its side-effects (throwing an error if the object is invalid) so you'd expect it to invisibly returns its primary input. Validation methods, however, are an exception to the rule)

### Helpers

If you want others to construct objects of your new class, you should also provide a helper method that makes their life easy as possible. This should have the same name as a class, and will often provide more defaults and more checks. 

A good example of a helper is `factor()`: the internal representation is quite different to how you might want to create it in practice. The simplest possible implementation looks soemthing like this:

```{r}
factor <- function(x, levels = unique(x)) {
  ind <- match(x, levels)
  validate_factor(new_factor(ind, levels))
}
factor(c("a", "a", "b"))
```

Constrast `factor` with `Date` and `POSIXct`. Neither of these have helpers in base R because there's no particularly natural way for them to be constructed. Instead they provide a coercion function that lets you create from existing base types. We'll come back to that idea in [coercion].

### Object styles

S3 gives you the freedom to build a new class on top of an existing base type. So far, we've focussed on the "vector" type of S3 object where you take an exist vector base type and add some attributes. Another common style is the "scalar" type, where you use a list with named elements. The constructors for the scalar type is slightly different, as you always put the elements in a list:

```{r}
new_my_class <- function(x, y, z) {
  structure(
    list(
      x = x,
      y = y,
      z = z
    ),
    class = "my_class"
  )
}
```

(For a real constructor, you'd also check that `x`, `y`, and `z` are the types that you expect.)

These are not the only styles, but they are the most common. Other less common but still useful structures are:

*   Environments with classes, which allow you to implement new styles of OO.
    This is beyond the scope of this book, because generally you shouldn't
    be creating your own OO style!

*   "Vector" lists, like data frames. This is technically a subtype of the
    "vector type", where the vector is a list.

*   Functions with classes. This is mostly useful to override the print method
    as you can't override the call method.

### Exercises

1.  Using the above system, categorise the objects returned by `lm()`,
    `factor()`, `table()`, `ecdf()`, `ordered()`, `I()`.

1.  Write a constructor for `difftime` objects. What base type are they
    built on? What attributes do they use?

1.  Write a constructor for `data.frame` objects. What base type is a data 
    frame build on? What attributes does it use? What are the restrictions 
    placed on the individual elements? What about the names?

1.  Enhance our `factor()` helper to have better behaviour when one or
    more `values` is not found in `levels`. What does `base::factor()` do
    in this situation?

1.  Carefully read the source code of `factor()` what does it do that
    our constructor does not?

## Generics

To determine if a function is an S3 generic, you can inspect its source code for a call to `UseMethod()`: that's the function that figures out the correct method to call, the process of __method dispatch__.  \indexc{UseMethod()}

```{r}
print
```

To add a new generic, create a function that calls `UseMethod()`. `UseMethod()` takes two arguments: the name of the generic function, and the argument to use for method dispatch. If you omit the second argument it will dispatch on the first argument to the function. 

There you don't pass any of the arguments of the generic to `UseMethod()`. `UseMethod()` uses black magic to find them out for itself. \indexc{UseMethod()} \index{S3!new generic}

```{r}
f <- function(x) UseMethod("f")
```

There are two main reasons to write your own generic:

* To avoid a big compound if statement
* To make it possible for others to extend your code



### Coercion

One common use of a generic is as a coercion method, i.e. `as_my_class`. (Base R often uses `as.` but I think it's better to use `_` to avoid method names like `as.data.frame.data.frame`).

You should always have `as_my_class.my_class <- function(x, ...) x` (i.e. coercing an object that's already of the right type should do nothing.). Generally each coercion method should do a little work then call the constructor.

```{r}
as_factor <- function(x) UseMethod("as_factor")

as_factor.factor <- function(x) x
as_factor.character <- function(x) new_factor(x, levels = unique(x))
as_factor.integer <- function(x) new_factor(x, levels = unique(x))
```

You may want to consider adding a "default" method. The default method is called if no specific method is found. We'll talk more about this in [inheritance]. This is a good place to add a more informative error message

```{r, error = TRUE}
as_factor(1)

as_factor.default <- function(x) {
  stop(
    "Don't know how to coerce object of class ", 
    paste(class(x), collapse = "/"), " into a factor", 
    call. = FALSE
  )
}
as_factor(1)
```

### Esoterica

Note that `UseMethod()` calls methods in a special way:

```{r}
g <- function(x) {
  x <- 10
  y <- 10
  UseMethod("g")
}
g.default <- function(x) c(x = x, y = y)

x <- 1
y <- 1
g(x)
```

Arguments to the generic are passed on as they were recieved by the generic, regardless of what's happened to them in the meantime. New variables, however, are passed on as is. You should never rely on this behaviour because it is esoteric. Instead, you should avoid performing complex calculations in the generic. Most generics should simply contain `UseMethod()` and nothing else.

### Group generics

Group generic methods add a little more complexity. Group generics make it possible to implement methods for multiple generics with one function. The four group generics and the functions they include are: \index{group generics} \index{S3!group generics}

* Math: `abs`, `sign`, `sqrt`, `floor`, `cos`, `sin`, `log`, `exp`, ...
* Ops: `+`, `-`, `*`, `/`, `^`, `%%`, `%/%`, `&`, `|`, `!`, `==`, `!=`, `<`, `<=`, `>=`, `>`
* Summary: `all`, `any`, `sum`, `prod`, `min`, `max`, `range`
* Complex: `Arg`, `Conj`, `Im`, `Mod`, `Re`

Note that inside a group generic function a special variable `.Generic` provides the actual generic function called.

### Internal generics

Some S3 generics, like `[`, `sum()`, and `cbind()`, don't call `UseMethod()` because they are implemented in C. Instead, they call the C functions `DispatchGroup()` or `DispatchOrEval()`. Functions that do method dispatch in C code are called __internal generics__. It's important to know which functions are internally generic, so you can write methods for them, and so you're aware of the slight differences in method dispatch. It's not easy to tell if a function is internally generic, because it just looks like a typical call to a C function:

```{r}
length
cbind
```

You can look in the documentation, in the global variable `.S3PrimitiveGenerics`, and in `?InternalMethods`.  All group generics are internal generics.

## Methods

A generic isn't useful without some methods. To add a method, you just create a regular function with the correct (`generic.class`) name: \index{methods!S3} \index{S3!methods}

```{r}
f.a <- function(x) "Class a"

a <- structure(list(), class = "a")
class(a)
f(a)
```

Adding a method to an existing generic works in the same way:

```{r}
mean.a <- function(x) "a"
mean(a)
```

As you can see, there's no check to make sure that the method returns the class compatible with the generic. It's up to you to make sure that your method doesn't violate the expectations of existing code.

### Method dispatch

S3 method dispatch is relatively simple. `UseMethod()` creates a vector of function names, like `paste0("generic", ".", c(class(x), "default"))` and looks for each in turn. The "default" class makes it possible to set up a fall back method for otherwise unknown classes. \index{S3!method dispatch rules}

```{r}
f <- function(x) UseMethod("f")
f.a <- function(x) "Class a"
f.default <- function(x) "Unknown class"

f(structure(list(), class = "a"))
# No method for b class, so uses method for a class
f(structure(list(), class = c("b", "a")))
# No method for c class, so falls back to default
f(structure(list(), class = "c"))
```

Once `UseMethod` has found the correct method, it's invoked in a special way. Rather than creating a new evaluation environment, it uses the environment of the current function call (the call to the generic), so any assignments or evaluations that were made before the call to UseMethod will be accessible to the method. The arguments that were used in the call to the generic are passed on to the method in the same order they were received.

Because methods are normal R functions, you can call them directly:

```{r}
c <- structure(list(), class = "c")
# Call the correct method:
f.default(c)
# Force R to call the wrong method:
f.a(c)
```

However, this is just as dangerous as changing the class of an object, so you shouldn't do it. Please don't point the loaded gun at your foot! The only reason to call the method directly is that sometimes you can get considerable performance improvements by skipping method dispatch. See [performance](#be-lazy) for details.



You can see all the methods that belong to a generic with `methods()`:

```{r}
methods("mean")
methods("t.test")
```

(Apart from methods defined in the base package, most S3 methods will not be visible: use `getS3method("generic", "class")` to read their source code.)

You can also list all generics that have a method for a given class:

```{r}
methods(class = "ts")
```

There's no way to list all S3 classes because there is no formal definition of an S3 class, as you'll learn shortly.


### Method families


* When implementing a vector class, you should implement these methods: `length`, `[`, `[<-`, `[[`, `[[<-`, `c`.  (If `[` is implemented `rev`, `head`, and `tail` should all work). 

* When implementing anything mathematical, implement `Ops`, `Math` and `Summary`.

* When implementing a matrix/array class, you should implement these methods: `dim` (gets you nrow and ncol), `t`, `dimnames` (gets you rownames and colnames), `dimnames<-` (gets you colnames<-, rownames<-), `cbind`, `rbind`.

* If you're implementing more complicated `print()` methods, it's a better idea to implement `format()` methods that return a string, and then implement `print.class <- function(x, ...) cat(format(x, ...), "\n"`. This makes for methods that are much easier to compose, because the side-effects are isolated to a single place.


### Double dispatch

How does `+` work.


### Exercises

1.  Read the source code for `t()` and `t.test()` and confirm that 
    `t.test()` is an S3 generic and not an S3 method. What happens if 
    you create an object with class `test` and call `t()` with it?
    
    ```{r, error = TRUE }
    x <- structure(1, class = "test")
    t(x)
    ```

1.  What classes have a method for the `Math` group generic in base R? Read 
    the source code. How do the methods work?

1.  R has two classes for representing date time data, `POSIXct` and 
    `POSIXlt`, which both inherit from `POSIXt`. Which generics have 
    different behaviours for the two classes? Which generics share the same
    behaviour?

1.  Which base generic has the greatest number of defined methods?

## Multiple "classes"

Not just inheritance. Better to think of S3 classes as an interface that defines a specific behaviour (and requried data).

<https://en.wikipedia.org/wiki/Composition_over_inheritance>

As discussed in the next section, R looks for methods in the order in which they appear in the class vector. So in this example, it would be like class A inherits from class B - if a method isn't defined for A, it will fall back to B. However, if you switched the order of the classes, the opposite would be true! This is because S3 doesn't define any formal relationship between classes, or even any definition of what an individual class is. If you're coming from a strict environment like Java, this will seem pretty frightening (and it is!) but it does give your users a tremendous amount of freedom. While it's very difficult to stop someone from doing something you don't want them to do, your users will never be held back because there is something you haven't implemented yet.

### Method dispatch


### "Duck" typing

`is.numeric()` is

> Methods for is.numeric should only return true if the base type of 
> the class is double or integer and values can reasonably be regarded 
> as numeric (e.g., arithmetic on them makes sense, and comparison 
> should be done via the base type).

(i.e. you should only ever override it to be false)

This is confusing because this definition of numeric is different to the definition of numeric used elsewhere: i.e. either an integer or a double. A better name would have been `is.numericish()` or similar.

### Interfaces

`POSIXt`

### Mixins

`I()`

```{r}
labelled <- function(x, label) {
  struture(x, 
    label = label,
    class = c("labelled", class(x))
  )
}

`[.labelled` <- function(x) {
  labelled(NextMethod(), attr(x, "label"))
}
```

### Conventions

S3 provides no rules as to the structure of inheritance. However, you will make your life easier if you impose some principles.

* A subclass should always be built on the same base type as a parent.

* If using attributes, a subclass should always contain at least the
  same attributes as the parent. If using named list elements, ensure
  you add, not subtract.

* Use a constructor to enforce the order of classes.

(Note that `POSIXct` and `POSIXlt` don't follow this definition because `POSIXct` is built on an double vector, and `POSIXlt` is build on a list. We'll come back to what the `POSIXt` "base class" does in [interface].)

### Constructor

If you want people to extend your class, you should faciliate this in the constructor by using `...` and `subclass`

```{r}
new_my_class <- function(x, y, ..., subclass = NULL) {
  structure(
    x,
    y = y,
    ...,
    class = c(subclass, "my_class")
  )
}
new_sub_class <- function(x, y, z, ..., subclass = NULL) {
  new_my_class(x, y, z, ..., subclass = c(subclass, "sub_class"))
}
```


### Default method

Sometimes you may want to customise just to have a more informative error for your audience.

### `NextMethod()`

The `NextMethod` function provides a simple inheritance mechanism, using the fact that the class of an S3 object is a vector. 

`NextMethod()` works like `UseMethod` but instead of dispatching on the first element of the class vector, it will dispatch based on the second (or subsequent) element:

```{r}
baz <- function(x) UseMethod("baz", x)
baz.A <- function(x) "A"
baz.B <- function(x) "B"

baz.C <- function(x) c("C", NextMethod())
ca <- structure(1, class = c("C", "A"))
cb <- structure(1, class = c("C", "B"))
baz(ca)
baz(cb)
```

The exact details are a little tricky: `NextMethod` doesn't actually work with the class attribute of the object, it uses a global variable (`.Class`) to keep track of which class to call next. This means that manually changing the class of the object will have no impact on the inheritance:

```{r}
# Turn object into class A - doesn't work!
baz.D <- function(x) {
  class(x) <- "A"
  NextMethod()
}
da <- structure(1, class = c("D", "A"))
db <- structure(1, class = c("D", "B"))
baz(da)
baz(db)
```

Methods invoked as a result of a call to `NextMethod` behave as if they had been invoked from the previous method. The arguments to the inherited method are in the same order and have the same names as the call to the current method, and are therefore are the same as the call to the generic. However, the expressions for the arguments are the names of the corresponding formal arguments of the current method. Thus the arguments will have values that correspond to their value at the time NextMethod was invoked. Unevaluated arguments remain unevaluated. Missing arguments remain missing.

If `NextMethod` is called in a situation where there is no second class it will return an error. 

### Writing flexible base methods


### Exercises

1.  The `ordered` class is a subtclass of `factor`, but it's implemented in 
    a very adhoc way in base R. Implement it in a principled way by
    building a constructor and an `as_ordered` generic.
    
    ```{r}
    f1 <- factor("a", c("a", "b"))
    as.factor(f1)  
    as.ordered(f1) # loses levels
    ```

## Base types

```{r}
x <- 1:10
class(x)
inherits(x, "integer")
inherits(x, "numeric")

foo <- function(x) UseMethod("foo")
foo.numeric <- function(x) TRUE
foo.default <- function(x) FALSE
foo(x)
```

### Interaction with base types

You can also call an S3 generic with a non-S3 object. Non-internal S3 generics will dispatch on the __implicit class__ of base types. (Internal generics don't do that for performance reasons.) The rules to determine the implicit class of a base type are somewhat complex, but are shown in the function below: \index{implicit class} \index{base types!implicit class}

```{r}
iclass <- function(x) {
  if (is.object(x)) {
    stop("x is not a primitive type", call. = FALSE)
  }

  c(
    if (is.matrix(x)) "matrix",
    if (is.array(x) && !is.matrix(x)) "array",
    typeof(x),
    if (is.integer(x) || is.double(x)) "numeric"
  )
}
iclass(matrix(1:5))
iclass(array(1.5))
```

```{r}
class(1:5)
```

### Internal generics

" internal generics only dispatch on objects for which is.object is true."

Internal generic have a slightly different dispatch mechanism to other generic functions: before trying the default method, they will also try dispatching on the __mode__ of an object, i.e. `mode(x)`. The following example shows the difference:

```{r, error = TRUE}
x <- structure(as.list(1:10), class = "myclass")
length(x)

mylength <- function(x) UseMethod("mylength", x)
mylength.list <- function(x) length(x)
mylength(x)
```

1.  Internal generics don't dispatch on the implicit class of base types.
    Carefully read `?"internal generic"` to determine why the length of `f` 
    and `g` is different in the example below. What function helps 
    distinguish between the behaviour of `f` and `g`?

    ```{r, eval = FALSE}
    f <- function() 1
    g <- function() 2
    class(g) <- "function"
    
    class(f)
    class(g)

    length.function <- function(x) "function"
    length(f)
    length(g)
    ```


### Numeric

We need a little extra discussion of the numeric "type" because it's used in three different ways in different places in R.

1.  In some places it's used as an alias for "double".  For example
    `as.numeric()` is identical to `as.double()`.
    
1.  In some places it means a base type of either integer or double. 

    ```{r, error = TRUE}
    mode(1)
    mode(1L)
    
    # Note the error messages here
    dummy <- function(x) UseMethod("dummy")
    dummy(1)
    dummy(1L)
    ```

1.  In some places it means an object build on a  base type of integer or 
    double that has numeric behaviour (i.e. arithmetic makes sense and you
    can order by numeric values)
    
    ```{r}
    is.numeric(1)
    is.numeric(1L)
    is.numeric(Sys.Date())
    is.numeric(factor("x"))
    ```

## Packaging

* Beware class clashes across packages. If there's any chance of confusion
  give your classes a common prefix.

