# S3 {#s3}

```{r, include = FALSE}
source("common.R")
```

## Introduction
\index{S3} 
\index{objects!S3|see{S3}} 

S3 is R's first and simplest OO system. S3 is informal and ad hoc, but it has a certain elegance in its minimalism: you can't take away any part of it and still have a useful OO system. Because of these reasons, I believe that S3 should be your default choice for OO programming: you should use it unless you have a compelling reason otherwise. S3 is the only OO system used in the base and stats packages, and it's the most commonly used system in CRAN packages. 

S3 is a very flexible system: it allows you to do a lot of things that are quite ill-advised. If you're coming from a strict environment like Java, this will seem pretty frightening (and it is!) but it does give R programmers a tremendous amount of freedom. While it's very difficult to prevent someone from doing something you don't want them to do, your users will never be held back because there is something you haven't implemented yet. Since S3 has few built-in constraints, the key to using it successfully is constraining yourself. This chapter will teach you the conventions you should (almost) always adhere to in order to use S3 safely.

The goal of this chapter is to show you how the S3 OOP system works, not how to use it effectively to solve data science problems. I'd recommend coupling the theoretical knoweledge from this chapter with the practical knowledge encoded in the [vctrs](https://vctrs.r-lib.org) package. The vctrs package makes it easier to create new classes by providing thoughtfully organised default methods.

### Outline {-}

### Prerequisites {-}

We'll use the sloop package to explore method dispatch, and the vctrs package for some niceties for creating new S3 classes.

```{r setup, messages = FALSE}
# install_github("r-lib/sloop")
# install_github("r-lib/vctrs")

library(sloop)
```

## Basics {#s3-basics}

An S3 object is a base type with a "class" attribute, and potential other data stored in attributes. For example, take the factor. Its base type is the integer vector, and it the value of the class attribute is "factor". It stores information about the "levels" in another attribute:

```{r}
f <- factor(c("a", "b", "c"))

typeof(f)
attributes(f)
```

You can get the "underlying" base type by `unclass()`ing it, which strips the class attribute, and hence custom behaviours:

```{r}
unclass(f)
```

An S3 object behaves differently from its underlying base type whenever it's passed to a __generic__ (short for generic function). A generic executes different code depending on the class of one of its arguments, almost always the first. Many base R functions are generic, including the important `print()`:

```{r}
print(f)
print(unclass(f))
```

And `mean()`:

```{r, error = TRUE}
mean(f)
mean(unclass(f))
```

The easiest way to tell if a function is to use `sloop::ftype()` and look for "generic" in the output:

```{r}
sloop::ftype(print)
sloop::ftype(mean)
sloop::ftype(`[`)
sloop::ftype(str)
```

Note that `str()` is generic, and some S3 classes use that generic to hide the internal details. For example, the `POSIXlt` class used to represent date-time data is actually built on top of a list, a fact which is hidden by its `str()` method:

```{r}
time <- strptime("2017-01-01", "%Y-%m-%d")
str(time)

str(unclass(time), list.len = 5)
```

The differences in behaviour is mediated through __methods__, functions that are called for specific classes on input. The generic is a middleman: its role is to find the right method for the job, the process of __method dispatch__.  The method is the function that implements the desired behaviour for a specific class. In other words the job of the generic is to find the right method; the job of the method is to do the work.

S3 methods are functions with a special naming scheme, `generic.class()`. For example, the Date method for the `mean()` generic is called `mean.Date()`, and the factor method for `print()` is called `print.factor()`. Generally, you can identify a method by the presence of `.` in the function name, but there are a number of important functions in base R that were written before S3, and use `.` to join words. Again, you can always check with `sloop::ftype()`:

```{r}
ftype(data.frame)
ftype(t.data.frame)
```

A generic isn't useful without some methods, which are just functions that follow a naming scheme (`generic.class`). Because a method is just a function with a special name, you _can_ call methods directly, but you generally _shouldn't_. 

Unlike most functions, you can't find the source code for most S3 methods[^base-s3] just by typing its name. That's because S3 methods are not usually exported: they live only inside the package, and are not available from the global environment. Instead, you can use `sloop::s3_get_method()`, which will work regardless of where the method lives:

```{r, error = TRUE}
weighted.mean.Date

s3_get_method(weighted.mean.Date)
```

[^base-s3]: The exceptions are methods found in the base package, like `t.data.frame`, and methods that you've created.

### Exercises

1.  The most important S3 objects in base R are factors, data frames,
    difftimes, and date/times (Dates, POSIXct, POSIXlt). You've already seen 
    the attributes and base type that factors are built on. What base types and 
    attributes are the others built on?

1.  What's the difference between `t.test()` and `t.data.frame()`?

1.  Describe the difference in behaviour in these two calls.

    ```{r}
    set.seed(1014)
    some_days <- as.Date("2017-01-31") + sample(10, 5)
    
    mean(some_days)
    mean(unclass(some_days))
    ```

1.  What does the `as.data.frame.data.frame()` method do? Why is
    it confusing? How should you avoid this confusion in your own
    code?

1.  What does the following code return? What base type is it built on?
    What attributes does it use?
    
    ```{r}
    x <- ecdf(rpois(100, 10))
    x
    ```

## Classes
\index{S3!classes} 
\index{classes!S3}
\indexc{class()}
\index{attributes!class}
 
S3 is a simple and ad hoc system, and has no formal definition of a class: to make an object an instance of a class, you simply set the __class attribute__. You can do that during creation with `structure()`, or after the fact with `class<-()`: 

```{r}
# Create and assign class in one step
x <- structure(list(), class = "my_class")

# Create, then set class
x <- list()
class(x) <- "my_class"
```

You can determine the class of an S3 object with `class(x)`, and see if an object inherits from a specific class using `inherits(x, "classname")`.

```{r}
class(x)
inherits(x, "my_class")
```

The class name can be any character vector, but I recommend using only letters and `_`. Avoid `.` because it makes life very confusing for reasons you'll learn shortly. When using a class in a package, I recommend including the package name in the class name. That ensures you won't accidental clash with a class defined by another package.

S3 has no checks for correctness which that means you can change the class of existing objects:

```{r, error = TRUE}
# Create a linear model
mod <- lm(log(mpg) ~ log(disp), data = mtcars)
class(mod)
print(mod)

# Turn it into a data frame (?!)
class(mod) <- "data.frame"

# Unsurprisingly this doesn't work very well
print(mod)
```

If you've used other OO languages, this might make you feel queasy. Surprisingly, however, this flexibility causes few problems: while you _can_ change the type of an object, you never _should_. R doesn't protect you from yourself: you can easily shoot yourself in the foot. As long as you don't aim the gun at your foot and pull the trigger, you won't have a problem.

To avoid foot-bullet intersections when creating your own class, I recommend that you provide:

* A low-level __constructor__, `new_x()`, that efficiently creates new objects 
  with the correct structure.
  
For more complicated classes, you may also want to provide:
  
* A __validator__, `validate_x()`, that performs more expensive checks that the 
  object has correct values.
   
* A __helper__, `x()`, that provides a convenient and neatly parameterised way 
  for others to create (construct and validate) objects of this class.

### Constructors

S3 doesn't provide a formal definition of a class, so it has no built-in way to ensure that all objects of a given class have the same structure (i.e. the same base and the same attributes with the same types). Instead, you should enforce a consistent structure yourself by using a __constructor__. A constructor is a function whose job is to construct objects of a given class, ensuring that they always have the same structure. 

There are three rules that a constructor should follow. It should:

1.  Be called `new_class_name()`.

1.  Have one argument for the base object, and one for each attribute. 

1.  Check the types of the base object and each attribute.

Base R generally does not provide constructors (three exceptions are the internal `.difftime()`, `.POSIXct()`, and `.POSIXlt()`) so we'll demonstrate these ideas by creating constructors for base classes. (If you want to use these constructors in your own code, please use use the versions exported by the vctrs package. They complete a few details that we skip here in order to focus on the core issues.)

We'll start with one of the simplest S3 classes in base R: Date, which is just a double with a class attribute. The constructor rules lead to the slightly awkward name `new_Date()`, because the existing base class uses a capital letter. I recommend using lower case class names to avoid this problem. 

```{r}
new_Date <- function(x = double()) {
  stopifnot(is.double(x))
  structure(x, class = "Date")
}

new_Date(c(-1, 0, 1))
```

The purpose of the constructor is to help the developer (you). That means you can keep them simple, and you don't need to optimise the error messages for user friendliness. If you expect others to create your objects, you should also create a friendly helper function, called `class_name()`, that we'll describe shortly.

A slightly more complicated example is `POSIXct`, which is used to represent date-times. It is again built on a double, but has an attribute that specifies the time zone, a length 1 character vector. R defaults to using the local time zone, which is represented by the empty string. To create the constructor, we need to make sure each attribute of the class gets an argument to the constructor. This gives us:

```{r}
new_POSIXct <- function(x = double(), tzone = "") {
  stopifnot(is.double(x))
  stopifnot(is.character(tzone), length(tzone) == 1)
  
  structure(x, 
    class = c("POSIXct", "POSIXt"),
    tzone = tzone
  )
}

new_POSIXct(1)
new_POSIXct(1, tzone = "UTC")
```

The constructor checks that `x` is a double, and that `tzone` is a length 1 character vector. Note that POSIXct uses a class _vector_; we'll come back to that in Section \@ref(inheritance).

Generally, the constructor should not check that the values are valid because such checks are often expensive. For example, our `new_POSIXct()` constructor does not check that `tzone` is a valid value, and we get a warning when the object is printed.

```{r, eval = FALSE}
x <- new_POSIXct(1, "Auckland NZ")
x
#> [1] "1970-01-01 00:00:01 GMT"
#> Warning messages:
#> 1: In as.POSIXlt.POSIXct(x, tz) : unknown timezone 'Auckland NZ'
```

### Validators

More complicated classes will require more complicated checks for validity. Take factors, for example. Since a constructor only checks that types are correct, it's possible to create malformed factors:

```{r, error = TRUE}
new_factor <- function(x, levels) {
  stopifnot(is.integer(x))
  stopifnot(is.character(levels))
  
  structure(
    x,
    levels = levels,
    class = "factor"
  )
}

new_factor(1:5, "a")
new_factor(0:1, "a")
```

Rather than encumbering the constructor with complicated checks, it's better to put them in a separate function. This is a good idea because it allows you to cheaply create new objects when you know that the values are correct, and to re-use the checks in other places.

```{r, error = TRUE}
validate_factor <- function(x) {
  values <- unclass(x)
  levels <- attr(x, "levels")
  
  if (!all(!is.na(values) & values > 0)) {
    stop(
      "All `x` values must be non-missing and greater than zero",
      call. = FALSE
    )
  }
  
  if (length(levels) < max(values)) {
    stop(
      "There must at least as many `levels` as possible values in `x`",
      call. = FALSE
    )
  }
  
  x
}

validate_factor(new_factor(1:5, "a"))
validate_factor(new_factor(0:1, "a"))
```

This function is called primarily for its side-effects (throwing an error if the object is invalid) so you'd expect it to invisibly return its primary input. However, unlike most functions called for their side effects, it's useful for validation methods to return visibly, as we'll see next.

### Helpers

If you want users to construct objects from your class, you should also provide a helper method that makes their life as easy as possible. This should have the same name as the class, and should be parameterised in a convenient way: exactly what this means will vary from case-to-case so it's hard to give general advice. However, we can look at a couple of concrete examples.

Take `factor()`, for example. It's a helper that constructs a factor from a convenient string representation. The simplest possible implementation looks something like the code below:

```{r, error = TRUE}
factor <- function(x, levels = unique(x)) {
  ind <- match(x, levels)
  validate_factor(new_factor(ind, levels))
}
factor(c("a", "a", "b"))

factor(c("a", "a", "b"), levels = "a")
```

What about `Date()` and `POSIXct()`? Base provides neithers `Date` nor `POSIXct` helper but instead provides `as.Date()` and `as.POSIXct()`, and `strptime()` and `ISODatetime()`. I think the most natural way to specify a date or date-time is from year-month-day components, which we could implement as follows:

```{r}
Date <- function(year, month, day) {
  as.Date(ISOdate(year, month, day, tz = ""))
}

POSIXct <- function(year, month, day, hour, minute, sec, tzone = "") {
  ISOdatetime(year, month, day, hour, minute, sec, tz = tzone)
}
```

These helpers fill a useful hole, but are not computationally efficient: behind the scenes `ISODatetime()` works by pasting the components into a string and then using `strptime()`. More efficient equivalents are `lubridate::make_datetime()` and `lubridate::make_date()`.

### Exercises

1.  Write a constructor for `difftime` objects. What base type are they
    built on? What attributes do they use? You'll need to consult the
    documentation, read some code, and perform some experiments.

1.  Write a constructor for `data.frame` objects. What base type is a data 
    frame built on? What attributes does it use? What are the restrictions 
    placed on the individual elements? What about the names?

1.  Enhance our `factor()` helper to have better behaviour when one or
    more `values` is not found in `levels`. What does `base::factor()` do
    in this situation?

1.  Carefully read the source code of `factor()`. What does it do that
    our constructor does not?
    
1.  Factors have an optional "contrasts" attribute. Read the help for `C()`, 
    briefly describe it's purpose. Rewrite our `new_factor()` constructor to
    include this attribute.

## Generics and methods
\indexc{UseMethod()} 
\index{S3!new generic}

The job of an S3 generic is to perform method dispatch, i.e. find the function designed to work specifically for the given class. S3 generics have a simple structure: they call `UseMethod()`, which then calls the right method. `UseMethod()` takes two arguments: the name of the generic function (required), and the argument to use for method dispatch (optional). If you omit the second argument it will dispatch based on the first argument, which is almost always what is desired.

```{r}
# Dispatches on x
mean
```

Note that you don't pass any of the arguments of the generic to `UseMethod()`; it uses deep magic to pass to the method automatically. Avoid doing any computation in a generic, because the semantics are complicated. Carefully read the "technical details" section "?UseMethod" for the full details.

### Method dispatch

How does `UseMethod()` work? It basically creates a vector of function names, `paste0("generic", ".", c(class(x), "default"))`, then looks for each method in turn. As soon as it finds a matching method, it calls it. If no matching method is found, it throws an error. We can see this in action with `sloop::s3_dispatch()`. You give it a call to an S3 generic, and it lists all the possible methods, noting which ones exist. For example, what method is called when you print a `POSIXct` object?

```{r}
x <- Sys.time()
s3_dispatch(print(x))
```

`s3_dispatch()` displays a `=>` next to the method that is called: this is always the first method that exists. It displays a `*` next to every other method that exists.

`print()` will look for three possible methods, of which two exist, and one, `print.POSIXct()`, will be called. The last method is always the "default" method. This is called if there are no more specific methods.

The rules of method dispatch get a bit more complicated with base objects, inheritance, internal generics, and group generics, as hinted at in the following example. We'll come back to those details in Sections \@ref(inhertiance) and \@ref(dispatch-details). 

```{r}
x <- matrix(1:10, nrow = 2)
s3_dispatch(mean(x))

s3_dispatch(sum(Sys.time()))
```

### Finding methods

You can see all the methods defined for a generic with `s3_methods_generic()`:

```{r}
s3_methods_generic("mean")
```

Or all methods for a class with:

```{r}
s3_methods_class("ordered")
```

### Example



### Arguments {#s3-arguments}

Methods must always have the same arguments as their generics. This is enforced in packages by `R CMD check`, but it's good practice anyway. There is one exception to this rule: if the generic has `...`, the method must still have all the same arguments (including `...`), but can also have its own additional arguments. This allows methods to take additional arguments, which is important because you don't know what additional arguments that a method for someone else's class might need. The downside of using `...`, however, is that any misspelled arguments will be silently swallowed[^ellipsis].

[^ellipsis]: See <https://github.com/hadley/ellipsis> for an experimental way of warning whenever arguments in `...` are ignored.

### Exercises

1.  Read the source code for `t()` and `t.test()` and confirm that 
    `t.test()` is an S3 generic and not an S3 method. What happens if 
    you create an object with class `test` and call `t()` with it? Why?
    
    ```{r}
    x <- structure(1:10, class = "test")
    t(x)
    ```

1.  Carefully read the documentation for `UseMethod()` and explain why the
    following code returns the results that it does. What two usual rules
    of function evaluation does `UseMethod()` violate?
    
    ```{r}
    g <- function(x) {
      x <- 10
      y <- 10
      UseMethod("g")
    }
    g.default <- function(x) c(x = x, y = y)
    
    x <- 1
    y <- 1
    g(x)
    ```

1.  Which base generic has the greatest number of defined methods?

1.  The `table()` function produces output with class "table". What attributes
    does this class possess? What generics does it have methods for?

## Object styles

S3 gives you the freedom to build a new class on top of any existing base type. So far, we've focussed on vector-style where you take an existing vector type and add some attributes. Importantly, a single vector-style object represents multiple values. There are two other important styles: scalar-style and record-style. These are both expanded upon in the vctrs package, but here I give the basic idea.

* Vectors
* Scalars
* Records

Data frames special.

Also build on other types, primarily functions.

Each __scalar__-style object represents a single "value", and are built on top of named lists. This is the style that you are most likely to use in practice. The constructor for the scalar type is slightly different because the arguments become named elements of the list, rather than attributes.

```{r}
new_scalar_class <- function(x, y, z) {
  structure(
    list(
      x = x,
      y = y,
      z = z
    ),
    class = "scalar_class"
  )
}
```

(For a real constructor, you'd also check that the `x`, `y`, and `z` fields are the types that you expect.)

In base R, the most important example of this style is `lm`, the class returned when you fit a linear model:

```{r}
mod <- lm(mpg ~ wt, data = mtcars)
typeof(mod)
names(mod)
```

The __record-style__ is a list of equal length vectors, which behaves like a regular vector. In base R, the most important example of this style is `POSIXlt()`. It behaves like a vector of length `n`, but is actually a list of nine vectors each of which is length `n`.

I'll use __fields__ to collectively refer to the names of a record-style or list-style class and the attributes of a 

Collectively, we'll call the attributes of a vector-style the names of a list-style class the __fields__ of an object.

When creating your own classes, you should pick the vector style if your class closely resembles an existing vector type. Otherwise, use a scalar (list) style. The scalar type is generally easier to work with because implementing a full range of convenient vectorised methods is usually a lot of work. It's typically obvious when you need to use a data-frame-style.

### Exercises

1.  Categorise the objects returned by `lm()`, `factor()`, `table()`, 
    `as.Date()`, `ecdf()`, `ordered()`, `I()` into "vector", "scalar", and 
    "other".

1.  What would a constructor function for `lm` objects, `new_lm()`, look like?
    Why is a constructor function less useful for linear models?

## Inheritance

The class attribute is not limited to a single string, but can be a character vector. This, along with S3 method dispatch and `NextMethod()`, gives a surprising amount of flexibility that can be used creatively to reduce code duplication. However, this flexibility can also lead to code that is hard to understand or reason about, so you are best constraining yourself to simple styles of inheritance. Here we will focus on defining subclasses that inherit their fields, and some behaviour, from a parent class.

Subclasses use a character __vector__ for the class attribute. There are two subclasses that you might have come across in base R:

* Generalised linear models are a generalisation of linear models that allow
  the error term to belong to a richer set of distributions, not just the normal
  distribution like the linear model. This is a natural case for the use of 
  inheritance and indeed, in R, `glm()` returns objects of class 
  `c("glm", "lm")`.

* Ordered factors are used when the levels of a factor have some intrinsic
  ordering, like `c("Good", "Better", "Best")`. Ordered factors are produced
  by `ordered()` which returns an object with class `c("ordered", "factor")`.

(I don't consider the relationship between `POSIXct` and `POSIXt` to be one of subclassing because the other child class `POSIXlt`, is implemented using a completely different base type. `POSIXt` is kind of like an interface, but not really since it provide substantial implementation.)

You can think of the glm class "inheriting" behaviour from the lm class, and the ordered class inheriting behaviour from the factor class because of the way method dispatch works. If there is a method available for the subclass, R will use it, otherwise it will fall back to the "parent" class. For example, if you "plot" a glm object, it falls back to the lm method, but if you compute the ANOVA, it uses a glm-specific method.

```{r}
mod1 <- glm(mpg ~ wt, data = mtcars)

s3_dispatch(plot(mod1))
s3_dispatch(anova(mod1))

x <- ordered(letters)
s3_dispatch(as.data.frame(x))
s3_dispatch(sum(x))

s3_dispatch(Sys.Date()[1])
```

### `NextMethod()`

Method dispatch usually terminates as soon as a matching method is found. However, methods can explicitly choose to call the next available method using `NextMethod()`, making it possible to inherit behaviour from the "parent" class. Like `UseMethod()`, you generally call `NextMethod()` without any arguments (any arguments will be passed on to the next method, as if they'd been supplied to the generic).

The most common use case of `NextMethod()` is `[`: how do you implement a subsetting method for your own class, without using `[`? `NextMethod()` resolves the issue by allowing you to use the underlying C code. You can see that the base `[.POSIXct` and `[.Date` methods use `NextMethod()` by looking at the output of `s3_dispatch()`:

```{r}
s3_dispatch(Sys.Date()[1])
s3_dispatch(Sys.time()[1])
```

The first method is indicated with a `=>`, but you'll see that the internal `[` method gets a `->`: this indicates that it's called by `NextMethod()`.

Let's demonstrate with a very simple toy example: a `secret` class that hides its output when printed:

```{r}
new_secret <- function(x = double()) {
  stopifnot(is.double(x))
  structure(x, class = "secret")
}

print.secret <- function(x, ...) {
  print(strrep("x", nchar(x)))
  invisible(x)
}

x <- new_secret(c(15, 1, 456))
x
```

This is great, but the default `[` method doesn't preserve the class:

```{r}
s3_dispatch(x[1])
x[1]
```

To fix this we need to provide a `[.secret` method. This will rely on the usual subsetting behaviour, but will wrap the results back with the secret class:

```{r}
`[.secret` <- function(x, ...) {
  new_secret(NextMethod())
}

s3_dispatch(x[1])
x[1]
```

Like `UseMethod()`, the precise semantics of `NextMethod()` are complex. It doesn't modify the class attribute of the object, but instead uses a special global variable (`.Class`) to keep track of which method to call next. This means that modifying the object being dispatched on will have no impact on the method call.

### Allowing subclassing

When you create a class, you need to decide if you want to allow subclasses, because it requires changes to the constructor and careful thought in your methods. To allow subclasses, the parent constructor needs to have `...` and `class` arguments:

```{r}
new_secret <- function(x, ..., class = character()) {
  stopifnot(is.double(x))

  structure(
    x,
    ...,
    class = c(class, "secret")
  )
}
```

Imagine we want to create a supersecret class which also hides the number of characters:

```{r}
new_supersecret <- function(x) {
  new_secret(x, class = "supersecret")
}

print.supersecret <- function(x, ...) {
  print(rep("xxxxx", length(x)))
  invisible(x)
}

x2 <- new_supersecret(c(15, 1, 456))
x2
```

There are three principles to adhere to when creating a subclass:

* A subclass should be built on the same base type as a parent.

* The `class()` of the subclass should be of the form 
  `c(subclass, parent_class)`

* The fields of the subclass should include the fields of the parent.

You also need to think carefully about your methods as you can no longer just call the constructor, as that limits methods to returning the parent type, and forces a subclasser to override every method, even if the implementation is basically the same. For secret, that means we need to revise the `[` method which currently always returns a `secret()`. We want to make sure that `[.secret` returns the same class as `x` even if it's a subclass.

```{r}
`[.secret` <- function(x, ...) {
  new_secret(NextMethod())
}

x2[1:3]
```

As far as I can tell, there's no way to solve this problem using base R, but the vctrs package provides a general approach to resolving this problem through the `vec_restore()` generic. It takes two objects: an object of the most specific subclass to restore to, and the object to restore.

```{r}
`[.secret` <- function(x, ...) {
  vctrs::vec_restore(NextMethod(), x)
}

vec_restore.secret <- function(x, to) new_secret(x)
vec_restore.supersecret <- function(x, to) new_supersecret(x)

x2[1:3]
```

(I only fully understand this issue and how to resolve it relatively recently, so it is not yet implemented in any tidyverse packages. Once this rolls out (hopefully by the time you are reading this book), it will be much easier to extend classes like data frame and have them work with dplyr with out too much effort.)

However, if you are going to create your own new class, I recommend using `vctrs::new_vctr()` because it will automatically provide a `[` method with this behaviour. You only need to override it you want to do something special with `[` or you have objects attributes that depend on the data component.

### Other approaches

If you read the source code for base R methods, you'll see a variety of other approaches used to call parent class methods.

*   `unclass(x)`

*   `print.default()`

I recommend avoiding these disparate approaches and instead standardising on `NextMethod()`.

### Constructors

And these properties should be enforced by the constructor. 

Then the implementation of the subclass constructor is simple: it checks the types of the new fields, then calls the parent constructor. 

```{r}
new_subclass <- function(x, y, z) {
  stopifnot(is.character(z))
  new_my_class(x, y, z = z, subclass = "subclass")
}
```

If you wanted to allow this subclass to be futher subclassed, you'd need to include `...` and `subclass` arguments:

```{r}
new_subclass <- function(x, y, z, ..., subclass = NULL) {
  stopifnot(is.character(z))
  
  new_my_class(x, y, z = z, ..., subclass = c(subclass, "subclass"))
}
```

If your subclass is more complicated, you'd also provide validator and helper functions, as described previously.

### Exercises

1.  The `ordered` class is a subclass of `factor`, but it's implemented in
    a very ad hoc way in base R. Implement it in a principled way by
    building a constructor and an `as_ordered` generic.
    
    ```{r}
    f1 <- factor("a", c("a", "b"))
    as.factor(f1)  
    as.ordered(f1) # loses levels
    ```

1.  How does `[.Date` support subclasses? How does it fail to support 
    subclasses?

1.  Explain what is happening in the following code.

    ```{r}
    generic2 <- function(x) UseMethod("generic2")
    generic2.a1 <- function(x) "a1"
    generic2.a2 <- function(x) "a2"
    generic2.b <- function(x) {
      class(x) <- "a1"
      NextMethod()
    }

    generic2(structure(list(), class = c("b", "a2")))
    ```

1.  What classes have a method for the `Math` group generic in base R? Read 
    the source code. How do the methods work?

1.  R has two classes for representing date time data, `POSIXct` and 
    `POSIXlt`, which both inherit from `POSIXt`. Which generics have 
    different behaviours for the two classes? Which generics share the same
    behaviour?

## Dispatch details

This chapter concludes with a few additional details about method dispatch that is not well documented elsewhere. It is safe to skip these details if you're new to S3.

### Environments and namespaces

The precise rules for where a generic looks for the methods are a little complicated because there are two paths for discovery:

1. In the calling environment of the function that called the generic.

1. In the special `.__S3MethodsTable__.` object in the function environment of 
   the generic. Every package has an `.__S3MethodsTable__` which lists all
   the S3 methods exported by the package.

These details are not usually important, but are necessary in order for S3 generics to find the correct method when the generic and method are in different packages.

### S3 and base types {#implicit-class}

What happens when you call an S3 generic with a non-S3 object, i.e. an object that doesn't have the class attribute set? You might think it would dispatch on what `class()` returns:

```{r}
class(matrix(1:5))
```

But unfortunately dispatch actually occurs on the __implicit class__, which has three components:

* "array" or "matrix" (if the object has dimensions).
* `typeof()` (with a few minor tweaks).
* If it's "integer" or "double", "numeric".

There is no base function that will compute the implicit class, but you can use a helper from the sloop package: \index{implicit class} \index{base types!implicit class}

```{r}
s3_class(matrix(1:5))
```

`s3_dispatch()` knows about the implicit class, so use it if you're ever in doubt about method dispatch:

```{r}
s3_dispatch(print(matrix(1:5)))
```

Note that this can lead to different dispatch for objects that look similar:

```{r}
x1 <- 1:5
class(x1)
s3_dispatch(mean(x1))

x2 <- structure(x1, class = "integer")
class(x2)
s3_dispatch(mean(x2))
```

### Internal generics

Some S3 generics, like `[`, `sum()`, and `cbind()`, don't call `UseMethod()` because they are implemented in C. Instead, they call the C functions `DispatchGroup()` or `DispatchOrEval()`. These functions are called __internal generics__, because they do dispatch internally, in C code. Internal generics only exist in base R, so you cannot create an internal generic in a package.

`s3_dispatch()` shows internal generics by including the name of the generic at the bottom of the method class. If this method is called, all the work happens in C code, typically using [switchpatch].

```{r}
s3_dispatch(Sys.time()[1])
```

For performance reasons, internal generics do not dispatch to methods unless the class attribute has been set (`is.object()` is true). This means that internal generics do not use the implicit class. Again, if you're confused, rely on `s3_dispatch()` to show you the difference.

```{r}
x <- sample(10)
class(x)
s3_dispatch(x[1])

class(mtcars) 
s3_dispatch(mtcars[1])
```


### Group generics

Group generics are the most complicated part of S3 method dispatch because they involve both `NextMethod()` and internal generics. Group generics are worth learning about, however, because they allow you to implement a whole swath of methods with one function. Like internal generics, they only exist in base R, and you cannot define your own group generic.

Base R has four group generics, which are made up of the following generics: \index{group generics} \index{S3!group generics}

* __Math__: `abs`, `sign`, `sqrt`, `floor`, `cos`, `sin`, `log`, `exp`, ...

* __Ops__: `+`, `-`, `*`, `/`, `^`, `%%`, `%/%`, `&`, `|`, `!`, `==`, `!=`, `<`, 
  `<=`, `>=`, `>`

* __Summary__: `all`, `any`, `sum`, `prod`, `min`, `max`, `range`

* __Complex__: `Arg`, `Conj`, `Im`, `Mod`, `Re`

Defining a single group generic for your class overrides the default behaviour for all of the members of the group. Methods for group generics are looked for only if the methods for the specific generic do not exist:

```{r}
s3_dispatch(sum(Sys.time()))
```

Most group generics involve a call to `NextMethod()`. For example, take `difftime()` objects. If you look at the method dispatch for `abs()`, you'll see there's a `Math` group generic defined.

```{r}
y <- as.difftime(10, units = "mins")
s3_dispatch(abs(y))
```

`Math.difftime` basically looks like this:

```{r}
Math.difftime <- function(x, ...) {
  new_difftime(NextMethod(), units = attr(x, "units"))
}
```

It dispatches to the next method, here the internal default, to perform the actual computation, then copies back over the class and attributes.

Note that inside a group generic function a special variable `.Generic` provides the actual generic function called. This can be useful when producing error messages, and can sometimes be useful if you need to manually re-call the generic with different arguments.

### Double dispatch

Generics in the "Ops" group, which includes the two-argument mathematical and logical operators like `-` and `&`, implement a special type of method dispatch. They dispatch on the type of _both_ of the arguments, so called __double dispatch__. This is necessary to preserve the commutative property of many operators, i.e. `a + b` should equal `b + a`. Take the following simple example:

```{r}
date <- as.Date("2017-01-01")
integer <- 1L

date + integer
integer + date
```

If `+` dispatched only on the first argument, it would return different values for the two cases. To overcome this problem, generics in the Ops group use a slightly different strategy from usual. Rather than doing a single method dispatch, they do two, one for each input. There are three possible outcomes of this lookup:

* The methods are the same, so it doesn't matter which method is used.

* The methods are different, and R falls back to the internal method with a warning.

* One method is internal, in which case R calls the other method.

For the example above, we can look at the possible methods for each argument, taking advantage of the fact that we can call `+` with a single argument.  In this case, the second argument would dispatch to the internal `+` function, so R will call `+.Date`.

```{r}
s3_dispatch(+date)
s3_dispatch(+integer)
```

Let's take a look at another case. What happens if you try to add a date to a factor? There is no method in common, so R calls the internal `+` method (which preserves the attributes of the LHS), with a warning.

```{r, error = TRUE}
factor <- factor("a")
s3_dispatch(+factor)

date + factor
factor + date
```

Finally, what happens if we try to substract a POSIXct from a POSIXlt? A common `-.POSIXt` method is found and called.

```{r}
dt1 <- as.POSIXct(date)
dt2 <- as.POSIXlt(date)

s3_dispatch(-dt1)
s3_dispatch(-dt2)

dt1 - dt2
```

### Exercises

1.  `Math.difftime()` is more complicated than I described. Why?

1.  Explain the differences in dispatch below:

    ```{r}
    x1 <- 1:5
    class(x1)
    s3_dispatch(x1[1])
    
    x2 <- structure(x1, class = "integer")
    class(x2)
    s3_dispatch(x2[1])
    ```
