# S3 {#s3}

S3 is R's first and simplest OO system. It is the only OO system used in the base and stats packages, and it's the most commonly used system in CRAN packages. S3 is informal and ad hoc, but it has a certain elegance in its minimalism: you can't take away any part of it and still have a useful OO system. S3 should be your default: you should use it unless you have a compelling reason to use a different system.\index{S3} \index{objects!S3|see{S3}}

## Basics {#s3-basics}

An S3 object is always built on a base type (typically a vector) and has the "class" attributte set. For example, take the factor. It's built on top of an integer vector, and has attributes "levels", and "class". It's the class attribute that makes this an S3 object.

```{r}
f <- factor("a")

typeof(f)
attributes(f)
```

S3 objects differ in their behaviour to the underlying base type because of __generic functions__, or generics for short. A generic function has different behaviour based on the class of one of its arguments (almost always the first). You can see this difference with the most important generic function: `print()`.

```{r}
print(f)
print(unclass(f))
```

Generic functions behave differently for different S3 classes because they have __methods__. A method is a function that implements the generic for a specific class. Note that in S3, methods belong to the generic, not the the object or the class. This reflects a different school of though to most programming languages, but is a good fit to the problem of data analysis. \index{functions!generics|see{generics}} \index{S3!generics} \index{generics!S3}

To determine if a function is an S3 generic, you can inspect its source code for a call to `UseMethod()`: that's the function that figures out the correct method to call, the process of __method dispatch__.  \indexc{UseMethod()}

```{r}
print
```

Some S3 generics, like `[`, `sum()`, and `cbind()`, don't call `UseMethod()` because they are implemented in C. Instead, they call the C functions `DispatchGroup()` or `DispatchOrEval()`. Functions that do method dispatch in C code are called __internal generics__ and we'll talk about them in [internal generics].

Given an object, the job of an S3 generic is to call the right S3 method. You can recognise S3 methods by their names, which look like `generic.class()`. For example, the Date method for the `mean()` generic is called `mean.Date()`, and the factor method for `print()` is called `print.factor()`. This is the reason that most modern style guides discourage the use of `.` in function names: it makes them look like S3 methods. For example, is `t.test()` the `t` method for `test` objects? Similarly, the use of `.` in class names can also be confusing: is `print.data.frame()` the `print()` method for `data.frames`, or the `print.data()` method for `frames`?  

You can see all the methods that belong to a generic with `methods()`:

```{r}
methods("mean")
methods("t.test")
```

(Apart from methods defined in the base package, most S3 methods will not be visible: use `getS3method("generic", "class")` to read their source code.)

You can also list all generics that have a method for a given class:

```{r}
methods(class = "ts")
```

There's no way to list all S3 classes because there is no formal definition of an S3 class, as you'll learn shortly.

### Exercises

1.  What does the `is.data.frame.data.frame()` method do?

1.  The most important S3 objects in base R are factors, data.frames,
    and date/times. You've already seen the attributes and type that
    factors are build on. What data structures and attributes are
    the others built on?

1.  What does the following code return? What base type is built on?
    What attributes does it use?
    
    ```{r}
    x <- ecdf(rpois(100, 10))
    x
    ```

## New classes

S3 is a simple and ad hoc system; it has no formal definition of a class. To make an object an instance of a class, you just take an existing base object and set the class attribute. You can do that during creation with `structure()`, or after the fact with `class<-()`: \index{S3!classes} \index{classes!S3}

```{r}
# Create and assign class in one step
foo <- structure(list(), class = "foo")

# Create, then set class
foo <- list()
class(foo) <- "foo"
```

You can determine the class of any object using `class(x)`, and see if an object inherits from a specific class using `inherits(x, "classname")`. \index{attributes!class}

```{r}
class(foo)
inherits(foo, "foo")
```

Class names can be any string, but I recommend using only lower case letters and `_`. Avoid `.`. Opinion is mixed whether to use underscores (`my_class`) or CamelCase (`MyClass`) for multi-word class names. Just pick one convention and stick with it.  You can include multiple names in a character vector. We'll come back to that idea in [inheritance].

S3 has no checks for correctness. This means you can change the class of existing objects:

```{r, error = TRUE}
# Create a linear model
mod <- lm(log(mpg) ~ log(disp), data = mtcars)
class(mod)
print(mod)

# Turn it into a data frame (?!)
class(mod) <- "data.frame"

# Unsurprisingly this doesn't work very well
print(mod)

# However, the data is still there
mod$coefficients
```

If you've used other OO languages, this might make you feel queasy. But surprisingly, this flexibility causes few problems: while you _can_ change the type of an object, you never should. R doesn't protect you from yourself: you can easily shoot yourself in the foot. As long as you don't aim the gun at your foot and pull the trigger, you won't have a problem.

### Constructor

Because S3 itself provides no checks for correctness, it's up to you to adopt conventions to protect yourself. The first convention is to use a constructor function: 

* Whenever you create an S3 object, you should always do so by calling 
  a constructor.

* Whenever you create an S3 class of your own, you should always provide
  a constructor.
  
The job of the constructor is to ensure that the S3 object always has the same structure (i.e. it always uses the same attributes, and they are defined up front) and to perform checks for validity. Generally, an S3 constructor should err on the side of strictness rather than helpfulness. Typically you will also provide a helpful function for the user.

```{r}
new_factor <- function(x, levels) {
  stopifnot(is.integer(x))
  stopifnot(is.character(levels))
  
  if (any(x <= 0)) {
    stop("All `x` values must be greater than zero")
  }
  
  if (length(levels) < max(x)) {
    stop("There must at least as many `levels` as possible values in `x`")
  }
  
  structure(
    x,
    levels = levels,
    class = "factor"
  )
}
```

`validity` method approach. A la `tibble::tibble()`.

If writing a package, you will need to consider whether or not to make the S3 constructor function available to the user. 

You should use it if it's available (like for `factor()` and `data.frame()`). This ensures that you're creating the class with the correct components. Constructor functions usually have the same name as the class.

Apart from developer supplied constructor functions,


### Object styles

Two basic ways to create an object in S3: with a list or with attributes.  With attributes is best if your object behaves like a vector, list if you're starting from scratch.

S3 objects are usually built on top of lists, or atomic vectors with attributes. (You can refresh your memory of attributes with [attributes](#attributes).) You can also turn functions into S3 objects. Other base types are either rarely seen in R, or have unusual semantics that don't work well with attributes.




## New methods

A generic isn't useful without some methods. To add a method, you just create a regular function with the correct (`generic.class`) name: \index{methods!S3} \index{S3!methods}

```{r}
f.a <- function(x) "Class a"

a <- structure(list(), class = "a")
class(a)
f(a)
```

Adding a method to an existing generic works in the same way:

```{r}
mean.a <- function(x) "a"
mean(a)
```

As you can see, there's no check to make sure that the method returns the class compatible with the generic. It's up to you to make sure that your method doesn't violate the expectations of existing code.

### Method dispatch

S3 method dispatch is relatively simple. `UseMethod()` creates a vector of function names, like `paste0("generic", ".", c(class(x), "default"))` and looks for each in turn. The "default" class makes it possible to set up a fall back method for otherwise unknown classes. \index{S3!method dispatch rules}

```{r}
f <- function(x) UseMethod("f")
f.a <- function(x) "Class a"
f.default <- function(x) "Unknown class"

f(structure(list(), class = "a"))
# No method for b class, so uses method for a class
f(structure(list(), class = c("b", "a")))
# No method for c class, so falls back to default
f(structure(list(), class = "c"))
```

Once `UseMethod` has found the correct method, it's invoked in a special way. Rather than creating a new evaluation environment, it uses the environment of the current function call (the call to the generic), so any assignments or evaluations that were made before the call to UseMethod will be accessible to the method. The arguments that were used in the call to the generic are passed on to the method in the same order they were received.

Because methods are normal R functions, you can call them directly:

```{r}
c <- structure(list(), class = "c")
# Call the correct method:
f.default(c)
# Force R to call the wrong method:
f.a(c)
```

However, this is just as dangerous as changing the class of an object, so you shouldn't do it. Please don't point the loaded gun at your foot! The only reason to call the method directly is that sometimes you can get considerable performance improvements by skipping method dispatch. See [performance](#be-lazy) for details.

### Exercises

1.  Read the source code for `t()` and `t.test()` and confirm that 
    `t.test()` is an S3 generic and not an S3 method. What happens if 
    you create an object with class `test` and call `t()` with it?

1.  What classes have a method for the `Math` group generic in base R? Read 
    the source code. How do the methods work?

1.  R has two classes for representing date time data, `POSIXct` and 
    `POSIXlt`, which both inherit from `POSIXt`. Which generics have 
    different behaviours for the two classes? Which generics share the same
    behaviour?

1.  Which base generic has the greatest number of defined methods?

1.  `UseMethod()` calls methods in a special way. Predict what the following
     code will return, then run it and read the help for `UseMethod()` to 
    figure out what's going on. Write down the rules in the simplest form
    possible.

    ```{r, eval = FALSE}
    y <- 1
    g <- function(x) {
      y <- 2
      UseMethod("g")
    }
    g.numeric <- function(x) y
    g(10)

    h <- function(x) {
      x <- 10
      UseMethod("h")
    }
    h.character <- function(x) paste("char", x)
    h.numeric <- function(x) paste("num", x)

    h("a")
    ```

1.  Internal generics don't dispatch on the implicit class of base types.
    Carefully read `?"internal generic"` to determine why the length of `f` 
    and `g` is different in the example below. What function helps 
    distinguish between the behaviour of `f` and `g`?

    ```{r, eval = FALSE}
    f <- function() 1
    g <- function() 2
    class(g) <- "function"
    
    class(f)
    class(g)

    length.function <- function(x) "function"
    length(f)
    length(g)
    ```

## New generics

To add a new generic, create a function that calls `UseMethod()`. `UseMethod()` takes two arguments: the name of the generic function, and the argument to use for method dispatch. If you omit the second argument it will dispatch on the first argument to the function. There's no need to pass any of the arguments of the generic to `UseMethod()` and you shouldn't do so. `UseMethod()` uses black magic to find them out for itself. \indexc{UseMethod()} \index{S3!new generic}

```{r}
f <- function(x) UseMethod("f")
```

There are two main reasons to write your own generic:

* To avoid a big compound if statement
* To make it possible for others to extend your package

## Method dispatch


### Default method


### Method families


* When implementing a vector class, you should implement these methods: `length`, `[`, `[<-`, `[[`, `[[<-`, `c`.  (If `[` is implemented `rev`, `head`, and `tail` should all work). 

* When implementing anything mathematical, implement `Ops`, `Math` and `Summary`.

* When implementing a matrix/array class, you should implement these methods: `dim` (gets you nrow and ncol), `t`, `dimnames` (gets you rownames and colnames), `dimnames<-` (gets you colnames<-, rownames<-), `cbind`, `rbind`.

* If you're implementing more complicated `print()` methods, it's a better idea to implement `format()` methods that return a string, and then implement `print.class <- function(x, ...) cat(format(x, ...), "\n"`. This makes for methods that are much easier to compose, because the side-effects are isolated to a single place.

### Inheritance

The `NextMethod` function provides a simple inheritance mechanism, using the fact that the class of an S3 object is a vector. This is very different behaviour to most other languages because it means that it's possible to have different inheritance hierarchies for different objects:

```{r}
baz <- function(x) UseMethod("baz", x)
baz.A <- function(x) "A"
baz.B <- function(x) "B"

ab <- structure(1, class = c("A", "B"))
ba <- structure(1, class = c("B", "A"))
baz(ab)
baz(ba)
```

`NextMethod()` works like `UseMethod` but instead of dispatching on the first element of the class vector, it will dispatch based on the second (or subsequent) element:

```{r}
baz.C <- function(x) c("C", NextMethod())
ca <- structure(1, class = c("C", "A"))
cb <- structure(1, class = c("C", "B"))
baz(ca)
baz(cb)
```

The exact details are a little tricky: `NextMethod` doesn't actually work with the class attribute of the object, it uses a global variable (`.Class`) to keep track of which class to call next. This means that manually changing the class of the object will have no impact on the inheritance:

```{r}
# Turn object into class A - doesn't work!
baz.D <- function(x) {
  class(x) <- "A"
  NextMethod()
}
da <- structure(1, class = c("D", "A"))
db <- structure(1, class = c("D", "B"))
baz(da)
baz(db)
```

Methods invoked as a result of a call to `NextMethod` behave as if they had been invoked from the previous method. The arguments to the inherited method are in the same order and have the same names as the call to the current method, and are therefore are the same as the call to the generic. However, the expressions for the arguments are the names of the corresponding formal arguments of the current method. Thus the arguments will have values that correspond to their value at the time NextMethod was invoked. Unevaluated arguments remain unevaluated. Missing arguments remain missing.

If `NextMethod` is called in a situation where there is no second class it will return an error. A selection of these errors are shown below so that you know what to look for.

```{r, error = TRUE}
c <- structure(1, class = "C")

baz(c)

baz.c(c)

baz.c(1)
```

As discussed in the next section, R looks for methods in the order in which they appear in the class vector. So in this example, it would be like class A inherits from class B - if a method isn't defined for A, it will fall back to B. However, if you switched the order of the classes, the opposite would be true! This is because S3 doesn't define any formal relationship between classes, or even any definition of what an individual class is. If you're coming from a strict environment like Java, this will seem pretty frightening (and it is!) but it does give your users a tremendous amount of freedom. While it's very difficult to stop someone from doing something you don't want them to do, your users will never be held back because there is something you haven't implemented yet.

### Interaction with base types

You can also call an S3 generic with a non-S3 object. Non-internal S3 generics will dispatch on the __implicit class__ of base types. (Internal generics don't do that for performance reasons.) The rules to determine the implicit class of a base type are somewhat complex, but are shown in the function below: \index{implicit class} \index{base types!implicit class}

```{r}
iclass <- function(x) {
  if (is.object(x)) {
    stop("x is not a primitive type", call. = FALSE)
  }

  c(
    if (is.matrix(x)) "matrix",
    if (is.array(x) && !is.matrix(x)) "array",
    if (is.double(x)) "double",
    if (is.integer(x)) "integer",
    mode(x)
  )
}
iclass(matrix(1:5))
iclass(array(1.5))
```

S3 implements a style of object oriented programming called generic-function OO. This is different to most programming languages, like Java, C++ and C#, which implement message-passing OO. In message-passing style, messages (methods) are sent to objects and the object determines which function to call. Typically this object has a special appearance in the method call, usually appearing before the name of the method/message: e.g. `canvas.drawRect("blue")`. S3 is different. While computations are still carried out via methods, a special type of function called a __generic function__ decides which method to call. Methods are defined in the same way as a normal function, but are called in a different way, as we'll see shortly.

The primary use of OO programming in R is for print, summary and plot methods. These methods allow us to have one generic function, e.g. `print()`, that displays the object differently depending on its type: printing a linear model is very different to printing a data frame.

### Internal generics

Some internal C functions are also generic, which means that the method dispatch is not performed by R function, but is instead performed by special C functions. It's important to know which functions are internally generic, so you can write methods for them, and so you're aware of the slight differences in method dispatch. It's not easy to tell if a function is internally generic, because it just looks like a typical call to a C:

```{r}
length
cbind
```

As well as `length` and `cbind`, internal generic functions include `dim`, `c`, `as.character`, `names` and `rep`. A complete list can be found in the global variable `.S3PrimitiveGenerics`, and more details are given in `?InternalMethods`.

Internal generic have a slightly different dispatch mechanism to other generic functions: before trying the default method, they will also try dispatching on the __mode__ of an object, i.e. `mode(x)`. The following example shows the difference:

```{r, error = TRUE}
x <- structure(as.list(1:10), class = "myclass")
length(x)

mylength <- function(x) UseMethod("mylength", x)
mylength.list <- function(x) length(x)
mylength(x)
```

### Double dispatch

How does `+` work.

### Group generics

Group generic methods add a little more complexity. Group generics make it possible to implement methods for multiple generics with one function. The four group generics and the functions they include are: \index{group generics} \index{S3!group generics}

* Math: `abs`, `sign`, `sqrt`, `floor`, `cos`, `sin`, `log`, `exp`, ...
* Ops: `+`, `-`, `*`, `/`, `^`, `%%`, `%/%`, `&`, `|`, `!`, `==`, `!=`, `<`, `<=`, `>=`, `>`
* Summary: `all`, `any`, `sum`, `prod`, `min`, `max`, `range`
* Complex: `Arg`, `Conj`, `Im`, `Mod`, `Re`

Group generics are a relatively advanced technique and are beyond the scope of this chapter but you can find out more about them in `?groupGeneric`. The most important thing to take away from this is to recognise that `Math`, `Ops`, `Summary`, and `Complex` aren't real functions, but instead represent groups of functions. Note that inside a group generic function a special variable `.Generic` provides the actual generic function called.
