---
title: Style guide
layout: default
---

# Style guide

Good coding style is like using correct punctuation when writing: you can manage without it, but it sure makes things easier to read. As with punctuation, there are many possible variations, and the main thing is to be consistent. The following guide describes the style that I use - you don't have to use it, but you need to have some consistent style that you do follow.  My style is  based on Google's [R style guide][1], with a few tweaks.

Good style is important because while your code only has one author, it will usually have multiple readers, and when you know you will be working with multiple people on the same code, it's a good idea to agree on a common style up-front.  No style is uniformly better than any other style, and if you're working with a group of people, you may need to sacrifice some of your most favourite types of style.

One package that can make adhering to a style guide easier is `formatR`, by Yihui Xie.  It can't do everything, but if you're starting with very poorly formatted R code, it will get you to a good place much more quickly than doing everything by hand.  Make sure to read [the notes on the wiki](https://github.com/yihui/formatR/wiki) before using it.

## Notation and naming

### File names

File names should end in `.R` and be meaningful.

    # Good
    fit-models.R
    utility-functions.R

    # Bad
    foo.r
    stuff.r

If files need to be run in sequence, prefix with numbers:

    0-download.R
    1-parse.R
    2-explore.R

### Identifiers

> "There are only two hard things in Computer Science: cache invalidation and 
> naming things." -- Phil Karlton

Variable and function names should be lowercase. Use `_` to separate words within a name. Generally, variable names should be nouns and function names should be verbs. Strive for concise but meaningful names (this is not easy!)

```{r, eval = FALSE}
# Good
day_one
day_1

# Bad
first_day_of_the_month
DayOne
dayone
djm1
```

Where possible, you should avoid using names of commonly used existing functions and variables.  This will cause confusion for the readers of your code.

```{r, eval = FALSE}
# Bad
T <- FALSE
c <- 10
mean <- function(x) sum(x)
```

## Syntax

### Spacing

Place spaces around all infix operators (`=`, `+`, `-`, `<-`, etc.). Always put a space after a comma, and never before (just like in regular English).

```{r, eval = FALSE}
# Good
average <- mean(feet / 12 + inches, na.rm = TRUE)

# Bad
average<-mean(feet/12+inches,na.rm=TRUE)
```

Place a space before left parentheses, except in a function call.

```{r, eval = FALSE}
# Good
if (debug) do(x)
plot(x, y)

# Bad
if(debug)do(x)
plot (x, y)
```

Extra spacing (i.e., more than one space in a row) is okay if it improves alignment of equals signs or arrows (`<-`).

```{r, eval = FALSE}
list(
  total = a + b + c, 
  mean  = (a + b + c) / n
)
```

Do not place spaces around code in parentheses or square brackets (unless there's a comma, in which case see above.

```{r, eval = FALSE}
# Good
if (debug) do(x)
diamonds[5, ]

# Bad
if ( debug )  do(x) # No spaces around debug
x[1,]  # Needs a space after the comma
x[1 ,]  # Space goes after comma not before
```

### Curly braces

An opening curly brace should never go on its own line and should always be followed by a new line; a closing curly brace should always go on its own line, unless followed by `else`.

Always indent the code inside the curly braces.

```{r, eval = FALSE}
# Good

if (y < 0 && debug) {
  message("Y is negative")
}

if (y == 0) {
  log(x)
} else {
  y ^ x
}

# Bad

if (y < 0 && debug)
message("Y is negative")

if (y == 0) {
  log(x)
} 
else {
  y ^ x
}
```

It's ok to leave very short statements on the same line:

```{r, eval = FALSE}
if (y < 0 && debug) message("Y is negative")
```

### Line length

Keep your lines less than 80 characters. This is the amount that will fit comfortably on a printed page at a reasonable size. If you find you are running out of room, this is probably an indication that you should encapsulate some of the work in a separate function.

### Indentation

When indenting your code, use two spaces. Never use tabs or mix tabs and spaces.

The only exception is if a function definition runs over multiple lines: indent the second line to line up with where the definition starts:

```{r, eval = FALSE}
long_function_name <- function(a = "a long argument", b = "another argument",
                               c = "another long argument") {
  # As usual code is indented by two spaces.
}
```

### Assignment

Use `<-`, not `=`, for assignment.

```{r}
# Good
x <- 5
# Bad
x = 5
```

## Organisation

### Commenting guidelines

Comment your code. Entire commented lines should begin with `#` and one space. Comments should explain the why, not the what.

Use commented lines of `-` and `=` to break up your files into scannable chunks.

```{r, eval = FALSE}
# Load data ---------------------------

# Plot data ---------------------------
```

 [1]: http://google-styleguide.googlecode.com/svn/trunk/google-r-style.html
