---
title: Untitled
output:
  md_document: {}
---

There are two parts to this problem:

1. checking that inputs are valid
2. coercing a list to a vector

## Checking valid inputs

First, I'd avoid `is()` because it's known to be slow. That gives:

```{r}
check_valid <- function (elem, mode) {
  if (length(elem) != 1) stop("Must be length 1")
  if (mode(elem) != mode) stop("Not desired type")

  TRUE
}
```

Now we need to figure out whether a loop or apply variant is faster. We'll benchmark with the worst possible case where all inputs are valid.

```{r}
worst <- as.list(0:101)

library(microbenchmark)
options(digits = 3)
microbenchmark(
  `for` = for(i in seq_along(worst)) check_valid(worst[[i]], "numeric"),
  lapply = lapply(worst, check_valid, "numeric"),
  vapply = vapply(worst, check_valid, "numeric", FUN.VALUE = logical(1))
)
```

The three methods are basically tied. `lapply()` is very slightly faster, probably because of the special C tricks that it uses

## Coercing list to vector

Now let's look at a few ways of coercing a list to a vector:

```{r}
change_mode <- function(x, mode) {
  mode(x) <- mode
  x
}

microbenchmark(
  change_mode = change_mode(worst, "numeric"),
  unlist = unlist(worst),
  as.vector = as.vector(worst, "numeric")
)
```

So it looks like you're already using the fastest method, and the total cost is dominated by the check.

## Alternative approach

Another idea is that we might be able to get a little faster by looping over the vector once, instead of once to check and once to coerce:

```{r}
as_atomic_for <- function (x, mode) {
  out <- vector(mode, length(x))

  for (i in seq_along(x)) {
    check_valid(x[[i]], mode)
    out[i] <- x[[i]]
  }

  out
}
microbenchmark(
  as_atomic_for(worst, "numeric")
)
```

That's definitely worse.

All in all, I think this suggests if you want to make this function faster, you should try vectorising the check function in Rcpp.
