---
output:
  md_document: {}
---

As well as reducing the number of tests performed, or otherwise optimising the running of the whole function, we might be able to make `assocstats` faster. We'll start by establishing a test case to make sure we don't accidentally make a faster function that's incorrect.

```{r}
x <- vcd::Arthritis$Improved
y <- vcd::Arthritis$Treatment
correct <- vcd::assocstats(table(x, y))$cramer
correct

is_ok <- function(x) stopifnot(all.equal(x, correct))
```

We'll start by making a version of `assocstats` that's very close to the original.

```{r}
cramer1 <- function (x, y) {
  mat <- table(x, y)

  tab <- summary(MASS::loglm(~1 + 2, mat))$tests

  phi <- sqrt(tab[2, 1] / sum(mat))
  cont <- sqrt(phi ^ 2 / (1 + phi ^ 2))

  sqrt(phi ^ 2 / min(dim(mat) - 1))
}
is_ok(cramer1(x, y))
```

The slowest operation here is going to be `loglm`, so before we try making that faster, it's worth looking for an alternative approach. A little googling finds [a useful blog post](http://sas-and-r.blogspot.com/2011/06/example-839-calculating-cramers-v.html). Let's also try that:

```{r}
cramer2 <- function(x, y) {
  chi <- chisq.test(x, y, correct=FALSE)$statistic[[1]]

  ulength_x <- length(unique(x))
  ulength_y <- length(unique(y))

  sqrt(chi / (length(x) * (min(ulength_x, ulength_y) - 1)))
}
is_ok(cramer2(x, y))
```

How does the performance stack up:

```{r}
library(microbenchmark)

microbenchmark(
  cramer1(x, y),
  cramer2(x, y)
)
```

`cramer2()` is faster. `chisq.test()` is likely to be the bottleneck, so lets see if we can make that function faster by doing less: `chisq.test()` does a lot more than compute the test-statistic, so it's likely that we can make it faster. A few minutes careful work reduces the function to:

```{r}
chisq_test <- function (x, y) {
  O <- table(x, y)
  n <- sum(O)

  E <- outer(rowSums(O), colSums(O), "*")/n

  sum((abs(O - E))^2 / E)
}
```

We can then create a new `cramer3()` that uses `chisq.test()`.

```{r}
cramer3 <- function(x, y) {
  chi <- chisq_test(x, y)

  ulength_x <- length(unique(x))
  ulength_y <- length(unique(y))

  sqrt(chi / (length(x) * (min(ulength_x, ulength_y) - 1)))
}
is_ok(cramer3(x, y))
microbenchmark(
  cramer1(x, y),
  cramer2(x, y),
  cramer3(x, y)
)
```

And now that we have our own simple version of `chisq.test()` we could eek out a little more speed by using the results of `table()` to figure out the number of unique elements in `x` and `y`.

```{r}
cramer4 <- function(x, y) {
  O <- table(x, y)
  n <- length(x)
  E <- outer(rowSums(O), colSums(O), "*")/n

  chi <- sum((abs(O - E))^2 / E)
  sqrt(chi / (length(x) * (min(dim(O)) - 1)))
}
is_ok(cramer4(x, y))

microbenchmark(
  cramer1(x, y),
  cramer2(x, y),
  cramer3(x, y),
  cramer4(x, y)
)

```

Not bad - we've made it 4 times faster just using R code. From here, you could try to get even more speed by:

* Using `tcrossprod()` instead of `outer()`
* Making a faster version of `table()` for this special (2d) case
* Using Rcpp to compute the test-statistic from the tabular data
