---
title: Untitled
output:
  md_document: {}

---


We begin by defining the lookup matrices. I've laid them out in a way that should make them easier to check against a reference, e.g. http://en.wikipedia.org/wiki/Verhoeff_algorithm.

```{r}
d5_mult <- matrix(as.integer(c(
  0, 1, 2, 3, 4, 5, 6, 7, 8, 9,
  1, 2, 3, 4, 0, 6, 7, 8, 9, 5,
  2, 3, 4, 0, 1, 7, 8, 9, 5, 6,
  3, 4, 0, 1, 2, 8, 9, 5, 6, 7,
  4, 0, 1, 2, 3, 9, 5, 6, 7, 8,
  5, 9, 8, 7, 6, 0, 4, 3, 2, 1,
  6, 5, 9, 8, 7, 1, 0, 4, 3, 2,
  7, 6, 5, 9, 8, 2, 1, 0, 4, 3,
  8, 7, 6, 5, 9, 3, 2, 1, 0, 4,
  9, 8, 7, 6, 5, 4, 3, 2, 1, 0
)), ncol = 10, byrow = TRUE)

d5_perm <- matrix(as.integer(c(
  0, 1, 2, 3, 4, 5, 6, 7, 8, 9,
  1, 5, 7, 6, 2, 8, 3, 0, 9, 4,
  5, 8, 0, 3, 7, 9, 6, 1, 4, 2,
  8, 9, 1, 6, 0, 4, 3, 5, 2, 7,
  9, 4, 5, 3, 1, 2, 6, 8, 7, 0,
  4, 2, 8, 6, 5, 7, 3, 9, 0, 1,
  2, 7, 9, 3, 8, 0, 6, 4, 1, 5,
  7, 0, 4, 6, 9, 1, 3, 2, 5, 8
)), ncol = 10, byrow = TRUE)

d5_inv <- as.integer(c(0, 4, 3, 2, 1, 5, 6, 7, 8, 9))
```

Next, we'll define the check function, and try it out with a test input. I've followed the derivation in wikipedia as closely as possible.

```{r}
p <- function(i, n_i) {
  d5_perm[(i %% 8) + 1, n_i + 1] + 1
}
d <- function(c, p) {
  d5_mult[c + 1, p]
}

verhoeff <- function(x) {
  #split and convert to numbers
  digs <- strsplit(as.character(x), "")[[1]]
  digs <- as.numeric(digs)
  digs <- rev(digs)   ## right to left algorithm

  ## apply algoritm - note 1-based indexing in R
  c <- 0
  for (i in 1:length(digs)) {
    c <- d(c, p(i, digs[i]))
  }

  d5_inv[c + 1]
}
verhoeff(142857)
```

This function is fundamentally iterative, as each iteration depends on the value of the previous. This means that we're unlikely to be able to vectorise in R, so if we want to vectorise, we'll need to use Rcpp.

However, before we turn to that, it's worth exploring if we can do the initial split faster. First we do a little microbenchmark to see if it's worth bothering:

```{r}
library(microbenchmark)
digits <- function(x) {
  digs <- strsplit(as.character(x), "")[[1]]
  digs <- as.numeric(digs)
  rev(digs)
}

microbenchmark(
  digits(142857),
  verhoeff(142857)
)
```

It looks like it! On my computer, `verhoeff_prepare()` accounts for about 50% of the run time. A little searching on stackoverflow reveals another approach to turning a [number into digits](http://stackoverflow.com/questions/18786432):

```{r}
digits2 <- function(x) {
   n <- floor(log10(x))
   x %/% 10^(0:n) %% 10
}
digits2(12345)
microbenchmark(
  digits(142857),
  digits2(142857)
)
```

`digits2()` is a lot faster than `digits()` but it has limited impact on the whole runtime.

```{r}
verhoeff2 <- function(x) {
  digs <- digits2(x)

  c <- 0
  for (i in 1:length(digs)) {
    c <- d(c, p(i, digs[i]))
  }

  d5_inv[c + 1]
}
verhoeff2(142857)

microbenchmark(
  verhoeff(142857),
  verhoeff2(142857)
)
```

It's always worth checking out the impact of byte-code compilation:

```{r}
verhoeff_c <- compiler::cmpfun(verhoeff)
verhoeff2_c <- compiler::cmpfun(verhoeff2)
microbenchmark(
  verhoeff(142857),
  verhoeff_c(142857),
  verhoeff2(142857),
  verhoeff2_c(142857)
)

```

It looks like it speeds it up by around 10%, which is fairly typical.

To make it even faster we could try C++.

```{r, engine='Rcpp'}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
int verhoeff3_c(IntegerVector digits, IntegerMatrix mult, IntegerMatrix perm,
                IntegerVector inv) {
  int n = digits.size();
  int c = 0;

  for(int i = 0; i < n; ++i) {
    int p = perm(i % 8, digits[i]);
    c = mult(c, p);
  }

  return inv[c];
}
```

```{r}
verhoeff3 <- function(x) {
  verhoeff3_c(digits(x), d5_mult, d5_perm, d5_inv)
}
verhoeff3(142857)

microbenchmark(
  verhoeff2(142857),
  verhoeff3(142857)
)
```

That doesn't yield much of an improvement. Maybe we can do better if we pass the number to C++ and process the digits in a loop:

```{r, engine = "Rcpp"}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
int verhoeff4_c(int number, IntegerMatrix mult, IntegerMatrix perm,
                IntegerVector inv) {
  int c = 0;
  int i = 0;

  for (int i = 0; number > 0; ++i, number /= 10) {
    int p = perm(i % 8, number % 10);
    c = mult(c, p);
  }

  return inv[c];
}
```

```{r}
verhoeff4 <- function(x) {
  verhoeff4_c(x, d5_mult, d5_perm, d5_inv)
}
verhoeff4(142857)

microbenchmark(
  verhoeff2(142857),
  verhoeff3(142857),
  verhoeff4(142857)
)
```

And we get a pay off: `verhoeff4()` is about 5 times faster than `verhoeff2()`.


