\chapter{Welcome}

This is the in-progress book site for \textbf{``Advanced R''}. The book
is designed primarily for R users who want to improve their programming
skills and understanding of the language. It should also be useful for
programmers coming to R from other languages, as it explains some of R's
quirks and shows how some parts that seem horrible do have a positive
side.

It will eventually be published as a real book in Chapman and Hall's
\href{http://www.crcpress.com/browse/series/crctherser}{R series}. The
final version of the book is due in June 2014, so it should be available
in late 2014. Thanks to the publisher, the wiki will continue to be
freely available after the book is published.

\section{Advanced R}

Learn all about how the R language works and how you can use it to solve
real problems. The book is broken down into the following chapters
(which you can also access from the drop-down menus at the top-right):

\{\% include programming-nav.html \%\}

\section{Package development}

Packages are the fundamental units of reproducible R code. They include
reusable R functions, the documentation that describes how to use them,
and sample data. In this section you'll learn how to turn your code into
packages that others can easily download and use.

Writing a package can seem overwhelming at first. So start with the
basics and improve it over time. It doesn't matter if your first version
isn't perfect as long as the next version is better.

\{\% include package-nav.html \%\}
