---
title: "Untitled"
output: html_document
---


If you're working with a list of functions, remember to use `call_fun`:

```{r}
call_fun <- function(f, ...) f(...)
f <- list(sum, mean, median, sd)
lapply(f, call_fun, x = runif(1e3))
```

Or you could create a variant, `fapply()`, specifically for working with lists of functions:

```{r}
fapply <- function(fs, ...) {
  out <- vector("list", length(fs))
  for (i in seq_along(fs)) {
    out[[i]] <- fs[[i]](...)
  }
  out
}
fapply(f, x = runif(1e3))
```

---

When writing functionals, you can expect your users to know of `partial()` and not use `...` For example, instead of implementing `lapply()` like:

```{r}
lapply2 <- function(x, f, ...) {
  out <- vector("list", length(x))
  for (i in seq_along(x)) {
    out[[i]] <- f(x[[i]], ...)
  }
  out
}
unlist(lapply2(1:5, log, base = 10))
```

we could implement it as:

```{r}
lapply3 <- function(x, f) {
  out <- vector("list", length(x))
  for (i in seq_along(x)) {
    out[[i]] <- f(x[[i]])
  }
  out
}
unlist(lapply3(1:5, partial(log, base = 10)))
```


---


If you look closely, you'll notice we're just applying a function to every element in a list. Since that's the job of `lapply`, we can further reduce the code:

```{r}
funs <- c(sum = sum, mean = mean, median = median)
funs2 <- lapply(funs, partial, na.rm = TRUE)
```

Next, let's think about a similar, but subtly different case. Say we have a numeric vector and we want to generate a list of means that are variously trimmed. The following code won't work because we want the first argument of `partial` to be the `mean` function. Since fixed matching overrides positional matching, we could instead try specifying the argument by name. But that won't work because the `trims` end up being the first argument of `mean`.

```{r, error = TRUE}
(trims <- seq(0, 0.9, length = 5))
funs3 <- lapply(trims, partial, `_f` = mean)
funs3[[1]]
sapply(funs3, call_fun, c(1:100, (1:50) * 100))
```

We can resolve this problem by telling `partial()` not to be lazy, and to eagerly evaluate its inputs:

```{r}
(trims <- seq(0, 0.9, length = 5))
funs3 <- lapply(trims, partial, `_f` = mean, .lazy = FALSE)
sapply(funs3, call_fun, c(1:100, (1:50) * 100))
```

But that doesn't work either because each function gets a promise to evaluate `t`, and that promise isn't evaluated until all of the functions are run, when `t = 0.9`. To make it work, you need to manually force the evaluation of `t`:

```{r}
funs5 <- lapply(trims, function(t) {
  force(t)
  partial(mean, trim = t)
})
funs5[[1]]
sapply(funs5, call_fun, c(1:100, (1:50) * 100))
```
