## Case study speeding up bootstrap fits of linear model

The `lm()` function is a good example.

1.  `model.frame()` handles non-standard evaluation to create a single
data frame containing all variables needed.

2.  `model.matrix()` converts a formula and a data frame to a matrix.

3.  `lm.fit()` performs the linear algebra

4.  The final results and combined with intermediate pieces to create
an S3 object.

```{r}
fit_model <- function(df) {
  coef(lm(mpg ~ disp + wt + cyl, data = df))
}
boot <- function(df) {
  df[sample(nrow(df), replace = TRUE), , drop = FALSE]
}
system.time(replicate(1000, fit_model(boot(mtcars))))

library(lineprof)
l <- lineprof(replicate(1000, fit_model(boot(mtcars))))
shine(l)
```

```{r}
library(biglm)
library(RcppArmadillo)

mtcars$cyl <- factor(mtcars$cyl)
mtcars_m <- model.matrix(mpg ~ disp + wt + cyl, mtcars)

library(microbenchmark)
microbenchmark(
  mtcars_m <- model.matrix(mpg ~ disp + wt + cyl, mtcars),
  coef(lm(mpg ~ disp + wt + cyl, data = mtcars)),
  coef(lm.fit(mtcars_m, mtcars$mpg)),

  coef(fastLm(mpg ~ disp + wt + cyl, data = mtcars)),
  coef(fastLmPure(mtcars_m, mtcars$mpg))[, 1]
)
```

```{r, eval = FALSE}
data(diamonds, package = "ggplot2")
diamonds_m <- model.matrix(price ~ carat + color, diamonds)

library(microbenchmark)
microbenchmark(
  diamonds_m <- model.matrix(price ~ carat + color, diamonds),
  coef(lm(price ~ carat + color, diamonds)),
  coef(lm.fit(diamonds_m, diamonds$price)),

  coef(fastLm(price ~ carat + color, diamonds)),
  coef(fastLmPure(diamonds_m, diamonds$price))[, 1],
  times = 10
)
```
