## Explicit scoping with `local`

Sometimes it's useful to be able to create a new scope without embedding inside a function. The `local()` function allows you to do exactly that. For example, to make an operation easier to understand, you can make temporary variables. In this example, `df()` is created in the global environment, but `x` and `y` are not:

```{r}
df <- local({
  x <- 1:10
  y <- runif(10)
  data.frame(x = x, y = y)
})
```

This is equivalent to:

```{r}
df <- (function() {
  x <- 1:10
  y <- runif(10)
  data.frame(x = x, y = y)
})()
```

(If you're familiar with JavaScript you've probably seen this pattern before. It's the immediately invoked function expression (IIFE). It's used extensively by many JavaScript libraries to avoid polluting the global namespace.)

## Evaluating code in a new environment

Another useful application of these ideas is `local()`. In the process of performing a data analysis, you may create variables that are necessary because they help break a complicated sequence of steps down in to easily digestible chunks, but are not needed afterwards. For example, in the following example, we might only want to keep the value of x:

```{r}
a <- 10
b <- 30
x <- a + b
rm(a)
rm(b)
```

It's useful to be able to store only the final result, preventing the intermediate results from cluttering your workspace.  One way to do that is to use a function:

```{r, error = TRUE}
x <- (function() {
  a <- 10
  b <- 30
  a + b
})()
x
a
```

(In JavaScript this is called the immediately invoked function expression (IIFE), and is used extensively in modern JavaScript to encapsulate different JavaScript libraries)

R provides a more concise tool to achieve this effect: `local()`.

```{r, error = TRUE}
x <- local({
  a <- 10
  b <- 30
  a + b
})
x
a
```

The idea of local is to create a new environment (inheriting from the current environment) and run the code in that. The essence of `local()` is captured in this code:

```{r}
local2 <- function(expr) {
  envir <- new.env(parent = parent.frame())
  eval(substitute(expr), envir)
}

a <- 100
local2({
  b <- a + sample(10, 1)
  my_get <<- function() b
})
my_get()

```

The real `local()` code is considerably more complicated because it adds a parameter to specify the environment. I don't think this is necessary because if you have an explicit environment parameter, then you can already evaluate code in that environment with `evalq()`.

### Exercises

1.  `local()` is hard to understand because it is very concise and uses
    some sutble features of evaluation (including non-standard evaluation
    of both arguments). Confirm that the following function works the same
    as `local()` and then explain how it works.

    ```{r}
    local3 <- function(expr, envir = new.env()) {
      call <- substitute(eval(quote(expr), envir))
      env <- parent.frame()

      eval(call, env)
    }
    ```
