
## Applications

To show how I've used some of these ideas in practice, the following two sections show applications of non-standard evaluation to plyr and ggplot2.

### `plyr::.` and `ggplot2::aes`

Both plyr and ggplot2 have ways of capturing what you want to do, and then performing that action later. ggplot2 uses the `aes()` to define a set of mappings between variables in your data and visual properties on your graphic. plyr uses the `.` function to capture the names (or more complicated expressions) of variables used to split a data frame into pieces. Let's look at the code:

```{r}
. <- function (..., .env = parent.frame()) {
  structure(
    as.list(match.call()[-1]),
    env = .env,
    class = "quoted"
  )
}

aes <- function (x = NULL, y = NULL, ...) {
  aes <- structure(
    as.list(match.call()[-1]),
    class = "uneval")
  class(aes) <- "uneval"
  ggplot2:::rename_aes(aes)
}
```

Both functions were written when I didn't know so much about non-standard evaluation, and if I was to write them today, I'd use the `dots()` helper function I showed previously.  I'd also think more about the environment in which the results of `aes()` should be evaluated, and how that integrates with ggplot2's rules for aesthetic mapping inheritance. That's a bit murky at the moment and leads to confusion when creating complex graphics across multiple functions.

ggplot2 and plyr provide slightly different ways to use standard evaluation so that you can refer to variables by reference. ggplot2 provides `aes_string()` which allows you to specify variables by the string representation of their name, and plyr uses S3 methods so that you can either supply an object of class quoted (as created with `.()`), or a regular character vector.

### Plyr: summarise, mutate and arrange

The plyr package also uses non-standard evaluation to complete the set of tools provided by the base `subset()` and `transform()` functions with `mutate()`, `summarise()` and `arrange()`. Each of these functions has the same interface: the first argument is a data frame and the subsequent arguments are evaluated in the context of that data frame (i.e. they look there first for variables, and then in the current environment) and they return a data frame.

The following code shows the essence of how these four functions work:

```{r}
subset2 <- function(.data, subset) {
sub <- eval(substitute(subset), .data, parent.frame())
sub <- sub & !is.na(sub)

.data[sub, , drop = FALSE]
}
arrange2 <- function (.data, ...) {
ord <- eval(substitute(order(...)), .data, parent.frame())
.data[ord, , drop = FALSE]
}

mutate2 <- function(.data, ...) {
cols <- named_dots(...)
data_env <- eval_df(.data, parent.frame(), cols)

out_cols <- union(names(.data), names(cols))
quickdf(mget(out_cols, data_env))
}
summarise2 <- function (.data, ...) {
cols <- named_dots(...)
data_env <- eval_df(.data, parent.frame(), cols)

quickdf(mget(names(cols), env))
}
eval_df <- function(data, env, expr) {
data_env <- list2env(data, parent = env)

for(nm in names(exprs)) {
data_env[[nm]] <- eval(data_env[[nm]], env)
}
data_env
}
```

You might be surprised to see the for loops in `eval_df`, but they are necessary because the computation of one variable might depend on the results of previous variables (this is the key difference between `mutate()` and `transform()`).
